---
title: "RNA-seq Analysis - Improved Version"
author: "Aleksandr Bykov (Enhanced by Claude Code)"
date: '`r Sys.Date()`'
output:
  html_document: 
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 8,
  dpi = 300
)

# Set random seed for reproducibility
set.seed(42)

# Record start time
analysis_start_time <- Sys.time()
message("Starting RNA-seq analysis at: ", analysis_start_time)
```

## Configuration and Setup

```{r configuration}
# Analysis configuration
config <- list(
  filtering = list(
    abs_filt_samples = 2,
    min_count_threshold = 10
  ),
  statistical = list(
    padj_cutoff = 0.05,
    log2FC_cutoff = 1,
    var_expl_needed = 0.6
  ),
  visualization = list(
    ntop_genes_pca = 1000,
    plot_width = 20,
    plot_height = 20,
    plot_units = "cm",
    top_pathways_show = 20
  ),
  fgsea = list(
    min_size = 15,
    max_size = 500,
    top_pathways_up = 10,
    top_pathways_down = 10
  )
)

# File paths configuration
PATHS <- list(
  data_dir = here::here("neuroblastoma", "data", "RNAseq"),
  results_dir = here::here("neuroblastoma", "results", "RNA-seq"),
  resources_dir = here::here("neuroblastoma", "resources"),
  cache_dir = here::here("neuroblastoma", "cache")
)

# Genes of interest for highlighting
GENES_OF_INTEREST <- list(
  all = c("VIM", "YAP1", "WWTR1", "JUN", "FOSL1", "FOSL2", "PHOX2B", "HAND2", "GATA3"),
  mesenchymal_markers = c("VIM", "CDH2", "FN1"),
  yap_taz_pathway = c("YAP1", "WWTR1", "TEAD1", "TEAD2", "TEAD3", "TEAD4"),
  ap1_factors = c("JUN", "JUNB", "JUND", "FOS", "FOSB", "FOSL1", "FOSL2"),
  adrenergic_markers = c("PHOX2B", "HAND2", "GATA3", "DBH", "TH")
)

print("Configuration loaded successfully")
```

```{r load_libraries}
# Load required libraries
suppressPackageStartupMessages({
  # Core analysis packages
  library(DESeq2)
  library(stringr)
  library(org.Hs.eg.db)
  library(biomaRt)
  
  # Enrichment analysis
  library(fgsea)
  library(hypeR)
  
  # Visualization
  library(ggplot2)
  library(ggrepel)
  library(RColorBrewer)
  library(pheatmap)
  
  # Data manipulation
  library(dplyr)
  library(tidyr)
  library(readr)
  
  # File operations
  library(openxlsx)
  library(here)
  
  # Utility functions from existing scripts
  import::from(
    .from = here::here("neuroblastoma/resources/UtilityScriptsRNA-seq.R"),
    "filterDatasets", "generatePCA", "generateEnsemblAnnotation", 
    "generateResults", "meanExprsPerGroup", "plotVolcano",
    .character_only = TRUE
  )
})

# Load custom utility functions
source(here::here("neuroblastoma/resources/rna_seq_analysis_utils.R"))

print("Libraries loaded successfully")
```

```{r setup_environment}
# Setup output directories
setup_output_dirs(PATHS$results_dir, c("plots", "tables", "reports"))
setup_output_dirs(PATHS$cache_dir)

# Setup BioMart with error handling
tryCatch({
  options(timeout = 30000)
  mart <- biomaRt::useMart(
    biomart = "ENSEMBL_MART_ENSEMBL", 
    dataset = "hsapiens_gene_ensembl",
    host = "https://www.ensembl.org"
  )
  message("BioMart connection established")
}, error = function(e) {
  warning("Failed to connect to BioMart: ", e$message)
  mart <- NULL
})

# Load gene sets with caching
gene_sets <- load_gene_sets(file.path(PATHS$cache_dir, "gene_sets.rds"))

print("Environment setup completed")
```

## Data Loading and Validation

```{r load_data}
# Load RNA-seq data
message("Loading RNA-seq data...")

rnaseq_dds <- readRDS(file.path(PATHS$data_dir, "rnaseq_deseq_global_deseq_data_set.rds"))
design(rnaseq_dds) <- as.formula("~ cell_line + cell_type")

annotation_data <- read.table(
  file = file.path(PATHS$data_dir, "rnaseq_deseq_global_annotation_gene.tsv"),
  sep = "\t",
  header = TRUE
)

# Standardize column names
colnames(annotation_data)[c(5, 7)] <- c("ensembl_id", "gene_symbol")

# Validate inputs
validate_inputs(rnaseq_dds, annotation_data)

# Display data summary
message("Data summary:")
message("- Samples: ", ncol(rnaseq_dds))
message("- Genes: ", nrow(rnaseq_dds))
message("- Cell types: ", paste(unique(rnaseq_dds$cell_type), collapse = ", "))
message("- Cell lines: ", paste(unique(rnaseq_dds$cell_line), collapse = ", "))
```

## Quality Control Analysis

### Initial PCA with All Samples

```{r initial_pca}
message("Performing initial quality control analysis...")

# Filter lowly expressed genes
rnaseq_dds_filt_initial <- filterDatasets(
  rnaseq_dds,
  abs_filt = TRUE,
  abs_filt_samples = config$filtering$abs_filt_samples
)

# Run DESeq2 pipeline
rnaseq_dds_filt_initial <- DESeq2::estimateSizeFactors(rnaseq_dds_filt_initial)
rnaseq_dds_filt_initial <- DESeq2::DESeq(rnaseq_dds_filt_initial)
vsd_initial <- DESeq2::vst(rnaseq_dds_filt_initial, blind = TRUE)

# Generate initial PCA
pca_initial <- generatePCA(
  transf_object = vsd_initial,
  cond_interest_varPart = c("cell_type", "cell_line"),
  color_variable = "cell_type",
  shape_variable = "cell_line",
  ntop_genes = config$visualization$ntop_genes_pca
) +
  scale_color_manual(values = c("navy", "firebrick3")) +
  scale_shape_manual(values = c(16, 7, 17, 15, 3)) +
  labs(title = "PCA: All Samples (Before QC)") +
  theme_publication()

# Save plot
safe_save_plot(
  pca_initial,
  file.path(PATHS$results_dir, "plots", "PCA_initial_all_samples.png"),
  width = config$visualization$plot_width,
  height = config$visualization$plot_height,
  units = config$visualization$plot_units
)

pca_initial
```

### Sample Quality Control and Filtering

```{r sample_filtering}
message("Removing problematic samples...")

# Remove STA_NB_6 sample (intermixed characteristics)
rnaseq_dds_clean <- rnaseq_dds[, rnaseq_dds$cell_line != "STA_NB_6"]
rnaseq_dds_clean$cell_line <- droplevels(rnaseq_dds_clean$cell_line)

message("Samples after filtering: ", ncol(rnaseq_dds_clean))
message("Removed samples: STA_NB_6")

# Re-validate after filtering
validate_inputs(rnaseq_dds_clean, annotation_data)
```

## Differential Expression Analysis

```{r differential_expression}
message("Performing differential expression analysis...")

# Filter lowly expressed genes
rnaseq_dds_filt <- filterDatasets(
  rnaseq_dds_clean,
  abs_filt = TRUE,
  abs_filt_samples = config$filtering$abs_filt_samples
)

# Run DESeq2 pipeline
rnaseq_dds_filt <- DESeq2::estimateSizeFactors(rnaseq_dds_filt)
rnaseq_dds_filt <- DESeq2::DESeq(rnaseq_dds_filt)

# Variance stabilization for visualization
vsd <- DESeq2::vst(rnaseq_dds_filt, blind = TRUE)

# Generate results
message("Available contrasts: ", paste(resultsNames(rnaseq_dds_filt), collapse = ", "))

deg_results <- generateResults(
  dds_object = rnaseq_dds_filt,
  coeff_name = "cell_type_MES_vs_ADR",
  cond_numerator = "MES",
  cond_denominator = "ADR",
  cond_variable = "cell_type",
  ensemblAnnot = annotation_data,
  log2FC_cutoff = config$statistical$log2FC_cutoff
)

# Validate statistical results
stat_validation <- validate_statistical_results(deg_results$results_all, config$statistical$padj_cutoff)
```

### Post-QC PCA Analysis

```{r post_qc_pca}
# Generate PCA after quality control
pca_clean <- generatePCA(
  transf_object = vsd,
  cond_interest_varPart = c("cell_type", "cell_line"),
  color_variable = "cell_type",
  shape_variable = "cell_line",
  ntop_genes = config$visualization$ntop_genes_pca
) +
  scale_color_manual(values = c("navy", "firebrick3")) +
  scale_shape_manual(values = c(16, 7, 17, 15, 3)) +
  labs(title = "PCA: Clean Dataset (After QC)") +
  theme_publication()

# Save plot
safe_save_plot(
  pca_clean,
  file.path(PATHS$results_dir, "plots", "PCA_clean_dataset.png"),
  width = config$visualization$plot_width,
  height = config$visualization$plot_height,
  units = config$visualization$plot_units
)

pca_clean
```

## Gene Set Creation and Export

```{r gene_set_export}
message("Creating MES/ADR gene signatures...")

# Create gene lists for downstream analysis
mes_adr_gene_list <- deg_results$results_signif %>%
  mutate(Term = if_else(log2FoldChange < 0, "ADRN", "MES"))

# Add Entrez IDs if BioMart is available
if (!is.null(mart)) {
  tryCatch({
    biomart_genes <- biomaRt::getBM(
      filters = "ensembl_gene_id",
      attributes = c("ensembl_gene_id", "entrezgene_id"),
      values = mes_adr_gene_list$ensembl_id,
      mart = mart
    )
    
    mes_adr_gene_list <- merge(
      mes_adr_gene_list,
      biomart_genes,
      by.x = "ensembl_id",
      by.y = "ensembl_gene_id"
    ) %>%
    select(Term, entrezgene_id) %>%
    rename(gs_id = Term, gene_id = entrezgene_id) %>%
    filter(!is.na(gene_id))
    
    # Export gene signature
    write.table(
      mes_adr_gene_list,
      file.path(PATHS$resources_dir, "mes_adrn_GS_frm_RNA_seq.tsv"),
      sep = "\t",
      row.names = FALSE,
      quote = FALSE
    )
    
    message("Gene signature exported successfully")
  }, error = function(e) {
    warning("Failed to add Entrez IDs: ", e$message)
  })
}
```

## Results Export

```{r results_export}
message("Exporting results to Excel...")

# Create Excel workbook
xlsx_out <- createWorkbook()

# Add worksheets
addWorksheet(xlsx_out, "Summary")
addWorksheet(xlsx_out, "Significant_DEGs")
addWorksheet(xlsx_out, "DE_Details")
addWorksheet(xlsx_out, "All_Results")

# Prepare summary data
analysis_summary <- generate_analysis_summary(deg_results)
summary_df <- data.frame(
  Metric = names(analysis_summary),
  Value = unlist(analysis_summary),
  stringsAsFactors = FALSE
)

# Write data to worksheets
writeData(xlsx_out, "Summary", summary_df)
writeData(xlsx_out, "Significant_DEGs", deg_results$results_signif)
writeData(xlsx_out, "DE_Details", deg_results$de_details)
writeData(xlsx_out, "All_Results", deg_results$results_all)

# Save workbook
saveWorkbook(
  xlsx_out,
  file.path(PATHS$results_dir, "tables", "MES_vs_ADR_comprehensive_results.xlsx"),
  overwrite = TRUE
)

message("Results exported to Excel successfully")
```

## Gene Set Enrichment Analysis

```{r gsea_analysis}
message("Performing gene set enrichment analysis...")

# Run GSEA for multiple gene sets
gsea_results <- run_gsea_analysis(
  gene_list = deg_results$results_signif$gene_symbol,
  gene_sets = gene_sets,
  background_size = nrow(rnaseq_dds_filt)
)

# Create visualization for each gene set
gsea_plots <- list()

for (set_name in names(gsea_results)) {
  if (!is.null(gsea_results[[set_name]])) {
    tryCatch({
      plot_obj <- hypeR::hyp_dots(
        gsea_results[[set_name]],
        merge = TRUE,
        fdr = config$statistical$padj_cutoff,
        top = config$visualization$top_pathways_show,
        abrv = 70,
        val = "fdr",
        title = paste("GSEA:", set_name)
      ) + theme_publication()
      
      gsea_plots[[set_name]] <- plot_obj
      
      # Save individual plots
      safe_save_plot(
        plot_obj,
        file.path(PATHS$results_dir, "plots", paste0("GSEA_", set_name, ".png")),
        width = config$visualization$plot_width,
        height = config$visualization$plot_height,
        units = config$visualization$plot_units
      )
      
      # Export to Excel
      hypeR::hyp_to_excel(
        gsea_results[[set_name]],
        file_path = file.path(PATHS$results_dir, "tables", paste0("GSEA_", set_name, ".xlsx"))
      )
      
    }, error = function(e) {
      warning("Failed to create plot for ", set_name, ": ", e$message)
    })
  }
}

# Display plots
for (plot_name in names(gsea_plots)) {
  if (!is.null(gsea_plots[[plot_name]])) {
    print(gsea_plots[[plot_name]])
  }
}
```

## Fast Gene Set Enrichment Analysis (fGSEA)

```{r fgsea_analysis}
message("Performing fGSEA analysis...")

# Prepare ranked gene list
deg_ranked <- resolve_duplicate_genes(
  deg_results$results_all,
  method = "highest_basemean"
) %>%
  arrange(desc(log2FoldChange))

ranked_genes <- setNames(deg_ranked$log2FoldChange, deg_ranked$gene_symbol)

# Run fGSEA
fgsea_results <- run_fgsea_batch(
  ranked_genes = ranked_genes,
  gene_sets = gene_sets,
  min_size = config$fgsea$min_size,
  max_size = config$fgsea$max_size,
  top_n = config$fgsea$top_pathways_up
)

# Display results
for (set_name in names(fgsea_results)) {
  if (!is.null(fgsea_results[[set_name]])) {
    message("Displaying fGSEA results for: ", set_name)
    
    # Show GSEA table plot
    print(fgsea_results[[set_name]]$gsea_plot)
    
    # Show dot plot
    print(fgsea_results[[set_name]]$dot_plot)
    
    # Save plots
    safe_save_plot(
      fgsea_results[[set_name]]$dot_plot,
      file.path(PATHS$results_dir, "plots", paste0("fGSEA_dotplot_", set_name, ".png")),
      width = config$visualization$plot_width,
      height = config$visualization$plot_height,
      units = config$visualization$plot_units
    )
  }
}
```

## Hippo Pathway Specific Analysis

```{r hippo_analysis}
message("Performing Hippo pathway specific analysis...")

# Define Hippo-related gene sets
hippo_gene_sets <- list(
  GOBP_Hippo = gene_sets$gobp[["genesets"]][grep("hippo|Hippo", names(gene_sets$gobp[["genesets"]]), value = TRUE)],
  REACTOME_Hippo = gene_sets$reactome[["genesets"]][grep("hippo|Hippo", names(gene_sets$reactome[["genesets"]]), value = TRUE)],
  Cordenonsi_YAP = gene_sets$oncogenic[["genesets"]][grep("Cordenonsi.*Yap", names(gene_sets$oncogenic[["genesets"]]), value = TRUE)],
  Wang_Hippo = gene_sets$wang_hippo[["genesets"]][["wang_hippo_set"]]
)

# Remove empty sets
hippo_gene_sets <- hippo_gene_sets[lengths(hippo_gene_sets) > 0]

if (length(hippo_gene_sets) > 0) {
  # Run fGSEA on Hippo pathways
  hippo_fgsea <- fgsea::fgsea(
    pathways = hippo_gene_sets,
    stats = ranked_genes,
    minSize = config$fgsea$min_size,
    maxSize = config$fgsea$max_size
  )
  
  # Create individual enrichment plots
  for (pathway_name in names(hippo_gene_sets)) {
    tryCatch({
      pathway_genes <- hippo_gene_sets[[pathway_name]]
      pathway_stats <- hippo_fgsea %>% filter(pathway == pathway_name)
      
      if (nrow(pathway_stats) > 0) {
        enrichment_plot <- fgsea::plotEnrichment(
          pathway = pathway_genes,
          stats = ranked_genes
        ) +
        labs(
          title = paste0(pathway_name, "\n",
                        "NES = ", round(pathway_stats$NES, 2), ", ",
                        "padj = ", signif(pathway_stats$padj, 3))
        ) +
        theme_publication()
        
        print(enrichment_plot)
        
        # Save plot
        safe_save_plot(
          enrichment_plot,
          file.path(PATHS$results_dir, "plots", paste0("fGSEA_hippo_", pathway_name, ".png")),
          width = config$visualization$plot_width,
          height = config$visualization$plot_height,
          units = config$visualization$plot_units
        )
      }
    }, error = function(e) {
      warning("Failed to create Hippo plot for ", pathway_name, ": ", e$message)
    })
  }
}
```

## Volcano Plot Visualization

```{r volcano_plot}
message("Creating enhanced volcano plot...")

# Create enhanced volcano plot
volcano_plot <- create_enhanced_volcano(
  results = deg_results$results_all,
  genes_highlight = GENES_OF_INTEREST$all,
  title = "MES vs ADR Differential Expression",
  padj_cutoff = config$statistical$padj_cutoff,
  lfc_cutoff = config$statistical$log2FC_cutoff
)

print(volcano_plot)

# Save volcano plot
safe_save_plot(
  volcano_plot,
  file.path(PATHS$results_dir, "plots", "volcano_plot_enhanced.png"),
  width = config$visualization$plot_width,
  height = config$visualization$plot_height,
  units = config$visualization$plot_units
)

# Create separate volcano plots for different gene categories
for (category in names(GENES_OF_INTEREST)) {
  if (category != "all") {
    category_plot <- create_enhanced_volcano(
      results = deg_results$results_all,
      genes_highlight = GENES_OF_INTEREST[[category]],
      title = paste("MES vs ADR:", str_to_title(gsub("_", " ", category))),
      padj_cutoff = config$statistical$padj_cutoff,
      lfc_cutoff = config$statistical$log2FC_cutoff
    )
    
    safe_save_plot(
      category_plot,
      file.path(PATHS$results_dir, "plots", paste0("volcano_", category, ".png")),
      width = config$visualization$plot_width,
      height = config$visualization$plot_height,
      units = config$visualization$plot_units
    )
  }
}
```

## Analysis Summary and Session Information

```{r analysis_summary}
# Calculate analysis duration
analysis_end_time <- Sys.time()
analysis_duration <- difftime(analysis_end_time, analysis_start_time, units = "mins")

# Generate comprehensive summary
final_summary <- generate_analysis_summary(deg_results, gsea_results)

# Create summary report
summary_report <- data.frame(
  Section = c("Analysis Info", "Data Summary", "Statistical Results", "Output Files"),
  Details = c(
    paste("Duration:", round(analysis_duration, 2), "minutes"),
    paste("Samples:", ncol(rnaseq_dds_clean), "| Genes tested:", final_summary$total_genes_tested),
    paste("Significant DEGs:", final_summary$significant_genes, 
          "(", final_summary$upregulated, "↑,", final_summary$downregulated, "↓)"),
    paste("Files saved to:", PATHS$results_dir)
  )
)

knitr::kable(summary_report, caption = "Analysis Summary")

# Display key statistics
message("=== ANALYSIS COMPLETED ===")
message("Total analysis time: ", round(analysis_duration, 2), " minutes")
message("Significant genes found: ", final_summary$significant_genes)
message("Results saved to: ", PATHS$results_dir)
```

```{r session_info}
# Save session information
session_info <- sessionInfo()
writeLines(
  capture.output(session_info),
  file.path(PATHS$results_dir, "reports", "session_info.txt")
)

# Display session info
session_info
```

---

**Analysis completed at:** `r Sys.time()`  
**Total runtime:** `r round(difftime(Sys.time(), analysis_start_time, units = "mins"), 2)` minutes  
**Output directory:** `r PATHS$results_dir`