---
title: "RNA-seq Analysis - Improved Version"
author: "Aleksandr Bykov (Enhanced by Claude Code)"
date: '`r Sys.Date()`'
output:
  html_document: 
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 8,
  dpi = 300
)

set.seed(42)
analysis_start_time <- Sys.time()
message("Starting RNA-seq analysis at: ", analysis_start_time)
```

## Configuration and Setup

```{r configuration}
config <- list(
  filtering = list(
    abs_filt_samples = 2,
    min_count_threshold = 10
  ),
  statistical = list(
    padj_cutoff = 0.05,
    log2FC_cutoff = 1,
    var_expl_needed = 0.6
  ),
  visualization = list(
    ntop_genes_pca = 1000,
    plot_width = 20,
    plot_height = 20,
    plot_units = "cm",
    top_pathways_show = 20
  ),
  fgsea = list(
    min_size = 15,
    max_size = 500,
    top_pathways_up = 10,
    top_pathways_down = 10
  )
)

PATHS <- list(
  data_dir = here::here("neuroblastoma", "data", "RNAseq"),
  results_dir = here::here("neuroblastoma", "results", "RNA-seq_MES_vs_ADR"),
  resources_dir = here::here("neuroblastoma", "resources")
)

GENES_OF_INTEREST <- list(
  all = c("VIM", "YAP1", "WWTR1", "JUN", "FOSL1", "FOSL2", "PHOX2B", "HAND2", "GATA3"),
  mesenchymal_markers = c("VIM", "CDH2", "FN1"),
  yap_taz_pathway = c("YAP1", "WWTR1", "TEAD1", "TEAD2", "TEAD3", "TEAD4"),
  ap1_factors = c("JUN", "JUNB", "JUND", "FOS", "FOSB", "FOSL1", "FOSL2"),
  adrenergic_markers = c("PHOX2B", "HAND2", "GATA3", "DBH", "TH")
)

print("Configuration loaded successfully")
```

```{r load_libraries}
suppressPackageStartupMessages({
  # Core analysis packages
  library(DESeq2)
  library(stringr)
  library(org.Hs.eg.db)
  
  # Enrichment analysis
  library(fgsea)
  library(hypeR)
  
  # Visualization
  library(ggplot2)
  library(ggrepel)
  library(RColorBrewer)
  library(pheatmap)
  
  # Data manipulation
  library(dplyr)
  library(tidyr)
  library(readr)
  
  # File operations
  library(openxlsx)
  library(here)
  
  # Load all utility functions from consolidated script
  source(here::here("neuroblastoma/resources/UtilityScriptsRNA-seq.R"))
  
})

print("Libraries loaded successfully")
```

```{r setup_environment}
# Setup directories
setup_output_dirs(PATHS$results_dir, c("plots", "tables", "reports"))

# Load gene sets 
message("Loading gene sets from MSigDB...")
gene_sets <- list(
  hallmark = hypeR::msigdb_gsets("Homo sapiens", "H", clean = TRUE),
  kegg = hypeR::msigdb_gsets("Homo sapiens", "C2", "CP:KEGG", clean = TRUE),
  reactome = hypeR::msigdb_gsets("Homo sapiens", "C2", "CP:REACTOME", clean = TRUE),
  gobp = hypeR::msigdb_gsets("Homo sapiens", "C5", "GO:BP", clean = TRUE),
  gocc = hypeR::msigdb_gsets("Homo sapiens", "C5", "GO:CC", clean = TRUE),
  gomf = hypeR::msigdb_gsets("Homo sapiens", "C5", "GO:MF", clean = TRUE),
  oncogenic = hypeR::msigdb_gsets("Homo sapiens", "C6", clean = TRUE),
  wang_hippo = list(genesets = list(wang_hippo_set = c("CCN1", "CCN2", "AMOTL2", "ANKRD1", "IGFBP3", "F3", "FJX1", "NUAK2", "LATS2", "CRIM1", "GADD45A","TGFB2", "PTPN14", "NT5E", "FOXF2", "AXL", "DOCK5", "ASAP1", "RBMS3", "MYOF", "ARHGEF17", "CCDC80")))
)

# # Extract individual gene sets for compatibility
# gs_hallmark <- gene_sets$hallmark
# gs_C2_kegg <- gene_sets$kegg
# gs_C2_reactome <- gene_sets$reactome
# gs_C5_GOBP <- gene_sets$gobp
# gs_C5_GOCC <- gene_sets$gocc
# gs_C5_GOMF <- gene_sets$gomf
# gs_C6_onco <- gene_sets$oncogenic
# gs_wang_hippo <- gene_sets$wang_hippo

# Setup BioMart with error handling
tryCatch({
  options(timeout = 30000)
  mart <- biomaRt::useMart(
    biomart = "ENSEMBL_MART_ENSEMBL", 
    dataset = "hsapiens_gene_ensembl",
    host = "https://www.ensembl.org"
  )
  message("BioMart connection established")
}, error = function(e) {
  warning("Failed to connect to BioMart: ", e$message)
  mart <- NULL
})

print("Environment setup completed")
```

## Data Loading and Validation

```{r load_data}
message("Loading RNA-seq data...")

rnaseq_dds <- readRDS(file.path(PATHS$data_dir, "rnaseq_deseq_global_deseq_data_set.rds"))
design(rnaseq_dds) <- as.formula("~ cell_line + cell_type")
annotation_data <- read.table(
  file = file.path(PATHS$data_dir, "rnaseq_deseq_global_annotation_gene.tsv"),
  sep = "\t",
  header = TRUE
)
# Standardize column names
colnames(annotation_data)[c(5, 7)] <- c("ensembl_id", "gene_symbol")

# Display data summary
message("Data summary:")
message("- Samples: ", ncol(rnaseq_dds))
message("- Genes: ", nrow(rnaseq_dds))
message("- Cell types: ", paste(unique(rnaseq_dds$cell_type), collapse = ", "))
message("- Cell lines: ", paste(unique(rnaseq_dds$cell_line), collapse = ", "))

```

## Quality Control Analysis

### Initial PCA with All Samples

```{r initial_pca}
message("Performing initial quality control analysis...")

# Filter lowly expressed genes
tmp_rnaseq_dds_filt <- filterDatasets(
  rnaseq_dds,
  abs_filt = TRUE,
  abs_filt_samples = config$filtering$abs_filt_samples
)
# Run DESeq2 pipeline
tmp_rnaseq_dds_filt <- DESeq2::estimateSizeFactors(tmp_rnaseq_dds_filt)
tmp_rnaseq_dds_filt <- DESeq2::DESeq(tmp_rnaseq_dds_filt)
vsd_initial <- DESeq2::vst(tmp_rnaseq_dds_filt, blind = TRUE)

# Generate initial PCA
pca_initial <- generatePCA(
  transf_object = vsd_initial,
  cond_interest_varPart = c("cell_type", "cell_line"),
  color_variable = "cell_type",
  shape_variable = "cell_line",
  ntop_genes = config$visualization$ntop_genes_pca
) +
  scale_color_manual(values = c("navy", "firebrick3")) +
  scale_shape_manual(values = c(16, 7, 17, 15, 3)) +
  labs(title = "PCA: All Samples (Before QC)")

# Save plot
ggsave(
  filename = file.path(PATHS$results_dir, "PCA_full_dataset.png"),
  plot = pca_initial,
  width = config$visualization$plot_width,
  height = config$visualization$plot_height,
  units = config$visualization$plot_units
)

pca_initial
```

```

### Sample Quality Control and Filtering

```{r sample_filtering}
message("Removing problematic samples...")

# Remove STA_NB_6 sample (intermixed characteristics)
rnaseq_dds <- rnaseq_dds[, rnaseq_dds$cell_line != "STA_NB_6"]
rnaseq_dds$cell_line <- droplevels(rnaseq_dds$cell_line)

message("Samples after filtering: ", ncol(rnaseq_dds))
message("Removed samples: STA_NB_6")

# Validate inputs after filtering
validate_inputs(rnaseq_dds, annotation_data)

# Filter lowly expressed genes
rnaseq_dds_filt <- filterDatasets(
  rnaseq_dds,
  abs_filt = TRUE,
  abs_filt_samples = config$filtering$abs_filt_samples
)
# Run DESeq2 pipeline
rnaseq_dds_filt <- DESeq2::estimateSizeFactors(rnaseq_dds_filt)
rnaseq_dds_filt <- DESeq2::DESeq(rnaseq_dds_filt)

# Variance stabilization for visualization
vsd <- DESeq2::vst(rnaseq_dds_filt, blind = TRUE)

# Generate PCA after quality control
pca_clean <- generatePCA(
  transf_object = vsd,
  cond_interest_varPart = c("cell_type", "cell_line"),
  color_variable = "cell_type",
  shape_variable = "cell_line",
  ntop_genes = config$visualization$ntop_genes_pca
) +
  scale_color_manual(values = c("navy", "firebrick3")) +
  labs(title = "PCA: Clean Dataset (After QC)") +
  theme_publication()

# Save plot using safe function
safe_save_plot(
  pca_clean,
  file.path(PATHS$results_dir, "plots", "PCA_clean_dataset.png"),
  width = config$visualization$plot_width,
  height = config$visualization$plot_height,
  units = config$visualization$plot_units
)

pca_clean
```

### Differential Gene Expression

```{r deg_analysis}

# Generate differential expression results
message("Available contrasts: ", paste(resultsNames(rnaseq_dds_filt), collapse = ", "))

deg_results <- generateResults(
  dds_object = rnaseq_dds_filt,
  coeff_name = "cell_type_MES_vs_ADR",
  cond_numerator = "MES",
  cond_denominator = "ADR",
  cond_variable = "cell_type",
  ensemblAnnot = annotation_data,
  log2FC_cutoff = config$statistical$log2FC_cutoff
)

# Validate statistical results
stat_validation <- validate_statistical_results(deg_results$results_all, config$statistical$padj_cutoff)

```

## Gene Set Creation and Export

```{r gene_set_export}
message("Creating MES/ADR gene signatures...")

# Create gene lists for downstream analysis
mes_adrn_gene_list <- deg_results$results_signif %>%
  mutate(Term = if_else(log2FoldChange < 0, "ADRN", "MES"))

# Add Entrez IDs if BioMart is available
if (!is.null(mart)) {
  tryCatch({
    biomart_genes <- biomaRt::getBM(
      filters = "ensembl_gene_id",
      attributes = c("ensembl_gene_id", "entrezgene_id"),
      values = mes_adrn_gene_list$ensembl_id,
      mart = mart
    )
    
    mes_adrn_gene_list <- merge(
      mes_adrn_gene_list,
      biomart_genes,
      by.x = "ensembl_id",
      by.y = "ensembl_gene_id"
    ) %>%
    select(Term, entrezgene_id) %>%
    rename(gs_id = Term, gene_id = entrezgene_id) %>%
    filter(!is.na(gene_id))
    
    # Export gene signature
    write.table(
      mes_adrn_gene_list,
      file.path(PATHS$resources_dir, "mes_adrn_GS_frm_RNA_seq.tsv"),
      sep = "\t",
      row.names = FALSE,
      quote = FALSE
    )
    
    message("Gene signature exported successfully")
  }, error = function(e) {
    warning("Failed to add Entrez IDs: ", e$message)
  })
}

```

## Results Export

```{r results_export}
message("Exporting results to Excel...")

# Create Excel workbook
xlsx_out <- createWorkbook()

# Add worksheets
addWorksheet(xlsx_out, "Summary")
addWorksheet(xlsx_out, "Significant_DEGs")
addWorksheet(xlsx_out, "DE_Details")
addWorksheet(xlsx_out, "All_Results")

# Prepare summary data
analysis_summary <- generate_analysis_summary(deg_results)
summary_df <- data.frame(
  Metric = names(analysis_summary),
  Value = unlist(analysis_summary),
  stringsAsFactors = FALSE
)

# Write data to worksheets
writeData(xlsx_out, "Summary", summary_df)
writeData(xlsx_out, "Significant_DEGs", deg_results$results_signif)
writeData(xlsx_out, "DE_Details", deg_results$de_details)
writeData(xlsx_out, "All_Results", deg_results$results_all)

# Save workbook
saveWorkbook(
  xlsx_out,
  file.path(PATHS$results_dir, "tables", "MES_vs_ADR_comprehensive_results.xlsx"),
  overwrite = TRUE
)

message("Results exported to Excel successfully")

```

```{r GSEA, echo=FALSE}
# Produce GSEA plots
C5_GOBP <- hypeR::hypeR(
  signature = deg_results$results_signif$gene_symbol,
  genesets = gs_C5_GOBP,
  test = "hypergeometric",
  background = nrow(rnaseq_dds_filt)
)
C5_GOCC <- hypeR::hypeR(
  signature = deg_results$results_signif$gene_symbol,
  genesets = gs_C5_GOCC,
  test = "hypergeometric",
  background = nrow(rnaseq_dds_filt)
)
C5_GOMF <- hypeR::hypeR(
  signature = deg_results$results_signif$gene_symbol,
  genesets = gs_C5_GOMF,
  test = "hypergeometric",
  background = nrow(rnaseq_dds_filt)
)
C2_kegg <- hypeR::hypeR(
  signature = deg_results$results_signif$gene_symbol,
  genesets = gs_C2_kegg,
  test = "hypergeometric",
  background = nrow(rnaseq_dds_filt)
)
C2_reactome <- hypeR::hypeR(
  signature = deg_results$results_signif$gene_symbol,
  genesets = gs_C2_reactome,
  test = "hypergeometric",
  background = nrow(rnaseq_dds_filt)
)
C6_onco <- hypeR::hypeR(
  signature = deg_results$results_signif$gene_symbol,
  genesets = gs_C6_onco,
  test = "hypergeometric",
  background = nrow(rnaseq_dds_filt)
)
wang_hippo <- hypeR::hypeR(
  signature = deg_results$results_signif$gene_symbol,
  genesets = gs_wang_hippo,
  test = "hypergeometric",
  background = nrow(rnaseq_dds_filt)
)

C5_GOBP_plot <- hypeR::hyp_dots(C5_GOBP, merge = TRUE, fdr = 0.05, top = 20, abrv = 70, val = "fdr", title = "GOBP: MES vs ADR") + theme_bw()
C5_GOCC_plot <- hypeR::hyp_dots(C5_GOCC, merge = TRUE, fdr = 0.05, top = 20, abrv = 70, val = "fdr", title = "GOCC: MES vs ADR") + theme_bw()
C5_GOMF_plot <- hypeR::hyp_dots(C5_GOMF, merge = TRUE, fdr = 0.05, top = 20, abrv = 70, val = "fdr", title = "GOMF: MES vs ADR") + theme_bw()
C2_kegg_plot <- hypeR::hyp_dots(C2_kegg, merge = TRUE, fdr = 0.05, top = 20, abrv = 70, val = "fdr", title = "KEGG: MES vs ADR") + theme_bw()
C2_rctm_plot <- hypeR::hyp_dots(C2_reactome, merge = TRUE, fdr = 0.05, top = 20, abrv = 70, val = "fdr", title = "REACTOME: MES vs ADR") + theme_bw()
C6_onco_plot <- hypeR::hyp_dots(C6_onco, merge = TRUE, fdr = 0.05, top = 20, abrv = 70, val = "fdr", title = "ONCO: MES vs ADR") + theme_bw()

plot(C5_GOBP_plot)
plot(C5_GOCC_plot)
plot(C5_GOMF_plot)
plot(C2_kegg_plot)
plot(C2_rctm_plot)

# optional saving to excel tables
hypeR::hyp_to_excel(C5_GOBP, file_path=file.path(deg_dir, "MES_vs_ADRN_GSEA_C5_GOBP.xlsx"))
hypeR::hyp_to_excel(C2_reactome, file_path=file.path(deg_dir, "MES_vs_ADRN_GSEA_C2_reactome.xlsx"))
hypeR::hyp_to_excel(C6_onco, file_path=file.path(deg_dir, "MES_vs_ADRN_GSEA_C6_onco.xlsx"))
# hypeR::hyp_to_excel(C5_GOCC, file_path=file.path(deg_dir, "MES_vs_ADRN_GSEA_C5_GOCC.xlsx"))
# hypeR::hyp_to_excel(C5_GOMF, file_path=file.path(deg_dir, "MES_vs_ADRN_GSEA_C5_GOMF.xlsx"))

# saving the report to a PDF file
pdf(file = file.path(deg_dir, "MES_vs_ADR_GSEA.pdf"), width = 10, height = 10)
plot(C5_GOBP_plot)
plot(C5_GOCC_plot)
plot(C5_GOMF_plot)
plot(C2_kegg_plot)
plot(C2_rctm_plot)
dev.off()

# Additionally - save as individual plots
ggsave(
  filename = paste0(deg_dir, "MES_vs_ADRN_GSEA_C5_GOBP.png"),
  plot = C5_GOBP_plot,
  width = 20, height = 20, units = "cm"
)
ggsave(
  filename = paste0(deg_dir, "MES_vs_ADRN_GSEA_C5_GOCC.png"),
  plot = C5_GOCC_plot,
  width = 20, height = 20, units = "cm"
)
ggsave(
  filename = paste0(deg_dir, "MES_vs_ADRN_GSEA_C5_GOMF.png"),
  plot = C5_GOMF_plot,
  width = 20, height = 20, units = "cm"
)
ggsave(
  filename = paste0(deg_dir, "MES_vs_ADRN_GSEA_C2_KEGG.png"),
  plot = C2_kegg_plot,
  width = 20, height = 20, units = "cm"
)
ggsave(
  filename = paste0(deg_dir, "MES_vs_ADRN_GSEA_C2_REACTOME.png"),
  plot = C2_rctm_plot,
  width = 20, height = 20, units = "cm"
)
```

Produce fgsea plots

```{r fgsea}
# Run FastGSEA for GOBP, GOCC, GOMF, KEGG, reactome terms
# Rank genes based on L2FC
# The slice_max is necessary to remove duplicated gene names (which are 6 in total). 
# The reason why they are present is that originally the DEG is done on Transcript IDs which can have identical 
# gene_symbol. And duplicated values are disrupting the fgsea, so they must be removed for a correct work.
# There is no ideal way to deal with this problem, so according to advice from the creator of fgsea
# https://www.biostars.org/p/9591990/#9591997 , The gene with the highest BaseMean kept.

de_genes_ranked <- deg_results$results_all
de_genes_ranked <- de_genes_ranked %>%
  group_by(gene_symbol) %>%
  slice_max(baseMean, n = 1, with_ties = FALSE) %>%
  ungroup()
de_genes_ranked <- de_genes_ranked %>% arrange(desc(log2FoldChange))
de_genes_ranked <- setNames(c(de_genes_ranked$log2FoldChange), c(de_genes_ranked$gene_symbol))

gene_set_list <- list(
  gs_hallmark = gs_hallmark,
  gs_C2_kegg = gs_C2_kegg,
  gs_C2_reactome = gs_C2_reactome,
  gs_C5_GOBP = gs_C5_GOBP,
  gs_C5_GOCC = gs_C5_GOCC,
  gs_C5_GOMF = gs_C5_GOMF
)

# pdf(file = file.path(deg_dir, "MES_vs_ADR_fGSEA_and_dotPLOT2.pdf"), width = 10, height = 10)
for (gene_set_name in names(gene_set_list)) {
  # gene_set_name <- "gs_hallmark"
  print(gene_set_name)
  gene_set <- gene_set_list[[gene_set_name]]

  fgseaRes <- fgsea(
    pathways = gene_set[["genesets"]],
    stats = de_genes_ranked,
    minSize = 15,
    maxSize = 500
  )
  topPathwaysUp <- fgseaRes[ES > 0][head(order(pval), n = 10), pathway]
  topPathwaysDown <- fgseaRes[ES < 0][head(order(pval), n = 10), pathway]
  topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

  p <- plotGseaTable(gene_set[["genesets"]][topPathways],
    de_genes_ranked,
    fgseaRes,
    gseaParam = 0.5
  ) +
    labs(title = gene_set_name)
  p[["theme"]][["plot.title"]] <- NULL
  print(p)

  # Dot plot
  fgseaRes_dot <- fgseaRes %>%
    select(pathway, NES, padj) %>%
    filter(pathway %in% topPathways) %>%
    arrange(NES) %>%
    mutate(pathway = factor(pathway, levels = pathway)) %>%
    mutate(padj_log2 = -log2(padj))

  p <- ggplot(fgseaRes_dot) +
    aes(x = NES, y = pathway, colour = NES, size = padj_log2) +
    geom_point(shape = "circle") +
    scale_color_distiller(palette = "RdBu", direction = -1) +
    scale_size(range = c(4, 10)) +
    theme_minimal() +
    labs(
      title = gene_set_name,
      size = "-log2p_adj"
    )
  print(p)
}
# dev.off()

```

Separate fgsea plots for hippo pathway 
```{r}
# Additional separate fsGSEAs for hippos
gene_set_list <- list(
  GOBP_Hippo_Signaling = gs_C5_GOBP[["genesets"]][["Hippo Signaling"]],
  REACTOME_Signaling_By_Hippo = gs_C2_reactome[["genesets"]][["Signaling By Hippo"]],
  C6_onko_Cordenonsi_Yap_Conserved_Signature = gs_C6_onco[["genesets"]][["Cordenonsi Yap Conserved Signature"]],
  Hippo_Wang = gs_wang_hippo$wang_hippo_set
)

fgseaRes <- fgsea(
  pathways = gene_set_list,
  stats = de_genes_ranked,
  minSize = 15,
  maxSize = 500
)

#pdf(file = file.path(deg_dir, "MES_vs_ADR_fGSEA_hippos.pdf"), width = 10, height = 8)
for(gene_set_to_plot_name in names(gene_set_list)){
 # gene_set_to_plot_name <- "Cordenonsi_Yap_Conserved_Signature"
  gene_set_to_plot <- gene_set_list[[gene_set_to_plot_name]]
  
  title_values <- fgseaRes %>% filter(pathway == gene_set_to_plot_name) %>% select(pathway, NES, padj)
  p <- plotEnrichment(pathway = gene_set_to_plot, stats = de_genes_ranked) + 
    labs(title = paste(title_values$pathway, "\n",
                       "NES =", title_values$NES, "\n",
                       "padj = ", title_values$padj))
  plot(p)
  
ggsave(
  filename = paste0(deg_dir, "fgsea_",gene_set_to_plot_name, ".pdf"),
  plot = p,
  width = 20, height = 20, units = "cm"
)
}
#dev.off()


```

Volcano plot for selected proteins
```{r volcano plot}
# Create a volcano plot
# Genes to highlight
genes_to_highlight <- c(
  "VIM",
  "YAP1",
  "WWTR1",
  "JUN",
  "FOSL1",
  "FOSL2",
  "PHOX2B",
  "HAND2",
  "GATA3"
)
volcano_plot <- plotVolcano(
  dds_results_obj = deg_results$results_all,
  genes_of_interest = genes_to_highlight,
  plot_title = "MES vs ADRN genes"
)
plot(volcano_plot)

ggsave(
  filename = paste0(deg_dir, "MES_vs_ADRN_volcano_plot.png"),
  plot = volcano_plot,
  width = 20, height = 20, units = "cm"
)

```
