---
title: "shRNA-seq Analysis (merged) - Improved Version"
author: "Aleksandr Bykov (Enhanced by Claude Code)"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

<!--
  This document merges two previously separate shRNA-seq analyses:
    * shRNA-seq_analysis.Rmd          - DESeq2-based dox vs ctrl comparisons + ssGSEA
    * 01_RNA_seq_shRNA_analysis.Rmd   - per-pair (no-replicate) fGSEA + GSVA barplot
  All file paths have been corrected to point at the current project layout
  (~/neuroblastoma/...). Dead links from the older draft (~/workspace/... and the
  non-existent cell_type_MES_vs_ADR.xlsx) have been replaced with the working
  MES_vs_ADR_comprehensive_results.xlsx source.
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 8,
  dpi = 300
)

set.seed(42)
analysis_start_time <- Sys.time()
message("Starting merged RNA-seq shRNA analysis at: ", analysis_start_time)
```

## Configuration and Setup

```{r configuration}
config <- list(
  filtering = list(
    abs_filt_samples = 2,
    min_count_threshold = 10
  ),
  statistical = list(
    padj_cutoff = 0.05,
    log2FC_cutoff = 1,
    var_expl_needed = 0.6
  ),
  visualization = list(
    ntop_genes_pca = 1000,
    plot_width = 20,
    plot_height = 20,
    plot_units = "cm",
    top_pathways_show = 20
  ),
  fgsea = list(
    min_size = 15,
    max_size = 500
  )
)

# NOTE: data lives under data/RNAseq (no hyphen); results under results/RNA-seq.
PATHS <- list(
  data_dir      = "~/neuroblastoma/data/RNAseq",
  results_dir   = "~/neuroblastoma/results/RNA-seq/RNA-seq_shRNA",
  pairwise_dir  = "~/neuroblastoma/results/RNA-seq/RNA-seq_shRNA_pairwise",
  resources_dir = "~/neuroblastoma/resources"
)

# Working reference to the MES/ADR signature table (replaces the old dead links).
MES_ADR_FILE <- "~/neuroblastoma/results/RNA-seq/RNA-seq_MES_vs_ADR/tables/MES_vs_ADR_comprehensive_results.xlsx"

# Working reference to the k975 YAP/TAZ inhibition 48h DE table.
K975_48H_FILE <- "~/neuroblastoma/results/RNA-seq/RNA-seq_yap_taz_inhibition/cell_type_48h_vs_control.xlsx"

print("Configuration loaded successfully")
```

```{r load_libraries}
suppressPackageStartupMessages({
  # Core analysis packages
  library(DESeq2)
  library(stringr)
  library(org.Hs.eg.db)

  # Enrichment analysis
  library(fgsea)
  library(hypeR)
  library(GSVA)

  # Visualization
  library(ggplot2)
  library(ggrepel)
  library(pheatmap)
  library(RColorBrewer)
  library(reshape2)

  # Data manipulation
  library(dplyr)
  library(tidyr)
  library(readr)

  # Batch correction
  library(limma)

  # File operations
  library(openxlsx)
  library(openxlsx2)
  library(here)

  # Load all utility functions from consolidated script
  source("~/neuroblastoma/resources/UtilityScriptsRNA-seq.R")
})

print("Libraries loaded successfully")
```

```{r setup_environment}
if(!dir.exists(PATHS$results_dir))  dir.create(PATHS$results_dir, recursive = TRUE)
if(!dir.exists(PATHS$pairwise_dir)) dir.create(PATHS$pairwise_dir, recursive = TRUE)

# Load gene sets directly (no caching to avoid repository bloat)
message("Loading gene sets from MSigDB...")
gs_hallmark    <- hypeR::msigdb_gsets(species = "Homo sapiens", "H", clean = TRUE)
gs_C2_reactome <- hypeR::msigdb_gsets(species = "Homo sapiens", "C2", "CP:REACTOME", clean = TRUE)
gs_C5_GOBP     <- hypeR::msigdb_gsets(species = "Homo sapiens", "C5", "GO:BP", clean = TRUE)
gs_C6_onco     <- hypeR::msigdb_gsets(species = "Homo sapiens", "C6", clean = TRUE)
gs_wang_hippo  <- list(wang_hippo_set = c("CCN1", "CCN2", "AMOTL2", "ANKRD1", "IGFBP3", "F3", "FJX1", "NUAK2", "LATS2", "CRIM1", "GADD45A",
                                          "TGFB2", "PTPN14", "NT5E", "FOXF2", "AXL", "DOCK5", "ASAP1", "RBMS3", "MYOF", "ARHGEF17", "CCDC80"))

print("Environment setup completed")
```

```{r load_shared_data}
message("Loading shRNA-seq data and annotation...")

annotation_data <- read.table(
  file = file.path(PATHS$data_dir, "rnaseq_deseq_global_annotation_gene.tsv"),
  sep = "\t",
  header = TRUE
)
colnames(annotation_data)[c(5, 7)] <- c("ensembl_id", "gene_symbol")

# MES/ADR signatures derived from our RNA-seq (MES vs ADRN comprehensive results, sheet 2).
# This replaces the old, non-working "cell_type_MES_vs_ADR.xlsx" links.
RNA_SEQ_data <- openxlsx2::read_xlsx(MES_ADR_FILE, sheet = 2)
mes_adrn_gene_list <- RNA_SEQ_data %>%
  mutate(Term = if_else(log2FoldChange < 0, "ADRN", "MES"))
sig_list_our_data <- list(
  Adrenergic  = mes_adrn_gene_list$gene_symbol[mes_adrn_gene_list$Term == "ADRN"],
  Mesenchymal = mes_adrn_gene_list$gene_symbol[mes_adrn_gene_list$Term == "MES"]
)
print("Shared data loaded successfully")
```

# Part 1 - DESeq2 dox vs ctrl comparisons

*(from shRNA-seq_analysis.Rmd)*

## 1a. Combined dox vs ctrl (all pairs)

All six samples (LUC/TGFP/YAP x ctrl/dox) modelled together with a single
dox-vs-ctrl contrast.

```{r combined_load_data}
message("Loading shRNA-seq data (all pairs)...")
load(file.path(PATHS$data_dir, "deseq2_shRNA.dds.RData"))

dds_reduced <- dds
# Sample order is ctrl/dox alternating across the three shRNA groups.
dds_reduced$Group3 <- factor(c("ctrl", "dox", "ctrl", "dox", "ctrl", "dox"), levels = c("ctrl", "dox"))
dds_reduced$Group1 <- factor(c(dds_reduced$Group1), levels = unique(dds_reduced$Group1))
design(dds_reduced) <- as.formula("~ Group3")
```

```{r combined_deseq}
# Filtering lowly expressed genes
dds_reduced_filt <- filterDatasets(dds_reduced,
                                   abs_filt = TRUE,
                                   abs_filt_samples = config$filtering$abs_filt_samples
)

# Estimate size factors and run DESeq2
dds_reduced_filt <- DESeq2::estimateSizeFactors(dds_reduced_filt)
dds_reduced_filt <- DESeq2::DESeq(dds_reduced_filt)
dds_reduced_filt_vsd <- DESeq2::vst(dds_reduced_filt, blind = TRUE) # blind = TRUE for QC

resultsNames(dds_reduced_filt)
```

```{r combined_get_degs, include=FALSE}
deg_results <- generateResults(
  dds_object = dds_reduced_filt,
  coeff_name = "Group3_dox_vs_ctrl",
  cond_numerator = "dox",
  cond_denominator = "ctrl",
  cond_variable = "Group3",
  ensemblAnnot = annotation_data,
  log2FC_cutoff = config$statistical$log2FC_cutoff
)

# Write the result to the xlsx file
XLSX_OUT <- openxlsx::createWorkbook()
openxlsx::addWorksheet(XLSX_OUT, "results_signif")
openxlsx::addWorksheet(XLSX_OUT, "de_details")
openxlsx::addWorksheet(XLSX_OUT, "results_all")
openxlsx::writeData(XLSX_OUT, x = deg_results$results_signif, sheet = "results_signif")
openxlsx::writeData(XLSX_OUT, x = deg_results$de_details, sheet = "de_details")
openxlsx::writeData(XLSX_OUT, x = deg_results$results_all, sheet = "results_all")
openxlsx::saveWorkbook(XLSX_OUT, file.path(PATHS$results_dir, "dox_vs_ctrl_all_pairs.xlsx"), overwrite = TRUE)
```

```{r combined_fgsea}
gene_set_list <- list(
  GOBP_Hippo_Signaling = gs_C5_GOBP[["genesets"]][["Hippo Signaling"]],
  REACTOME_Signaling_By_Hippo = gs_C2_reactome[["genesets"]][["Signaling By Hippo"]],
  C6_onko_Cordenonsi_Yap_Conserved_Signature = gs_C6_onco[["genesets"]][["Cordenonsi Yap Conserved Signature"]],
  Hippo_Wang = gs_wang_hippo$wang_hippo_set
)

deg_results_ranked <- deg_results$results_all
deg_results_ranked <- deg_results_ranked %>%
  group_by(gene_symbol) %>%
  slice_max(baseMean, n = 1, with_ties = FALSE) %>%
  ungroup()
deg_results_ranked <- deg_results_ranked %>% arrange(desc(log2FoldChange))
deg_results_ranked <- setNames(c(deg_results_ranked$log2FoldChange),
                               c(deg_results_ranked$gene_symbol))

fgseaRes_dox_vs_ctrl <- fgsea(
  pathways = gene_set_list,
  stats = deg_results_ranked,
  minSize = 15,
  maxSize = 2500
)

for(gene_set_to_plot_name in names(gene_set_list)){
  gene_set_to_plot <- gene_set_list[[gene_set_to_plot_name]]

  title_values <- fgseaRes_dox_vs_ctrl %>% dplyr::filter(pathway == gene_set_to_plot_name) %>% dplyr::select(pathway, NES, padj)
  p <- plotEnrichment(pathway = gene_set_to_plot, stats = deg_results_ranked) +
    labs(title = paste(title_values$pathway, "\n",
                       "NES =", title_values$NES, "\n",
                       "padj = ", title_values$padj))

  plot(p)

  ggsave(
    filename = file.path(PATHS$results_dir, paste0("dox_vs_ctrl_all_pairs_fgsea_", gene_set_to_plot_name, ".pdf")),
    plot = p,
    width = 20, height = 20, units = "cm"
  )
}
```

```{r combined_gsva}
# prepare vsd matrix with gene symbols
vsd_counts_matrix <- assay(dds_reduced_filt_vsd)
gene_symbols <- annotation_data[, 'gene_symbol'][match(rownames(vsd_counts_matrix),
                                                       annotation_data[, 'ensembl_id'])]
# Drop unmapped symbols and collapse duplicates to the highest-mean row;
# ssgseaParam() requires unique, non-NA rownames.
keep <- !is.na(gene_symbols) & gene_symbols != ""
vsd_counts_matrix <- vsd_counts_matrix[keep, ]
gene_symbols <- gene_symbols[keep]
keep_idx <- order(rowMeans(vsd_counts_matrix), decreasing = TRUE)
keep_idx <- keep_idx[!duplicated(gene_symbols[keep_idx])]
vsd_counts_matrix <- as.matrix(vsd_counts_matrix[keep_idx, ])
rownames(vsd_counts_matrix) <- gene_symbols[keep_idx]

# Score MES/ADR terms (defined from our own RNA-seq profiles)
ssgsea_param <- GSVA::ssgseaParam(
  vsd_counts_matrix,
  sig_list_our_data,
  minSize = 1,
  maxSize = Inf,
  normalize = TRUE
)
ssgsea_mes_adr_cellines <- GSVA::gsva(
  ssgsea_param,
  verbose = TRUE,
  BPPARAM = BiocParallel::MulticoreParam(2)
)
ssgsea_mes_adr_heatmap <- pheatmap::pheatmap(ssgsea_mes_adr_cellines,
                                             scale = "row",
                                             cluster_rows = TRUE,
                                             cluster_cols = FALSE,
                                             color = colorRampPalette(c("navy", "white", "firebrick3"))(50),
                                             show_colnames = TRUE)
```

```{r combined_plot_gsva_heatmap}
ssgsea_mes_adr_heatmap
```

## 1b. YAP shRNA vs control shRNA (dox-treated)

Restricted to the three dox-treated samples, contrasting the YAP-shRNA sample
against the two control-shRNA (LUC, TGFP) dox samples. **Verify the sample order
matches the `c("ctrl", "ctrl", "dox")` labelling for your dataset.**

```{r yap_contrast_load_data}
load(file.path(PATHS$data_dir, "deseq2_shRNA.dds.RData"))
dds_reduced <- dds[, c(2, 4, 6)]
dds_reduced$Group3 <- factor(c("ctrl", "ctrl", "dox"), levels = c("ctrl", "dox"))
design(dds_reduced) <- as.formula("~ Group3")
```

```{r yap_contrast_deseq}
dds_reduced_filt <- filterDatasets(dds_reduced,
                                   abs_filt = TRUE,
                                   abs_filt_samples = config$filtering$abs_filt_samples
)

dds_reduced_filt <- DESeq2::estimateSizeFactors(dds_reduced_filt)
dds_reduced_filt <- DESeq2::DESeq(dds_reduced_filt)
dds_reduced_filt_vsd <- DESeq2::vst(dds_reduced_filt, blind = TRUE)

resultsNames(dds_reduced_filt)
```

```{r yap_contrast_get_degs, include=FALSE}
deg_results <- generateResults(
  dds_object = dds_reduced_filt,
  coeff_name = "Group3_dox_vs_ctrl",
  cond_numerator = "dox",
  cond_denominator = "ctrl",
  cond_variable = "Group3",
  ensemblAnnot = annotation_data,
  log2FC_cutoff = config$statistical$log2FC_cutoff
)

XLSX_OUT <- openxlsx::createWorkbook()
openxlsx::addWorksheet(XLSX_OUT, "results_signif")
openxlsx::addWorksheet(XLSX_OUT, "de_details")
openxlsx::addWorksheet(XLSX_OUT, "results_all")
openxlsx::writeData(XLSX_OUT, x = deg_results$results_signif, sheet = "results_signif")
openxlsx::writeData(XLSX_OUT, x = deg_results$de_details, sheet = "de_details")
openxlsx::writeData(XLSX_OUT, x = deg_results$results_all, sheet = "results_all")
openxlsx::saveWorkbook(XLSX_OUT, file.path(PATHS$results_dir, "YAP_shRNA_vs_control_shRNA_dox.xlsx"), overwrite = TRUE)
```

```{r yap_contrast_fgsea}
gene_set_list <- list(
  GOBP_Hippo_Signaling = gs_C5_GOBP[["genesets"]][["Hippo Signaling"]],
  REACTOME_Signaling_By_Hippo = gs_C2_reactome[["genesets"]][["Signaling By Hippo"]],
  C6_onko_Cordenonsi_Yap_Conserved_Signature = gs_C6_onco[["genesets"]][["Cordenonsi Yap Conserved Signature"]],
  Hippo_Wang = gs_wang_hippo$wang_hippo_set
)

deg_results_ranked <- deg_results$results_all
deg_results_ranked <- deg_results_ranked %>%
  group_by(gene_symbol) %>%
  slice_max(baseMean, n = 1, with_ties = FALSE) %>%
  ungroup()
deg_results_ranked <- deg_results_ranked %>% arrange(desc(log2FoldChange))
deg_results_ranked <- setNames(c(deg_results_ranked$log2FoldChange),
                               c(deg_results_ranked$gene_symbol))

fgseaRes_yap_contrast <- fgsea(
  pathways = gene_set_list,
  stats = deg_results_ranked,
  minSize = 15,
  maxSize = 2500
)

for(gene_set_to_plot_name in names(gene_set_list)){
  gene_set_to_plot <- gene_set_list[[gene_set_to_plot_name]]

  title_values <- fgseaRes_yap_contrast %>% dplyr::filter(pathway == gene_set_to_plot_name) %>% dplyr::select(pathway, NES, padj)
  p <- plotEnrichment(pathway = gene_set_to_plot, stats = deg_results_ranked) +
    labs(title = paste(title_values$pathway, "\n",
                       "NES =", title_values$NES, "\n",
                       "padj = ", title_values$padj))

  plot(p)

  ggsave(
    filename = file.path(PATHS$results_dir, paste0("YAP_shRNA_vs_control_shRNA_dox_fgsea_", gene_set_to_plot_name, ".pdf")),
    plot = p,
    width = 20, height = 20, units = "cm"
  )
}
```

```{r yap_contrast_gsva}
vsd_counts_matrix <- assay(dds_reduced_filt_vsd)
gene_symbols <- annotation_data[, 'gene_symbol'][match(rownames(vsd_counts_matrix),
                                                       annotation_data[, 'ensembl_id'])]
# Drop unmapped symbols and collapse duplicates to the highest-mean row;
# ssgseaParam() requires unique, non-NA rownames.
keep <- !is.na(gene_symbols) & gene_symbols != ""
vsd_counts_matrix <- vsd_counts_matrix[keep, ]
gene_symbols <- gene_symbols[keep]
keep_idx <- order(rowMeans(vsd_counts_matrix), decreasing = TRUE)
keep_idx <- keep_idx[!duplicated(gene_symbols[keep_idx])]
vsd_counts_matrix <- as.matrix(vsd_counts_matrix[keep_idx, ])
rownames(vsd_counts_matrix) <- gene_symbols[keep_idx]

ssgsea_param <- GSVA::ssgseaParam(
  vsd_counts_matrix,
  sig_list_our_data,
  minSize = 1,
  maxSize = Inf,
  normalize = TRUE
)
ssgsea_mes_adr_cellines <- GSVA::gsva(
  ssgsea_param,
  verbose = TRUE,
  BPPARAM = BiocParallel::MulticoreParam(2)
)
ssgsea_mes_adr_heatmap_yap <- pheatmap::pheatmap(ssgsea_mes_adr_cellines,
                                                 scale = "row",
                                                 cluster_rows = TRUE,
                                                 cluster_cols = FALSE,
                                                 color = colorRampPalette(c("navy", "white", "firebrick3"))(50),
                                                 show_colnames = TRUE)
```

```{r yap_contrast_plot_gsva_heatmap}
ssgsea_mes_adr_heatmap_yap
```

# Part 2 - Per-pair fGSEA (no replicates)

*(from 01_RNA_seq_shRNA_analysis.Rmd)*

Each DOX vs CTRL pair (LUC, TGFP, YAP) is analysed individually using a
per-gene log2 fold-change ranking, since these pairs have no replicates.

```{r pairwise_analysis}
# Reload the raw shRNA dds (Part 1 mutated dds_reduced in place).
load(file.path(PATHS$data_dir, "deseq2_shRNA.dds.RData"))

# Examine the data structure
message("Examining data structure...")
sample_info <- as.data.frame(colData(dds))
sample_info$sample_name <- rownames(sample_info)
print(sample_info)

# Filter lowly expressed genes (keep genes with >= 10 counts in >= 2 samples)
keep <- rowSums(counts(dds) >= 10) >= 2
dds_filtered <- dds[keep, ]
dds_filtered <- estimateSizeFactors(dds_filtered)
normalized_counts <- counts(dds_filtered, normalized = TRUE)

# Map ensembl ids to gene symbols; keep highest-mean gene per duplicated symbol.
gene_symbols <- annotation_data$gene_symbol[match(rownames(normalized_counts), annotation_data$ensembl_id)]
gene_symbols[is.na(gene_symbols)] <- rownames(normalized_counts)[is.na(gene_symbols)]

mean_expr <- rowMeans(normalized_counts)
df_temp <- data.frame(
  ensembl_id  = rownames(normalized_counts),
  gene_symbol = gene_symbols,
  mean_expr   = mean_expr,
  stringsAsFactors = FALSE
) %>%
  group_by(gene_symbol) %>%
  slice_max(mean_expr, n = 1, with_ties = FALSE) %>%
  ungroup()

normalized_counts <- normalized_counts[df_temp$ensembl_id, ]
rownames(normalized_counts) <- df_temp$gene_symbol

# k975 YAP/TAZ inhibition 48h up/down gene sets (working link).
k975_48h_results <- openxlsx::read.xlsx(K975_48H_FILE, sheet = 1)
k975_48h_up   <- k975_48h_results %>% dplyr::filter(log2FoldChange > 0) %>% dplyr::pull(gene_symbol)
k975_48h_down <- k975_48h_results %>% dplyr::filter(log2FoldChange < 0) %>% dplyr::pull(gene_symbol)

# Gene sets for the per-pair fGSEA.
gene_set_list_pairwise <- list(
  Hallmark_EMT = gs_hallmark[["genesets"]][["Epithelial Mesenchymal Transition"]],
  Hallmark_Hippo = gs_hallmark[["genesets"]][["Hippo Signaling"]],
  GOBP_Hippo_Signaling = gs_C5_GOBP[["genesets"]][["Hippo Signaling"]],
  REACTOME_Signaling_By_Hippo = gs_C2_reactome[["genesets"]][["Signaling By Hippo"]],
  C6_onko_Cordenonsi_Yap_Conserved_Signature = gs_C6_onco[["genesets"]][["Cordenonsi Yap Conserved Signature"]],
  Hippo_Wang = gs_wang_hippo$wang_hippo_set,
  k975_48h_down = k975_48h_down,
  k975_48h_up = k975_48h_up
)

# --- helper: log2FC + fGSEA for a single CTRL/DOX pair ---------------------
analyze_sample_pair <- function(ctrl_sample, dox_sample, pair_name, normalized_counts, gene_sets) {
  message(paste("Analyzing pair:", pair_name))

  ctrl_counts <- normalized_counts[, ctrl_sample]
  dox_counts  <- normalized_counts[, dox_sample]

  pseudocount <- 1
  log2fc <- log2((dox_counts + pseudocount) / (ctrl_counts + pseudocount))

  valid_genes <- is.finite(log2fc) & !is.na(log2fc)
  log2fc <- log2fc[valid_genes]

  set.seed(42)  # break ties reproducibly
  log2fc <- log2fc + runif(length(log2fc), -1e-8, 1e-8)
  log2fc_sorted <- sort(log2fc, decreasing = TRUE)

  fgsea_results <- fgsea(
    pathways = gene_sets,
    stats = log2fc_sorted,
    minSize = 15,
    maxSize = 2500
  )
  fgsea_results$pair <- pair_name
  fgsea_results$ctrl_sample <- ctrl_sample
  fgsea_results$dox_sample <- dox_sample

  list(fgsea_results = fgsea_results, log2fc_ranked = log2fc_sorted, pair_name = pair_name)
}

# --- helper: enrichment plots for every pathway of a pair ------------------
create_enrichment_plots <- function(analysis_result, gene_sets, output_dir) {
  fgsea_res <- analysis_result$fgsea_results
  log2fc_ranked <- analysis_result$log2fc_ranked
  pair_name <- analysis_result$pair_name

  if(nrow(fgsea_res) > 0) {
    for(i in 1:nrow(fgsea_res)) {
      pathway_name  <- fgsea_res$pathway[i]
      pathway_genes <- gene_sets[[pathway_name]]

      if(!is.null(pathway_genes) && length(pathway_genes) > 0) {
        tryCatch({
          p <- plotEnrichment(pathway_genes, log2fc_ranked) +
            labs(
              title = paste(pair_name, "-", pathway_name),
              subtitle = paste("NES =", round(fgsea_res$NES[i], 3),
                               "| padj =", format(fgsea_res$padj[i], scientific = TRUE, digits = 3))
            ) +
            theme_minimal()
          filename <- file.path(output_dir, paste0(pair_name, "_", gsub("[^A-Za-z0-9]", "_", pathway_name), "_enrichment.pdf"))
          ggsave(filename, p, width = 10, height = 8)
        }, error = function(e) {
          message(paste("Warning: Could not create plot for", pathway_name, ":", e$message))
        })
      }
    }
  } else {
    message(paste("No pathways found for", pair_name))
  }
}

# --- identify CTRL/DOX pairs ----------------------------------------------
sample_pairs <- list()
if("Group1" %in% colnames(sample_info)) {
  for(group in unique(sample_info$Group1)) {
    group_samples <- sample_info[sample_info$Group1 == group, ]
    if(nrow(group_samples) == 2) {
      sample_names <- rownames(group_samples)
      ctrl_idx <- grep("ctrl|control|CTRL|CONTROL", sample_names, ignore.case = TRUE)
      dox_idx  <- grep("dox|DOX|treat|TREAT", sample_names, ignore.case = TRUE)
      if(length(ctrl_idx) == 0) ctrl_idx <- 1
      if(length(dox_idx) == 0)  dox_idx <- 2
      sample_pairs[[paste0(group, "_DOX_vs_CTRL")]] <- list(
        ctrl = sample_names[ctrl_idx[1]],
        dox  = sample_names[dox_idx[1]]
      )
    }
  }
}
if(length(sample_pairs) == 0) {
  # Fallback: assume 1,2 = LUC; 3,4 = TGFP; 5,6 = YAP (ctrl, dox).
  message("Using fallback sample pairing based on position...")
  sample_names <- colnames(dds_filtered)
  if(length(sample_names) >= 6) {
    sample_pairs <- list(
      "LUC_DOX_vs_CTRL"  = list(ctrl = sample_names[1], dox = sample_names[2]),
      "TGFP_DOX_vs_CTRL" = list(ctrl = sample_names[3], dox = sample_names[4]),
      "YAP_DOX_vs_CTRL"  = list(ctrl = sample_names[5], dox = sample_names[6])
    )
  }
}
message("Identified sample pairs:")
print(sample_pairs)

# --- run analysis per pair -------------------------------------------------
all_results <- list()
combined_fgsea <- data.frame()

for(pair_name in names(sample_pairs)) {
  pair_info <- sample_pairs[[pair_name]]
  result <- analyze_sample_pair(
    ctrl_sample = pair_info$ctrl,
    dox_sample  = pair_info$dox,
    pair_name   = pair_name,
    normalized_counts = normalized_counts,
    gene_sets   = gene_set_list_pairwise
  )
  all_results[[pair_name]] <- result

  combined_fgsea <- if(nrow(combined_fgsea) == 0) result$fgsea_results else rbind(combined_fgsea, result$fgsea_results)

  create_enrichment_plots(result, gene_set_list_pairwise, PATHS$pairwise_dir)

  fgsea_for_excel <- result$fgsea_results
  fgsea_for_excel$leadingEdge <- sapply(fgsea_for_excel$leadingEdge, function(x) paste(x, collapse = "; "))
  openxlsx2::write_xlsx(
    fgsea_for_excel,
    file.path(PATHS$pairwise_dir, paste0(pair_name, "_fgsea_results.xlsx"))
  )
}

# Save combined results
combined_fgsea_for_excel <- combined_fgsea
if(nrow(combined_fgsea_for_excel) > 0) {
  combined_fgsea_for_excel$leadingEdge <- sapply(combined_fgsea_for_excel$leadingEdge, function(x) paste(x, collapse = "; "))
}
openxlsx2::write_xlsx(combined_fgsea_for_excel, file.path(PATHS$pairwise_dir, "combined_fgsea_results.xlsx"))
```

```{r pairwise_summary}
# Summary heatmap of NES values across pairs (significant pathways only).
heatmap_data <- combined_fgsea %>%
  select(pathway, pair, NES, padj) %>%
  filter(padj < 0.05) %>%
  dcast(pathway ~ pair, value.var = "NES", fill = 0)

if(nrow(heatmap_data) > 1 && ncol(heatmap_data) > 2) {
  rownames(heatmap_data) <- heatmap_data$pathway
  heatmap_data <- heatmap_data[, -1, drop = FALSE]

  pheatmap(
    heatmap_data,
    cluster_rows = nrow(heatmap_data) > 1,
    cluster_cols = ncol(heatmap_data) > 1,
    color = colorRampPalette(c("blue", "white", "red"))(100),
    main = "NES Heatmap - Significant Pathways (padj < 0.05)",
    filename = file.path(PATHS$pairwise_dir, "fgsea_NES_heatmap.pdf"),
    width = 10,
    height = 8
  )
  message("Heatmap created successfully")
} else {
  message("Insufficient data for heatmap generation (need at least 2 significant pathways)")
}

# Summary table
summary_table <- combined_fgsea %>%
  filter(padj < 0.05) %>%
  arrange(pair, padj) %>%
  select(pair, pathway, NES, pval, padj, size)
openxlsx2::write_xlsx(summary_table, file.path(PATHS$pairwise_dir, "significant_pathways_summary.xlsx"))
```

```{r pairwise_gsva}
# GSVA / ssGSEA for MES/ADR signatures across all samples.
message("Running GSVA analysis for MES/ADR signatures...")

tryCatch({
  dds_for_gsva <- dds_filtered
  dds_for_gsva <- DESeq(dds_for_gsva)
  vsd_for_gsva <- vst(dds_for_gsva, blind = TRUE)

  vsd_counts_matrix <- assay(vsd_for_gsva)
  gene_symbols_gsva <- annotation_data$gene_symbol[match(rownames(vsd_counts_matrix), annotation_data$ensembl_id)]
  gene_symbols_gsva[is.na(gene_symbols_gsva)] <- rownames(vsd_counts_matrix)[is.na(gene_symbols_gsva)]

  df_temp_gsva <- data.frame(
    ensembl_id  = rownames(vsd_counts_matrix),
    gene_symbol = gene_symbols_gsva,
    mean_expr   = rowMeans(vsd_counts_matrix),
    stringsAsFactors = FALSE
  ) %>%
    group_by(gene_symbol) %>%
    slice_max(mean_expr, n = 1, with_ties = FALSE) %>%
    ungroup()

  vsd_counts_matrix <- vsd_counts_matrix[df_temp_gsva$ensembl_id, ]
  rownames(vsd_counts_matrix) <- df_temp_gsva$gene_symbol

  ssgsea_param <- GSVA::ssgseaParam(
    as.matrix(vsd_counts_matrix),
    sig_list_our_data,
    minSize = 1,
    maxSize = Inf,
    normalize = TRUE
  )
  ssgsea_mes_adr <- GSVA::gsva(
    ssgsea_param,
    verbose = TRUE,
    BPPARAM = BiocParallel::MulticoreParam(2)
  )

  pheatmap::pheatmap(
    ssgsea_mes_adr,
    scale = "row",
    cluster_rows = TRUE,
    cluster_cols = FALSE,
    color = colorRampPalette(c("navy", "white", "firebrick3"))(50),
    show_colnames = TRUE,
    main = "MES/ADR Signatures - ssGSEA Scores",
    filename = file.path(PATHS$pairwise_dir, "gsva_mes_adr_signatures_heatmap.pdf"),
    width = 10,
    height = 6
  )

  gsva_scores_df <- as.data.frame(t(ssgsea_mes_adr))
  gsva_scores_df$sample <- rownames(gsva_scores_df)
  gsva_scores_df$treatment <- sapply(strsplit(gsva_scores_df$sample, "_"), function(x) x[1])
  gsva_scores_df$condition <- sapply(strsplit(gsva_scores_df$sample, "_"), function(x) x[2])
  openxlsx2::write_xlsx(gsva_scores_df, file.path(PATHS$pairwise_dir, "gsva_mes_adr_scores.xlsx"))

  gsva_long <- gsva_scores_df %>%
    select(-sample) %>%
    pivot_longer(cols = c("Adrenergic", "Mesenchymal"), names_to = "signature", values_to = "score")

  p_gsva <- ggplot(gsva_long, aes(x = paste(treatment, condition, sep = "_"), y = score, fill = signature)) +
    geom_bar(stat = "identity", position = "dodge") +
    theme_minimal() +
    labs(
      title = "MES/ADR Signature Scores by Treatment",
      x = "Sample", y = "ssGSEA Score", fill = "Signature"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  print(p_gsva)

  ggsave(
    filename = file.path(PATHS$pairwise_dir, "gsva_mes_adr_barplot.pdf"),
    plot = p_gsva,
    width = 10, height = 6
  )
  message("GSVA analysis completed successfully")
}, error = function(e) {
  message(paste("Warning: Could not complete GSVA analysis:", e$message))
})
```

```{r session_summary}
message("Analysis complete!")
message(paste("DESeq2 comparison results saved to:", PATHS$results_dir))
message(paste("Per-pair fGSEA results saved to:", PATHS$pairwise_dir))
message(paste("Number of sample pairs analyzed:", length(sample_pairs)))
message(paste("Total significant pathways (padj < 0.05):", nrow(summary_table)))
message(paste("Elapsed:", format(Sys.time() - analysis_start_time)))
```
