---
title: "RNA-seq Integration Analysis - Full Dataset"
author: "Aleksandr Bykov (Enhanced by Claude Code)"
date: '`r Sys.Date()`'
output:
  html_document: 
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 8,
  dpi = 300
)

set.seed(42)
analysis_start_time <- Sys.time()
message("Starting RNA-seq integration analysis at: ", analysis_start_time)
```

## Configuration and Setup

```{r configuration}
config <- list(
  filtering = list(
    abs_filt_samples = 2,
    min_count_threshold = 10
  ),
  statistical = list(
    padj_cutoff = 0.05,
    log2FC_cutoff = 1,
    var_expl_needed = 0.6
  ),
  visualization = list(
    ntop_genes_pca = 1000,
    plot_width = 20,
    plot_height = 20,
    plot_units = "cm",
    top_pathways_show = 20
  ),
  fgsea = list(
    min_size = 15,
    max_size = 500,
    top_pathways_up = 10,
    top_pathways_down = 10
  )
)

PATHS <- list(
  data_dir = here::here("neuroblastoma", "data", "RNAseq"),
  results_dir = here::here("neuroblastoma", "results", "RNA-seq_TDI"),
  resources_dir = here::here("neuroblastoma", "resources"),
  cache_dir = here::here("neuroblastoma", "cache")
)

GENES_OF_INTEREST <- list(
  all = c("VIM", "YAP1", "WWTR1", "JUN", "FOSL1", "FOSL2", "PHOX2B", "HAND2", "GATA3"),
  mesenchymal_markers = c("VIM", "CDH2", "FN1"),
  yap_taz_pathway = c("YAP1", "WWTR1", "TEAD1", "TEAD2", "TEAD3", "TEAD4"),
  ap1_factors = c("JUN", "JUNB", "JUND", "FOS", "FOSB", "FOSL1", "FOSL2"),
  adrenergic_markers = c("PHOX2B", "HAND2", "GATA3", "DBH", "TH")
)

print("Configuration loaded successfully")
```

```{r load_libraries}
suppressPackageStartupMessages({
  # Core analysis packages
  library(DESeq2)
  library(stringr)
  library(org.Hs.eg.db)
  
  # Enrichment analysis
  library(fgsea)
  library(hypeR)
  library(GSVA)
  
  # Visualization
  library(ggplot2)
  library(ggrepel)
  library(RColorBrewer)
  library(pheatmap)
  
  # Data manipulation
  library(dplyr)
  library(tidyr)
  library(readr)
  
  # File operations
  library(openxlsx)
  library(here)
  
  # Utility functions from existing scripts
  import::from(
    .from = here::here("neuroblastoma/resources/UtilityScriptsRNA-seq.R"),
    "filterDatasets", "generatePCA", "generateEnsemblAnnotation", 
    "generateResults", "meanExprsPerGroup", "plotVolcano",
    .character_only = TRUE
  )
})

print("Libraries loaded successfully")
```

```{r setup_environment}
if(!dir.exists(PATHS$results_dir)) dir.create(PATHS$results_dir, recursive = TRUE)

gs_hallmark <- hypeR::msigdb_gsets(species = "Homo sapiens", category = c("H"), clean = TRUE)
gs_C2_kegg <- hypeR::msigdb_gsets(species = "Homo sapiens", category = c("C2"), subcategory = "CP:KEGG", clean = TRUE)
gs_C2_reactome <- hypeR::msigdb_gsets(species = "Homo sapiens", category = c("C2"), subcategory = "CP:REACTOME", clean = TRUE)
gs_C5_GOBP <- hypeR::msigdb_gsets(species = "Homo sapiens", category = c("C5"), subcategory = "GO:BP", clean = TRUE)
gs_C5_GOCC <- hypeR::msigdb_gsets(species = "Homo sapiens", category = c("C5"), subcategory = "GO:CC", clean = TRUE)
gs_C5_GOMF <- hypeR::msigdb_gsets(species = "Homo sapiens", category = c("C5"), subcategory = "GO:MF", clean = TRUE)
gs_C6_onco <- hypeR::msigdb_gsets(species = "Homo sapiens", category = c("C6"), clean = TRUE)
gs_wang_hippo <- list(wang_hippo_set = c("CCN1", "CCN2", "AMOTL2", "ANKRD1", "IGFBP3", "F3", "FJX1", "NUAK2", "LATS2", "CRIM1", "GADD45A",
                                         "TGFB2", "PTPN14", "NT5E", "FOXF2", "AXL", "DOCK5", "ASAP1", "RBMS3", "MYOF", "ARHGEF17", "CCDC80"))

print("Environment setup completed")
```

## Data Loading and Preparation

```{r load_data}
message("Loading RNA-seq integration data...")

annotationData <- read.table(
  file = file.path(PATHS$data_dir, "rnaseq_deseq_global_annotation_gene.tsv"),
  sep = "\t",
  header = TRUE
)
colnames(annotationData)[c(5, 7)] <- c("ensembl_id", "gene_symbol")

load(file.path(PATHS$data_dir, "OE_WWTR1_mCherry_JunDN_experiment.dds.RData"))

samples_to_select_full <- c("OE_WWTR1_DMSO_RNA_S165191_R1", 
                            "OE_WWTR1_TDI_RNA_S165188_R1", 
                            "OE_WWTR1_dox_RNA_S165189_R1", 
                            "OE_WWTR1_dox_TDI_RNA_S165178_R1",
                            "OE_mCherrry_TDI_RNA_S165184_R1", 
                            "OE_mCherry_DMSO_RNA_S165187_R1", 
                            "OE_mCherry_dox_RNA_S165185_R1", 
                            "OE_mCherry_dox_TDI_RNA_S165190_R1",
                            "OE_WWTR1_DMSO_RNA_S165191_R2", 
                            "OE_WWTR1_TDI_RNA_S165188_R2", 
                            "OE_WWTR1_dox_RNA_S165189_R2", 
                            "OE_WWTR1_dox_TDI_RNA_S165178_R2",
                            "OE_mCherrry_TDI_RNA_S165184_R2", 
                            "OE_mCherry_DMSO_RNA_S165187_R2", 
                            "OE_mCherry_dox_RNA_S165185_R2", 
                            "OE_mCherry_dox_TDI_RNA_S165190_R2")

samples_to_select_TDI_vs_DMSO <- c("OE_WWTR1_DMSO_RNA_S165191_R1", 
                                   "OE_WWTR1_TDI_RNA_S165188_R1", 
                                   "OE_mCherrry_TDI_RNA_S165184_R1", 
                                   "OE_mCherry_DMSO_RNA_S165187_R1", 
                                   "OE_WWTR1_DMSO_RNA_S165191_R2", 
                                   "OE_WWTR1_TDI_RNA_S165188_R2", 
                                   "OE_mCherrry_TDI_RNA_S165184_R2", 
                                   "OE_mCherry_DMSO_RNA_S165187_R2")

print("Data loaded successfully")
```

## TDI vs DMSO Analysis (Combined)

### Data Processing and Sample Annotation

```{r process_tdi_dmso}
message("Processing TDI vs DMSO analysis...")

TDI_dds <- dds[,colnames(dds) %in% samples_to_select_TDI_vs_DMSO]
sample_names <- colData(TDI_dds)$sample

parsed <- str_match(
  sample_names,
  "^OE_([^_]+)_((?:[^_]+_)*[^_]+)_RNA_.*_(R\\d+)$"
)
parsed[,2][parsed[, 2] == "mCherrry"] <- "mCherry"

parsed_df <- data.frame(
  genotype  = parsed[, 2],
  treatment = parsed[, 3],
  replicate = parsed[, 4],
  stringsAsFactors = TRUE
)

colData(TDI_dds) <- cbind(colData(TDI_dds), parsed_df)
TDI_dds$treatment <- factor(TDI_dds$treatment, levels = c("DMSO", "TDI"))
design(TDI_dds) <- as.formula("~ genotype + treatment")

message("Sample metadata parsed and design formula set")
```

### Quality Control Analysis

```{r qc_analysis}
TDI_dds_filt <- filterDatasets(TDI_dds,
                               abs_filt = TRUE,
                               abs_filt_samples = config$filtering$abs_filt_samples)

TDI_dds_filt <- DESeq2::estimateSizeFactors(TDI_dds_filt)
TDI_dds_filt <- DESeq2::DESeq(TDI_dds_filt)
TDI_dds_filt_vsd <- DESeq2::vst(TDI_dds_filt, blind = TRUE)

tmp_pca_deg <- generatePCA(
  transf_object = TDI_dds_filt_vsd,
  cond_interest_varPart = c("genotype", "treatment"),
  color_variable = "genotype",
  shape_variable = "treatment",
  ntop_genes = config$visualization$ntop_genes_pca
) +
  scale_color_manual(values = c("navy", "firebrick3")) +
  scale_shape_manual(values = c(16, 7)) +
  ggtitle("RNA-seq TDI dataset")

ggsave(
  filename = file.path(PATHS$results_dir, "PCA_TDI_vs_DMSO_combined.png"),
  plot = tmp_pca_deg,
  width = config$visualization$plot_width,
  height = config$visualization$plot_height,
  units = config$visualization$plot_units
)

tmp_pca_deg
```

### Differential Expression Analysis

```{r differential_expression}
resultsNames(TDI_dds_filt)

deg_results <- generateResults(
  dds_object = TDI_dds_filt,
  coeff_name = "treatment_TDI_vs_DMSO",
  cond_numerator = "TDI",
  cond_denominator = "DMSO",
  cond_variable = "treatment",
  ensemblAnnot = annotationData,
  log2FC_cutoff = config$statistical$log2FC_cutoff
)

deg_results_wwtr_mcherry <- deg_results

XLSX_OUT <- openxlsx::createWorkbook()
openxlsx::addWorksheet(XLSX_OUT, "results_signif")
openxlsx::addWorksheet(XLSX_OUT, "de_details")
openxlsx::addWorksheet(XLSX_OUT, "results_all")
openxlsx::writeData(XLSX_OUT, x = deg_results$results_signif, sheet = "results_signif")
openxlsx::writeData(XLSX_OUT, x = deg_results$de_details, sheet = "de_details")
openxlsx::writeData(XLSX_OUT, x = deg_results$results_all, sheet = "results_all")
openxlsx::saveWorkbook(XLSX_OUT, file.path(PATHS$results_dir, "treatment_TDI_vs_DMSO.xlsx"), overwrite = T)

print(paste("Significant genes found:", nrow(deg_results$results_signif)))
```

## WWTR1-Only Analysis

### Data Processing

```{r wwtr1_only_analysis}
message("Processing WWTR1-only analysis...")

samples_to_select_TDI_vs_DMSO_WWTR1 <- c("OE_WWTR1_DMSO_RNA_S165191_R1", 
                                          "OE_WWTR1_TDI_RNA_S165188_R1", 
                                          "OE_WWTR1_DMSO_RNA_S165191_R2", 
                                          "OE_WWTR1_TDI_RNA_S165188_R2")

TDI_dds_WWTR1 <- dds[,colnames(dds) %in% samples_to_select_TDI_vs_DMSO_WWTR1]
sample_names <- colData(TDI_dds_WWTR1)$sample

parsed <- str_match(
  sample_names,
  "^OE_([^_]+)_((?:[^_]+_)*[^_]+)_RNA_.*_(R\\d+)$"
)

parsed_df <- data.frame(
  genotype  = parsed[, 2],
  treatment = parsed[, 3],
  replicate = parsed[, 4],
  stringsAsFactors = TRUE
)

colData(TDI_dds_WWTR1) <- cbind(colData(TDI_dds_WWTR1), parsed_df)
TDI_dds_WWTR1$treatment <- factor(TDI_dds_WWTR1$treatment, levels = c("DMSO", "TDI"))
design(TDI_dds_WWTR1) <- as.formula("~ treatment")

TDI_dds_filt_WWTR1 <- filterDatasets(TDI_dds_WWTR1,
                                      abs_filt = TRUE,
                                      abs_filt_samples = config$filtering$abs_filt_samples)

TDI_dds_filt_WWTR1 <- DESeq2::estimateSizeFactors(TDI_dds_filt_WWTR1)
TDI_dds_filt_WWTR1 <- DESeq2::DESeq(TDI_dds_filt_WWTR1)
TDI_dds_filt_vsd_WWTR1 <- DESeq2::vst(TDI_dds_filt_WWTR1, blind = TRUE)

deg_results_WWTR1 <- generateResults(
  dds_object = TDI_dds_filt_WWTR1,
  coeff_name = "treatment_TDI_vs_DMSO",
  cond_numerator = "TDI",
  cond_denominator = "DMSO",
  cond_variable = "treatment",
  ensemblAnnot = annotationData,
  log2FC_cutoff = config$statistical$log2FC_cutoff
)

write.table(
  x = deg_results_WWTR1$results_signif,
  file = file.path(PATHS$results_dir, "deg_results_wwtr.tsv"),
  sep = "\t",
  row.names = F,
  quote = F
)

print(paste("WWTR1-only significant genes found:", nrow(deg_results_WWTR1$results_signif)))
```

## Multi-condition Analysis (Full WWTR1 Dataset)

### Data Preparation for Complex Design

```{r multi_condition_analysis}
message("Processing multi-condition analysis...")

samples_to_select_full_WWTR1 <- c("OE_WWTR1_DMSO_RNA_S165191_R1", 
                                  "OE_WWTR1_TDI_RNA_S165188_R1", 
                                  "OE_WWTR1_dox_RNA_S165189_R1", 
                                  "OE_WWTR1_dox_TDI_RNA_S165178_R1",
                                  "OE_WWTR1_DMSO_RNA_S165191_R2", 
                                  "OE_WWTR1_TDI_RNA_S165188_R2", 
                                  "OE_WWTR1_dox_RNA_S165189_R2", 
                                  "OE_WWTR1_dox_TDI_RNA_S165178_R2")

TDI_dds_full <- dds[,colnames(dds) %in% samples_to_select_full_WWTR1]
sample_names <- colData(TDI_dds_full)$sample

parsed <- str_match(
  sample_names,
  "^OE_([^_]+)_((?:[^_]+_)*[^_]+)_RNA_.*_(R\\d+)$"
)

parsed_df <- data.frame(
  genotype  = parsed[, 2],
  treatment = parsed[, 3],
  replicate = parsed[, 4],
  stringsAsFactors = TRUE
)

colData(TDI_dds_full) <- cbind(colData(TDI_dds_full), parsed_df)
TDI_dds_full$treatment <- factor(TDI_dds_full$treatment, levels = c("DMSO", "TDI", "dox", "dox_TDI"))
design(TDI_dds_full) <- as.formula("~ treatment")

TDI_dds_filt_full <- filterDatasets(TDI_dds_full,
                                    abs_filt = TRUE,
                                    abs_filt_samples = config$filtering$abs_filt_samples)

TDI_dds_filt_full <- DESeq2::estimateSizeFactors(TDI_dds_filt_full)
TDI_dds_filt_full <- DESeq2::DESeq(TDI_dds_filt_full)
```

### TDI vs DMSO Comparison

```{r tdi_dmso_comparison}
deg_results_TDI_vs_DMSO <- generateResults(
  dds_object = TDI_dds_filt_full,
  coeff_name = "treatment_TDI_vs_DMSO",
  cond_numerator = "TDI",
  cond_denominator = "DMSO",
  cond_variable = "treatment",
  ensemblAnnot = annotationData,
  log2FC_cutoff = config$statistical$log2FC_cutoff
)

XLSX_OUT <- openxlsx::createWorkbook()
openxlsx::addWorksheet(XLSX_OUT, "results_signif")
openxlsx::addWorksheet(XLSX_OUT, "de_details")
openxlsx::addWorksheet(XLSX_OUT, "results_all")
openxlsx::writeData(XLSX_OUT, x = deg_results_TDI_vs_DMSO$results_signif, sheet = "results_signif")
openxlsx::writeData(XLSX_OUT, x = deg_results_TDI_vs_DMSO$de_details, sheet = "de_details")
openxlsx::writeData(XLSX_OUT, x = deg_results_TDI_vs_DMSO$results_all, sheet = "results_all")
openxlsx::saveWorkbook(XLSX_OUT, file.path(PATHS$results_dir, "treatment_TDI_vs_DMSO_full.xlsx"), overwrite = T)
```

### doxTDI vs dox Comparison

```{r dox_tdi_comparison}
deg_results_doxTDI_vs_dox <- generateResults(
  dds_object = TDI_dds_filt_full,
  coef_contrast = c("treatment", "dox_TDI", "dox"),
  cond_numerator = "dox_TDI",
  cond_denominator = "dox",
  cond_variable = "treatment",
  ensemblAnnot = annotationData,
  log2FC_cutoff = config$statistical$log2FC_cutoff
)

XLSX_OUT <- openxlsx::createWorkbook()
openxlsx::addWorksheet(XLSX_OUT, "results_signif")
openxlsx::addWorksheet(XLSX_OUT, "de_details")
openxlsx::addWorksheet(XLSX_OUT, "results_all")
openxlsx::writeData(XLSX_OUT, x = deg_results_doxTDI_vs_dox$results_signif, sheet = "results_signif")
openxlsx::writeData(XLSX_OUT, x = deg_results_doxTDI_vs_dox$de_details, sheet = "de_details")
openxlsx::writeData(XLSX_OUT, x = deg_results_doxTDI_vs_dox$results_all, sheet = "results_all")
openxlsx::saveWorkbook(XLSX_OUT, file.path(PATHS$results_dir, "treatment_doxTDI_vs_dox.xlsx"), overwrite = T)
```

## Hippo Pathway-Specific Enrichment Analysis

```{r hippo_pathway_analysis}
message("Performing Hippo pathway-specific analysis...")

gene_set_list <- list(
  GOBP_Hippo_Signaling = gs_C5_GOBP[["genesets"]][["Hippo Signaling"]],
  REACTOME_Signaling_By_Hippo = gs_C2_reactome[["genesets"]][["Signaling By Hippo"]],
  C6_onko_Cordenonsi_Yap_Conserved_Signature = gs_C6_onco[["genesets"]][["Cordenonsi Yap Conserved Signature"]],
  Hippo_Wang = gs_wang_hippo$wang_hippo_set
)

# Function to perform fGSEA analysis
run_fgsea_analysis <- function(deg_results, gene_sets, title_prefix) {
  de_genes_ranked <- deg_results %>%
    group_by(gene_symbol) %>%
    slice_max(baseMean, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    arrange(desc(log2FoldChange))
  
  de_genes_ranked <- setNames(de_genes_ranked$log2FoldChange, de_genes_ranked$gene_symbol)
  
  fgseaRes <- fgsea(
    pathways = gene_sets,
    stats = de_genes_ranked,
    minSize = config$fgsea$min_size,
    maxSize = config$fgsea$max_size
  )
  
  for(gene_set_to_plot_name in names(gene_sets)){
    gene_set_to_plot <- gene_sets[[gene_set_to_plot_name]]
    
    title_values <- fgseaRes %>% 
      filter(pathway == gene_set_to_plot_name) %>% 
      select(pathway, NES, padj)
    
    if(nrow(title_values) > 0) {
      p <- plotEnrichment(pathway = gene_set_to_plot, stats = de_genes_ranked) + 
        labs(title = paste(title_prefix, "\n",
                           title_values$pathway, "\n",
                           "NES =", round(title_values$NES, 2), "\n",
                           "padj = ", signif(title_values$padj, 3)))
      
      ggsave(
        filename = file.path(PATHS$results_dir, paste0(title_prefix, "_fgsea_", gene_set_to_plot_name, ".pdf")),
        plot = p,
        width = config$visualization$plot_width, 
        height = config$visualization$plot_height, 
        units = config$visualization$plot_units
      )
      
      print(p)
    }
  }
  
  return(fgseaRes)
}

# Run fGSEA for TDI vs DMSO
fgsea_TDI_vs_DMSO <- run_fgsea_analysis(
  deg_results_TDI_vs_DMSO$results_all, 
  gene_set_list, 
  "TDI_vs_DMSO"
)

# Run fGSEA for doxTDI vs dox
fgsea_doxTDI_vs_dox <- run_fgsea_analysis(
  deg_results_doxTDI_vs_dox$results_all, 
  gene_set_list, 
  "doxTDI_vs_dox"
)
```

## MES/ADR Signature Analysis

```{r mes_adr_analysis}
message("Performing MES/ADR signature analysis...")

# Load MES/ADR signatures
RNA_SEQ_data <- openxlsx2::read_xlsx(file.path(PATHS$results_dir, "..", "RNA-seq", "cell_type_MES_vs_ADR.xlsx"), sheet = 1)
mes_adrn_gene_list <- RNA_SEQ_data %>%
  mutate(Term = if_else(log2FoldChange < 0, "ADRN", "MES"))

sig_list_our_data <- list(
  Aderenergic = mes_adrn_gene_list$gene_symbol[mes_adrn_gene_list$Term == "ADRN"],
  Mesenchymal = mes_adrn_gene_list$gene_symbol[mes_adrn_gene_list$Term == "MES"]
)

# GSVA analysis
vsd_counts_matrix <- assay(TDI_dds_filt_vsd)
row.names(vsd_counts_matrix) <- annotationData[, 'gene_symbol'][match(row.names(vsd_counts_matrix), annotationData[, 'ensembl_id'])]

ssgsea_mes_adr_cellines <- GSVA::gsva(vsd_counts_matrix,
                                      sig_list_our_data,
                                      method = c("ssgsea"),
                                      min.sz = 1, max.sz = Inf, 
                                      ssgsea.norm = TRUE, verbose = TRUE, parallel.sz = 10)

ssgsea_mes_adr_ncc_noradr_heatmap <- pheatmap::pheatmap(ssgsea_mes_adr_cellines,
                                                        scale = "row",
                                                        cluster_rows = TRUE,
                                                        cluster_cols = TRUE,
                                                        color = colorRampPalette(c("navy", "white", "firebrick3"))(50),
                                                        show_colnames = TRUE,
                                                        main = "MES/ADR Signature Analysis")

ggsave(
  filename = file.path(PATHS$results_dir, "ssgsea_mes_adr_heatmap.pdf"),
  plot = ssgsea_mes_adr_ncc_noradr_heatmap,
  width = config$visualization$plot_width,
  height = config$visualization$plot_height / 2,
  units = config$visualization$plot_units
)

ssgsea_mes_adr_ncc_noradr_heatmap
```

## Integration with Original Dataset

```{r integration_analysis}
message("Integrating with original dataset...")

samples_to_select_OG <- c("SH_A_RNA_S130494","SH_M_RNA_S130497")
rnaseq_OG <- readRDS(file.path(PATHS$data_dir, "rnaseq_deseq_global_deseq_data_set.rds"))
rnaseq_OG <- rnaseq_OG[,colnames(rnaseq_OG) %in% samples_to_select_OG]

# Combine count data
TDI_dds_counts <- as.data.frame(TDI_dds_full@assays@data@listData$counts)
rnaseq_OG_counts <- as.data.frame(rnaseq_OG@assays@data@listData$counts)

TDI_dds_counts$gene_id <- rownames(TDI_dds_full)
rnaseq_OG_counts$gene_id <- rownames(rnaseq_OG)

TDI_plus_OG_counts <- merge.data.frame(TDI_dds_counts, rnaseq_OG_counts, by = "gene_id")
row.names(TDI_plus_OG_counts) <- TDI_plus_OG_counts$gene_id
TDI_plus_OG_counts <- select(TDI_plus_OG_counts, !gene_id)

# Create metadata
parsed <- str_match(
  colnames(TDI_plus_OG_counts),
  "^OE_([^_]+)_((?:[^_]+_)*[^_]+)_RNA_.*_(R\\d+)$"
)
parsed[,2][parsed[, 2] == "mCherrry"] <- "mCherry"

# Manual addition of control samples
parsed[c(9,10), 1] <- colnames(TDI_plus_OG_counts)[c(9,10)]
parsed[c(9,10), 2] <- c("outside_control_ADR", "outside_control_MES")
parsed[c(9,10), 3] <- c("ADR", "MES")
parsed[c(9,10), 4] <- "R1"

parsed_df <- data.frame(
  genotype  = as.factor(parsed[, 2]),
  treatment = as.factor(parsed[, 3]),
  replicate = as.factor(parsed[, 4]),
  stringsAsFactors = TRUE
)
parsed_df$is_outer_control <- as.factor(c(rep("not_control", 8), rep("control", 2)))

TDI_plus_OG <- DESeqDataSetFromMatrix(countData = TDI_plus_OG_counts,
                                     colData = parsed_df,
                                     design = as.formula("~ 1"))

TDI_plus_OG_filt <- filterDatasets(TDI_plus_OG,
                                  abs_filt = TRUE,
                                  abs_filt_samples = config$filtering$abs_filt_samples)

TDI_plus_OG_filt <- DESeq2::estimateSizeFactors(TDI_plus_OG_filt)
TDI_plus_OG_filt <- DESeq2::DESeq(TDI_plus_OG_filt)
TDI_plus_OG_filt_vsd <- DESeq2::vst(TDI_plus_OG_filt, blind = TRUE)

# Batch correction
transf_batch_NObatch_experiment <- TDI_plus_OG_filt_vsd
transf_batch_NObatch_experiment_count <- limma::removeBatchEffect(
  SummarizedExperiment::assay(transf_batch_NObatch_experiment),
  transf_batch_NObatch_experiment$is_outer_control
)
SummarizedExperiment::assay(transf_batch_NObatch_experiment) <- transf_batch_NObatch_experiment_count

pca_deg_NObatch_experiment <- generatePCA(
  transf_object = transf_batch_NObatch_experiment[,c(1,3,5,7,9,10)], 
  cond_interest_varPart = c("genotype","treatment"), 
  color_variable = "genotype", 
  shape_variable = "treatment",
  ntop_genes = config$visualization$ntop_genes_pca
) + ggtitle("Integration with Original Dataset (Batch Corrected)")

ggsave(
  filename = file.path(PATHS$results_dir, "PCA_integration_with_originals.pdf"),
  plot = pca_deg_NObatch_experiment,
  width = config$visualization$plot_width, 
  height = config$visualization$plot_height, 
  units = config$visualization$plot_units
)

pca_deg_NObatch_experiment
```

## Summary and Session Information

```{r analysis_summary}
analysis_end_time <- Sys.time()
analysis_duration <- difftime(analysis_end_time, analysis_start_time, units = "mins")

# Create summary report
summary_report <- data.frame(
  Analysis = c("TDI vs DMSO (Combined)", "WWTR1-only", "TDI vs DMSO (Full)", "doxTDI vs dox"),
  Significant_Genes = c(
    nrow(deg_results_wwtr_mcherry$results_signif),
    nrow(deg_results_WWTR1$results_signif), 
    nrow(deg_results_TDI_vs_DMSO$results_signif),
    nrow(deg_results_doxTDI_vs_dox$results_signif)
  ),
  Files_Generated = c(
    "treatment_TDI_vs_DMSO.xlsx",
    "deg_results_wwtr.tsv", 
    "treatment_TDI_vs_DMSO_full.xlsx",
    "treatment_doxTDI_vs_dox.xlsx"
  )
)

knitr::kable(summary_report, caption = "Integration Analysis Summary")

message("=== INTEGRATION ANALYSIS COMPLETED ===")
message("Total analysis time: ", round(analysis_duration, 2), " minutes")
message("Results saved to: ", PATHS$results_dir)
```

```{r session_info}
session_info <- sessionInfo()
writeLines(
  capture.output(session_info),
  file.path(PATHS$results_dir, "integration_session_info.txt")
)

session_info
```

---

**Analysis completed at:** `r Sys.time()`  
**Total runtime:** `r round(difftime(Sys.time(), analysis_start_time, units = "mins"), 2)` minutes  
**Output directory:** `r PATHS$results_dir`