---
title: "ATAC-seq addtitional analysis - YAP/TAZ inhibition using K-975"
author: "Aleksandr Bykov (Enhanced by Claude Code)"
date: '`r Sys.Date()`'
output:
  html_document: 
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 8,
  dpi = 300
)

```


```{r setup-libraries}
# Set up the environment ####
deg_dir <- "~/neuroblastoma/results/ATAC-seq/ATAC-seq_drug_treatment/"
if(!dir.exists(deg_dir)){dir.create(deg_dir)}
## Loading libraries ####
library(dplyr)
library(tidyr)
library(DESeq2)
library(limma)
library(ggplot2)
library(BSgenome.Hsapiens.UCSC.hg38)
library(IRanges)
library(patchwork)
import::from(stringr, str_extract, str_detect)
import::from(GenomicRanges,GRanges)
import::from(rtracklayer, liftOver, import.chain)
import::from(openxlsx, addWorksheet, writeData, saveWorkbook)
import::from(openxlsx2, read_xlsx)
import::from(.from = TFBSTools, getMatrixSet)
import::from(chromVAR, addGCBias)
import::from(.from = RColorBrewer, brewer.pal)
import::from(.from = "~/neuroblastoma/resources/utilityScripts.R",
             "generatePCA", 
             "extract_results_nextflow_output",
             "meanExprsPerGroup_nextflow",
             "extract_results_DDS_HIC",
             "plotVolcano",
             "get_the_poissoon_p_val")
import::from(
  .from = here::here("~/neuroblastoma/resources/UtilityScriptsRNA-seq.R"),
  "filterDatasets",
  .character_only = TRUE
)
```

```{r setup-jaspar}
# BugFix - JASPAR2022 cannot be pulled from the web automatically, so it need to be loaded manually
download.file(url = "https://jaspar2022.genereg.net/download/database/JASPAR2022.sqlite",
              destfile = "~/neuroblastoma/resources/JASPAR2022.sqlite")
JASPAR2022 <-  "~/neuroblastoma/resources/JASPAR2022.sqlite"

```

```{r load-atac-data}
# Analysis part ###########################################################################
# Load and prepare ATAC-seq data for drug treatment analysis
# We use only filtered consensus peaks. And combined annotation from genecode, homer and reg.
tmp_dir <- "~/neuroblastoma/data/ATACseq/"
path_to_the_ATAC_counts_data <- file.path(tmp_dir, "consensus_peaks.mLb.clN.rds")
path_annnotation_ATAC_data <- file.path(tmp_dir, "consensus_peaks.mLb.clN.annotatePeaks.txt")
ATAC_counts_data <- readRDS(path_to_the_ATAC_counts_data)
ATAC_annotation_data <- read.csv(path_annnotation_ATAC_data, sep = "\t")

colnames(ATAC_annotation_data)[1] <- "PeakId"
ATAC_annotation_data <- ATAC_annotation_data %>%
  mutate(PeakId = factor(PeakId, levels = row.names(ATAC_counts_data))) %>%
  arrange(PeakId)

#remove Scaffolds chrs
ATAC_annotation_data <- ATAC_annotation_data[ATAC_annotation_data$Chr %in% c(1:22, "X", "Y"), ]
ATAC_counts_data <- ATAC_counts_data[rownames(ATAC_counts_data) %in% ATAC_annotation_data$PeakId, ]

#add ranges
rowRanges(ATAC_counts_data) <- GenomicRanges::GRanges(seqnames = paste0("chr", ATAC_annotation_data$Chr), 
                                              ranges = IRanges::IRanges(start = ATAC_annotation_data$Start, 
                                                                        end = ATAC_annotation_data$End
                                              )
)
mcols(ATAC_counts_data) <- DataFrame(mcols(ATAC_counts_data), ATAC_annotation_data)
```


```{r subset-drug-treatment-samples}
# Subset to drug treatment samples (CM and SH cell lines)
# Extract metadata for control, 24h, and 48h time points
samples_to_subset <- c(
  "CM_control_ATAC_S165169_REP1",
  "CM_24h_ATAC_S165170_REP1",
  "CM_48h_ATAC_S165167_REP1",
  "SH_control_ATAC_S165166_REP1",
  "SH_24h_ATAC_S165164_REP1",
  "SH_48h_ATAC_S165165_REP1",
  "SH_DMSO_ctrl_rep1_REP1",
  "SH_DMSO_ctrl_rep2_REP1",
  "SH_24h_rep1_REP1",
  "SH_24h_rep2_REP1",
  "SH_48h_rep1_REP1",
  "SH_48h_rep2_REP1",
  "CM_DMSO_ctrl_rep1_REP1",
  "CM_DMSO_ctrl_rep2_REP1",
  "CM_24h_rep1_REP1",
  "CM_24h_rep2_REP1",
  "CM_48h_rep1_REP1",
  "CM_48h_rep2_REP1"
)


drg_trtm_ATAC <- ATAC_counts_data[, colData(ATAC_counts_data)$sample %in% samples_to_subset]

colData(drg_trtm_ATAC)$cell_line <- as.factor(case_when(
  str_detect(drg_trtm_ATAC$sample, "^CM") ~ "CLBMa",
  str_detect(drg_trtm_ATAC$sample, "^SH") ~ "SK_N_SH",
  .default = colData(drg_trtm_ATAC)$sample
))
colData(drg_trtm_ATAC)$time <- as.factor(case_when(
  str_detect(drg_trtm_ATAC$sample, "24h") ~ "24h",
  str_detect(drg_trtm_ATAC$sample, "48h") ~ "48h",
  str_detect(drg_trtm_ATAC$sample, "DMSO") ~ "control",
  str_detect(drg_trtm_ATAC$sample, "control") ~ "control",
  .default = colData(drg_trtm_ATAC)$cell_line
))
colData(drg_trtm_ATAC)$time <- relevel(colData(drg_trtm_ATAC)$time, ref = "control")

# Batch: legacy samples (single replicate, named *_ATAC_S165xxx_REP1) vs the new
# batch (2 replicates, named *_rep1/rep2_REP1). Every cell_line x time combination
# is present in BOTH batches, so batch is balanced and can be corrected for.
colData(drg_trtm_ATAC)$batch <- as.factor(dplyr::if_else(
  str_detect(drg_trtm_ATAC$sample, "_ATAC_S\\d+"), "legacy", "new"
))

# alternative 
colData(drg_trtm_ATAC)$batch <- as.factor(case_when(drg_trtm_ATAC$sample %in% 
                                                      c("CM_48h_rep1_REP1",
                                                        "CM_24h_rep1_REP1",
                                                        "SH_DMSO_ctrl_rep1_REP1",
                                                        "CM_DMSO_ctrl_rep1_REP1",
                                                        "SH_24h_rep1_REP1",
                                                        "SH_48h_rep1_REP1") ~ "new_1",
                                                    
                                                    drg_trtm_ATAC$sample %in% 
                                                      c("CM_24h_rep2_REP1",
                                                        "CM_48h_rep2_REP1",
                                                        "SH_DMSO_ctrl_rep2_REP1",
                                                        "CM_DMSO_ctrl_rep2_REP1",
                                                        "SH_24h_rep2_REP1",
                                                        "SH_48h_rep2_REP1") ~ "new_2",
                                          
                                                    drg_trtm_ATAC$sample %in% 
                                                      c("SH_48h_ATAC_S165165_REP1",
                                                        "CM_24h_ATAC_S165170_REP1",
                                                        "CM_48h_ATAC_S165167_REP1",
                                                        "CM_control_ATAC_S165169_REP1",
                                                        "SH_24h_ATAC_S165164_REP1",
                                                        "SH_control_ATAC_S165166_REP1") ~ "legacy",
                                                    ))



# Variable of interest (time) stays last; batch and cell_line are adjusted for.
design(drg_trtm_ATAC) <- as.formula("~ batch + cell_line + time")

#design(drg_trtm_ATAC) <- as.formula("~ time")

#drg_trtm_ATAC@rowRanges@elementMetadata@listData[["PeakId"]]

#
RNA_SEQ_data <- openxlsx2::read_xlsx("~/neuroblastoma/results/RNA-seq/RNA-seq_MES_vs_ADR/tables/MES_vs_ADR_comprehensive_results.xlsx", sheet = 2)
RNA_SEQ_data <- RNA_SEQ_data %>%
  dplyr::mutate(Term = if_else(log2FoldChange < 0, "ADRN", "MES")) %>%
  dplyr::select(gene_symbol, Term)
mcols(drg_trtm_ATAC)$gene_category_our_RNAseq <- "ND"
mcols(drg_trtm_ATAC)$gene_category_our_RNAseq <- case_when(
  mcols(drg_trtm_ATAC)$Gene.Name %in% RNA_SEQ_data[RNA_SEQ_data$Term == "ADRN", 1] ~ "g_ADRN",
  mcols(drg_trtm_ATAC)$Gene.Name %in% RNA_SEQ_data[RNA_SEQ_data$Term == "MES", 1] ~ "g_MES",
  TRUE ~ mcols(drg_trtm_ATAC)$gene_category_our_RNAseq
)

# Add metadata - marks peaks that are around 3kb from TSS
mcols(drg_trtm_ATAC)$is_promoter_3kb <- FALSE
mcols(drg_trtm_ATAC)$is_promoter_3kb[abs(mcols(drg_trtm_ATAC)$Distance.to.TSS) <= 3000] <- TRUE
```

```{r differential-analysis}
# Perform differential accessibility analysis
# Filter low-count peaks and run DESeq2 normalization
drg_trtm_ATAC <- filterDatasets(drg_trtm_ATAC,
                                abs_filt = TRUE,
                                abs_filt_samples = 2
)

design(drg_trtm_ATAC) <- as.formula("~ batch + cell_line + time")

#design(drg_trtm_ATAC) <- as.formula("~ batch +time")
drg_trtm_ATAC <- DESeq(drg_trtm_ATAC)
vsd <- vst(drg_trtm_ATAC, blind = F)

# For visualization only: the design corrects for batch statistically, but the VST
# matrix still carries the batch structure, so PCA/heatmaps would still split by
# batch. Strip it here (NEVER feed this back into DESeq2 - the model handles it).
vsd_uncorrected <- vsd
assay(vsd) <- limma::removeBatchEffect(
  assay(vsd),
  batch  = vsd$batch_detailed,
  design = model.matrix(~ cell_line + time, data = as.data.frame(colData(vsd)))
)

pca_dar_uncorrected_time <- generatePCA(transf_object = vsd_uncorrected,
                       cond_interest_varPart = c("time", "cell_line"),
                       color_variable = "time",
                       shape_variable = "cell_line",
                       ntop_genes = 1000) +
  ggtitle("Original dataset (batch effect present)")



```


```{r plot-pca}
# Generate PCA plot to visualize sample relationships
# Left: batch effect visible; Right: after limma::removeBatchEffect
pca_dar_uncorrected_time
```

```{r extract-dar-results}
# Extract differential accessibility results for 48h vs control comparison
resultsNames(drg_trtm_ATAC)
coeff_name <- "time_48h_vs_control"
cond_numerator <- "48h"
cond_denominator <- "control"
cond_variable <- "time"
padj_cutoff = 0.05
log2FC_cutoff = 1
sample_result <- results(drg_trtm_ATAC, contrast = c(cond_variable, cond_numerator, cond_denominator))
ATAC_dds_results_48h <- extract_results_nextflow_output(dds_object = drg_trtm_ATAC,
                                            dds_result = sample_result,
                                            coeff_name = coeff_name,
                                            cond_numerator = cond_numerator,
                                            cond_denominator = cond_denominator,
                                            cond_variable = cond_variable,
                                            padj_cutoff = padj_cutoff,
                                            log2FC_cutoff = log2FC_cutoff)
# Write the result to the xlsx file
XLSX_OUT <- openxlsx::createWorkbook()
openxlsx::addWorksheet(XLSX_OUT, "results_signif")
openxlsx::addWorksheet(XLSX_OUT, "de_details")
openxlsx::addWorksheet(XLSX_OUT, "results_all")
openxlsx::writeData(XLSX_OUT, x = ATAC_dds_results_48h$results_signif, sheet = "results_signif")
openxlsx::writeData(XLSX_OUT, x = ATAC_dds_results_48h$de_details, sheet = "de_details")
openxlsx::writeData(XLSX_OUT, x = ATAC_dds_results_48h$results_all, sheet = "results_all")
openxlsx::saveWorkbook(XLSX_OUT, file.path(deg_dir,  paste0("DAR_results_time_48h_vs_control.xlsx")), overwrite = T)


resultsNames(drg_trtm_ATAC)
coeff_name <- "time_24h_vs_control"
cond_numerator <- "24h"
cond_denominator <- "control"
cond_variable <- "time"
padj_cutoff = 0.05
log2FC_cutoff = 1
sample_result <- results(drg_trtm_ATAC, contrast = c(cond_variable, cond_numerator, cond_denominator))
ATAC_dds_results_24h <- extract_results_nextflow_output(dds_object = drg_trtm_ATAC,
                                            dds_result = sample_result,
                                            coeff_name = coeff_name,
                                            cond_numerator = cond_numerator,
                                            cond_denominator = cond_denominator,
                                            cond_variable = cond_variable,
                                            padj_cutoff = padj_cutoff,
                                            log2FC_cutoff = log2FC_cutoff)

XLSX_OUT <- openxlsx::createWorkbook()
openxlsx::addWorksheet(XLSX_OUT, "results_signif")
openxlsx::addWorksheet(XLSX_OUT, "de_details")
openxlsx::addWorksheet(XLSX_OUT, "results_all")
openxlsx::writeData(XLSX_OUT, x = ATAC_dds_results_24h$results_signif, sheet = "results_signif")
openxlsx::writeData(XLSX_OUT, x = ATAC_dds_results_24h$de_details, sheet = "de_details")
openxlsx::writeData(XLSX_OUT, x = ATAC_dds_results_24h$results_all, sheet = "results_all")
openxlsx::saveWorkbook(XLSX_OUT, file.path(deg_dir,  paste0("DAR_results_time_24h_vs_control.xlsx")), overwrite = T)
```


```{r export-bed-files}
# Export differentially accessible peaks (24h and 48h vs control) and the
# consensus peak set as BED files for downstream tools (IGV, bedtools, GREAT...).

# Helper: turn a DAR results table into a GRanges and write it as a BED file.
# The HOMER annotation coordinates (Chr/Start/End) are 1-based; rtracklayer::export
# writes proper 0-based BED starts automatically. log2FoldChange is stored as the
# BED score so up/down peaks stay distinguishable in genome browsers.
export_dars_to_bed <- function(dar_results, bed_path) {
  dar_results <- dar_results[!is.na(dar_results$Chr) & !is.na(dar_results$Start) & !is.na(dar_results$End), ]
  dar_gr <- GenomicRanges::GRanges(
    seqnames = ifelse(grepl("^chr", dar_results$Chr), dar_results$Chr, paste0("chr", dar_results$Chr)),
    ranges   = IRanges::IRanges(start = dar_results$Start, end = dar_results$End),
    strand   = ifelse(dar_results$Strand %in% c("+", "-"), dar_results$Strand, "*")
  )
  names(dar_gr)     <- dar_results$PeakId
  dar_gr$name       <- dar_results$PeakId
  dar_gr$score      <- dar_results$log2FoldChange
  rtracklayer::export.bed(dar_gr, con = bed_path)
  invisible(dar_gr)
}

# Significant DARs (padj < 0.05 & |log2FC| > 1, as filtered above)
export_dars_to_bed(ATAC_dds_results_48h$results_signif,
                   file.path(deg_dir, "DAR_peaks_time_48h_vs_control.bed"))
export_dars_to_bed(ATAC_dds_results_24h$results_signif,
                   file.path(deg_dir, "DAR_peaks_time_24h_vs_control.bed"))

# Consensus peaks used in the differential analysis (post count-filtering)
consensus_gr <- SummarizedExperiment::rowRanges(drg_trtm_ATAC)
names(consensus_gr)   <- mcols(drg_trtm_ATAC)$PeakId
consensus_gr$name     <- mcols(drg_trtm_ATAC)$PeakId
mcols(consensus_gr)   <- mcols(consensus_gr)[, "name", drop = FALSE]
rtracklayer::export.bed(consensus_gr,
                        con = file.path(deg_dir, "consensus_peaks.bed"))
```



```{r generate-heatmaps}
# Generate heatmaps of differentially accessible regions
# Focus on peaks associated with MES and ADRN gene signatures
## Now the same analysis, but for the promoter regions - ± 3000bp around TSSs ####
ATAC_dds <- drg_trtm_ATAC
#ATAC_dds <- ATAC_dds[mcols(ATAC_dds)$is_promoter_3kb,]

# Uncoment only for analysis of promoters
# ATAC_dds <- DESeq(ATAC_dds)
# 
# vsd <- vst(ATAC_dds, blind = F)
# 
# # Visualization only: remove batch so heatmap columns cluster by treatment, not batch.
# assay(vsd) <- limma::removeBatchEffect(
#   assay(vsd),
#   batch  = vsd$batch,
#   design = model.matrix(~ cell_line + time, data = as.data.frame(colData(vsd)))
# )
# 
# pca_dar <- generatePCA(transf_object = vsd,
#                        cond_interest_varPart = c("time", "cell_line"), 
#                        color_variable = "time", 
#                        shape_variable = "cell_line",
#                        ntop_genes = 1000) +
#   ggtitle("3kb Promoters")
# plot(pca_dar)

# # extracting results
# resultsNames(ATAC_dds)
# coeff_name <- "time_48h_vs_control"
# cond_numerator <-  "48h"
# cond_denominator <-  "control"
# cond_variable <- "time"
# padj_cutoff = 0.05
# log2FC_cutoff = 1
# 
# ATAC_dds_results <- extract_results_nextflow_output(dds_object = ATAC_dds,
#                                         coeff_name = coeff_name,
#                                         cond_numerator = cond_numerator,
#                                         cond_denominator = cond_denominator,
#                                         cond_variable = cond_variable,
#                                         padj_cutoff = padj_cutoff,
#                                         log2FC_cutoff = log2FC_cutoff)

ATAC_dds_results <- ATAC_dds_results_48h

# Making Heatmap
metadata_heatmap <- as.data.frame(colData(ATAC_dds))
heatmap_counts <- SummarizedExperiment::assay(vsd)
rownames(heatmap_counts) <- vsd@rowRanges$PeakId

all_ADR_peaks <- c(mcols(drg_trtm_ATAC)$PeakId[mcols(drg_trtm_ATAC)$gene_category_our_RNAseq == "g_ADRN"])
all_MES_peaks <- c(mcols(drg_trtm_ATAC)$PeakId[mcols(drg_trtm_ATAC)$gene_category_our_RNAseq == "g_MES"])

# removed experiment batch effects! ? use experimen+cell_line removed???
ATAC_signif <- ATAC_dds_results$results_signif
heatmap_counts <- heatmap_counts[rownames(heatmap_counts) %in% ATAC_dds_results$results_signif$PeakId, ]
heatmap_counts_MES_spec_genes <- heatmap_counts[rownames(heatmap_counts) %in% all_MES_peaks, ]
heatmap_counts_ADRN_spec_genes <- heatmap_counts[rownames(heatmap_counts) %in% all_ADR_peaks, ]

annotation_col <- metadata_heatmap %>%
  dplyr::select(cell_line, time) %>% 
  dplyr::arrange(time, cell_line)
heatmap_counts <- heatmap_counts[, match(rownames(annotation_col), colnames(heatmap_counts))]

heatmap_counts_MES_spec_genes <- heatmap_counts_MES_spec_genes[, match(rownames(annotation_col), colnames(heatmap_counts_MES_spec_genes))]
# Map PeakId -> Gene.Name by matching on PeakId so labels follow the matrix row
# order. Do NOT use `ATAC_signif[PeakId %in% rownames, "Gene.Name"]`: that returns
# names in ATAC_signif's padj order, which scrambles labels vs the peak-ordered rows.
rownames(heatmap_counts_MES_spec_genes) <- ATAC_signif$Gene.Name[match(rownames(heatmap_counts_MES_spec_genes), ATAC_signif$PeakId)]

heatmap_counts_ADRN_spec_genes <- heatmap_counts_ADRN_spec_genes[, match(rownames(annotation_col), colnames(heatmap_counts_ADRN_spec_genes))]
rownames(heatmap_counts_ADRN_spec_genes) <- ATAC_signif$Gene.Name[match(rownames(heatmap_counts_ADRN_spec_genes), ATAC_signif$PeakId)]

#heatmap_row_annot_MES_spec_genes <- data.frame(gene_name = ATAC_signif[ATAC_signif$PeakId %in% row.names(heatmap_counts_MES_spec_genes), "Gene.Name"],
#                                               row.names =  row.names(heatmap_counts_MES_spec_genes))

ensembl2symbol_annot <- ATAC_dds_results$results_all %>%
  dplyr::select(PeakId, Gene.Name)

color.scheme <- colorRampPalette(c("navy", "white", "firebrick3"))(50)
ann_colors = list(
  time = c( control = "#005f73", `24h` = "#0a9396", `48h` = "#94d2bd"),
  cell_line = c(CLBMa = "#E9D8A6", SK_N_SH = "#D9BE6D")
)

heatmap <- pheatmap::pheatmap(heatmap_counts,
                              main = "Heatmap of signif. All DARs 48h vs control.",
                              scale = "row",
                              annotation_col = annotation_col,
                              annotation_colors = ann_colors,
                              show_colnames = FALSE,
                              show_rownames = TRUE,
                              cluster_cols = FALSE,
                              color = color.scheme,
                              fontsize = 10, fontsize_row = 10) 
ggsave(filename = file.path(deg_dir, "heatmap_48h_vs_control_all_DARs_ATAC.png"), 
       plot = heatmap,
       width = 18, height = 20, units = "cm")
ggsave(filename = file.path(deg_dir, "heatmap_48h_vs_control_all_DARs_ATAC.pdf"), 
       plot = heatmap,
       width = 18, height = 20, units = "cm")

heatmap <- pheatmap::pheatmap(heatmap_counts_MES_spec_genes,
                              main = "Heatmap of signif. DAR ADRN vs MES all DARs \n MES-specific genes only (defined by RNA-seq)",
                              scale = "row",
                              annotation_col = annotation_col,
                              annotation_colors = ann_colors,
                              #annotation_row = heatmap_row_annot_MES_spec_genes,
                              show_colnames = FALSE,
                              show_rownames = TRUE,
                              cluster_cols = FALSE,
                              color = color.scheme,
                              fontsize = 10, fontsize_row = 10) #height=10, cellwidth = 11, cellheight = 11
ggsave(filename = file.path(deg_dir, "heatmap_48h_vs_control_all_DARs_MES-RNAseq-specific.png"), 
       plot=heatmap,
       width = 18, height = 20, units = "cm")
ggsave(filename = file.path(deg_dir, "heatmap_48h_vs_control_all_DARs_MES-RNAseq-specific.pdf"), 
       plot=heatmap,
       width = 18, height = 20, units = "cm")

heatmap <- pheatmap::pheatmap(heatmap_counts_ADRN_spec_genes,
                              main = "Heatmap of signif. DAR ADRN vs MES all DARs \n ADRN-specific genes only (defined by RNA-seq)",
                              scale = "row",
                              annotation_col = annotation_col,
                              annotation_colors = ann_colors,
                              #annotation_row = row_annot_symbols,
                              show_colnames = FALSE,
                              show_rownames = TRUE,
                              cluster_cols = FALSE,
                              color = color.scheme,
                              fontsize = 10, fontsize_row = 10) #height=10, cellwidth = 11, cellheight = 11
ggsave(filename = file.path(deg_dir, "heatmap_48h_vs_control_all_DARs_ADRN-RNAseq-specific.png"), 
       plot=heatmap,
       width = 18, height = 20, units = "cm")
ggsave(filename = file.path(deg_dir, "heatmap_48h_vs_control_all_DARs_ADRN-RNAseq-specific.pdf"),
       plot=heatmap,
       width = 18, height = 20, units = "cm")
```


```{r generate-heatmaps-merged-replicates}
# Same DAR heatmaps as above, but with replicates MERGED: every replicate of a
# given time x cell_line group (legacy single-rep + new rep1/rep2) is averaged
# into a single column. Uses the same batch-corrected VST matrices, so only the
# column dimension changes.

# Collapse the columns of a count matrix to one averaged column per
# time x cell_line group. Grouping is read from metadata_heatmap (rows = samples),
# so it stays correct regardless of sample naming.
merge_replicates <- function(mat, metadata, time_order = c("control", "24h", "48h")) {
  meta          <- metadata[colnames(mat), c("cell_line", "time"), drop = FALSE]
  sample_group  <- paste(as.character(meta$time), as.character(meta$cell_line), sep = "_")
  cell_order    <- levels(droplevels(as.factor(meta$cell_line)))

  # Desired column order: time (control -> 24h -> 48h), then cell_line
  wanted        <- expand.grid(time = time_order, cell_line = cell_order, stringsAsFactors = FALSE)
  wanted$group  <- paste(wanted$time, wanted$cell_line, sep = "_")
  wanted        <- wanted[wanted$group %in% sample_group, , drop = FALSE]

  merged <- sapply(wanted$group, function(g) {
    rowMeans(mat[, colnames(mat)[sample_group == g], drop = FALSE])
  })
  merged <- as.matrix(merged)
  rownames(merged) <- rownames(mat)
  colnames(merged) <- wanted$group

  attr(merged, "annotation") <- data.frame(
    time      = factor(wanted$time, levels = time_order),
    cell_line = factor(wanted$cell_line, levels = cell_order),
    row.names = wanted$group
  )
  merged
}

heatmap_counts_merged           <- merge_replicates(heatmap_counts, metadata_heatmap)
heatmap_counts_MES_merged       <- merge_replicates(heatmap_counts_MES_spec_genes, metadata_heatmap)
heatmap_counts_ADRN_merged      <- merge_replicates(heatmap_counts_ADRN_spec_genes, metadata_heatmap)
annotation_col_merged           <- attr(heatmap_counts_merged, "annotation")

heatmap <- pheatmap::pheatmap(heatmap_counts_merged,
                              main = "Heatmap of signif. All DARs 48h vs control (merged replicates).",
                              scale = "row",
                              annotation_col = annotation_col_merged,
                              annotation_colors = ann_colors,
                              show_colnames = TRUE,
                              show_rownames = TRUE,
                              cluster_cols = FALSE,
                              color = color.scheme,
                              fontsize = 10, fontsize_row = 10)
ggsave(filename = file.path(deg_dir, "heatmap_48h_vs_control_all_DARs_ATAC_merged_replicates.png"),
       plot = heatmap,
       width = 18, height = 20, units = "cm")
ggsave(filename = file.path(deg_dir, "heatmap_48h_vs_control_all_DARs_ATAC_merged_replicates.pdf"),
       plot = heatmap,
       width = 18, height = 20, units = "cm")

heatmap <- pheatmap::pheatmap(heatmap_counts_MES_merged,
                              main = "Heatmap of signif. DAR ADRN vs MES all DARs \n MES-specific genes only (merged replicates)",
                              scale = "row",
                              annotation_col = annotation_col_merged,
                              annotation_colors = ann_colors,
                              show_colnames = TRUE,
                              show_rownames = TRUE,
                              cluster_cols = FALSE,
                              color = color.scheme,
                              fontsize = 10, fontsize_row = 10)
ggsave(filename = file.path(deg_dir, "heatmap_48h_vs_control_all_DARs_MES-RNAseq-specific_merged_replicates.png"),
       plot = heatmap,
       width = 18, height = 20, units = "cm")
ggsave(filename = file.path(deg_dir, "heatmap_48h_vs_control_all_DARs_MES-RNAseq-specific_merged_replicates.pdf"),
       plot = heatmap,
       width = 18, height = 20, units = "cm")

heatmap <- pheatmap::pheatmap(heatmap_counts_ADRN_merged,
                              main = "Heatmap of signif. DAR ADRN vs MES all DARs \n ADRN-specific genes only (merged replicates)",
                              scale = "row",
                              annotation_col = annotation_col_merged,
                              annotation_colors = ann_colors,
                              show_colnames = TRUE,
                              show_rownames = TRUE,
                              cluster_cols = FALSE,
                              color = color.scheme,
                              fontsize = 10, fontsize_row = 10)
ggsave(filename = file.path(deg_dir, "heatmap_48h_vs_control_all_DARs_ADRN-RNAseq-specific_merged_replicates.png"),
       plot = heatmap,
       width = 18, height = 20, units = "cm")
ggsave(filename = file.path(deg_dir, "heatmap_48h_vs_control_all_DARs_ADRN-RNAseq-specific_merged_replicates.pdf"),
       plot = heatmap,
       width = 18, height = 20, units = "cm")
```

```{r generate-heatmaps-MES-by-cellline}
# MES-specific-gene heatmaps (the line-406 heatmap) split by cell line, in two
# flavours per cell line:
#   (1) individual replicate columns (CM only / SH only)
#   (2) replicates merged into per-timepoint group-mean columns (CM only / SH only)
# Reuses heatmap_counts_MES_spec_genes, metadata_heatmap, merge_replicates(),
# ann_colors and color.scheme from the chunks above. Existing heatmaps untouched.

time_levels <- c("control", "24h", "48h")

# Drop rows that would break scale = "row" (zero/undefined variance across the
# columns being plotted) so a single-cell-line or 3-column matrix still renders.
drop_unscalable_rows <- function(mat) {
  keep <- apply(mat, 1, function(x) {
    s <- stats::sd(x, na.rm = TRUE)
    is.finite(s) && s > 0 && !any(is.na(x))
  })
  mat[keep, , drop = FALSE]
}

for (cl in c("CLBMa", "SK_N_SH")) {

  # Samples for this cell line, ordered control -> 24h -> 48h.
  cl_meta <- metadata_heatmap[metadata_heatmap$cell_line == cl, c("cell_line", "time"), drop = FALSE]
  cl_meta <- cl_meta[order(factor(as.character(cl_meta$time), levels = time_levels)), , drop = FALSE]
  cl_samples <- rownames(cl_meta)

  ## (1) Individual replicate columns -----------------------------------------
  mat_ind <- heatmap_counts_MES_spec_genes[, cl_samples, drop = FALSE]
  mat_ind <- drop_unscalable_rows(mat_ind)
  annotation_col_ind <- cl_meta
  annotation_col_ind$time <- factor(as.character(annotation_col_ind$time), levels = time_levels)

  hm_ind <- pheatmap::pheatmap(
    mat_ind,
    main = sprintf("MES-specific DARs (RNA-seq) - %s\n48h vs control, individual replicates", cl),
    scale = "row",
    annotation_col = annotation_col_ind,
    annotation_colors = ann_colors,
    show_colnames = FALSE,
    show_rownames = TRUE,
    cluster_cols = FALSE,
    color = color.scheme,
    fontsize = 10, fontsize_row = 10)
  ggsave(filename = file.path(deg_dir, sprintf("heatmap_48h_vs_control_MES-RNAseq-specific_%s.png", cl)),
         plot = hm_ind, width = 14, height = 20, units = "cm")
  ggsave(filename = file.path(deg_dir, sprintf("heatmap_48h_vs_control_MES-RNAseq-specific_%s.pdf", cl)),
         plot = hm_ind, width = 14, height = 20, units = "cm")

  ## (2) Replicates merged (per-timepoint group means) ------------------------
  mat_merged <- merge_replicates(heatmap_counts_MES_spec_genes[, cl_samples, drop = FALSE],
                                 metadata_heatmap, time_order = time_levels)
  annotation_col_merged_cl <- attr(mat_merged, "annotation")
  mat_merged <- drop_unscalable_rows(mat_merged)

  hm_merged <- pheatmap::pheatmap(
    mat_merged,
    main = sprintf("MES-specific DARs (RNA-seq) - %s\n48h vs control, merged replicates", cl),
    scale = "row",
    annotation_col = annotation_col_merged_cl,
    annotation_colors = ann_colors,
    show_colnames = TRUE,
    show_rownames = TRUE,
    cluster_cols = FALSE,
    color = color.scheme,
    fontsize = 10, fontsize_row = 10)
  ggsave(filename = file.path(deg_dir, sprintf("heatmap_48h_vs_control_MES-RNAseq-specific_%s_merged_replicates.png", cl)),
         plot = hm_merged, width = 12, height = 20, units = "cm")
  ggsave(filename = file.path(deg_dir, sprintf("heatmap_48h_vs_control_MES-RNAseq-specific_%s_merged_replicates.pdf", cl)),
         plot = hm_merged, width = 12, height = 20, units = "cm")
}
```

```{r notch-gene-accessibility}
# Accessibility of the three Notch-pathway genes MAML2, NOTCH1 and NOTCH2,
# shown separately for each gene and for both cell lines (CLBMa vs SK_N_SH).
#
# Peaks are selected by GENOMIC OVERLAP with each gene locus (gene body + a 5 kb
# promoter flank), NOT via the HOMER `Gene.Name` column. HOMER assigns every peak
# to its single nearest TSS, so NOTCH1's peaks are labelled with small features
# that sit inside the locus (C9orf163, MIR4673, NALT1 ...) and a Gene.Name filter
# returns ZERO peaks for NOTCH1. Overlap selection recovers them.
#
# Accessibility = the batch-corrected VST matrix `vsd` built in the
# differential-analysis chunk (same values used for the DAR heatmaps), so the
# strong ATAC batch effect is already removed for visualisation.
notch_dir <- file.path(deg_dir, "notch_accessibility")
if (!dir.exists(notch_dir)) dir.create(notch_dir, recursive = TRUE)

# hg38 gene-body coordinates (Ensembl/GENCODE), padded by a promoter flank.
notch_loci <- GenomicRanges::GRanges(
  seqnames = c("chr11", "chr9", "chr1"),
  ranges   = IRanges::IRanges(
    start = c(95976614, 136494433, 119911553),
    end   = c(96343547, 136545786, 120100779)),
  gene = c("MAML2", "NOTCH1", "NOTCH2")
)
promoter_flank <- 5000
notch_loci <- GenomicRanges::resize(
  notch_loci, width = GenomicRanges::width(notch_loci) + 2 * promoter_flank, fix = "center")

# Map consensus peaks (rows of vsd) to a gene locus by overlap.
peak_gr <- SummarizedExperiment::rowRanges(vsd)
hits    <- GenomicRanges::findOverlaps(peak_gr, notch_loci)
peak_map <- data.frame(row_idx = S4Vectors::queryHits(hits),
                       gene    = notch_loci$gene[S4Vectors::subjectHits(hits)])
peak_map <- peak_map[!duplicated(peak_map$row_idx), ]   # first locus wins if a peak overlaps two
message("Notch peaks selected per gene (by genomic overlap):")
print(table(peak_map$gene))

# Assemble long table: one row per peak x sample, with gene + sample metadata.
notch_acc <- SummarizedExperiment::assay(vsd)[peak_map$row_idx, , drop = FALSE]
rownames(notch_acc) <- mcols(peak_gr)$PeakId[peak_map$row_idx]
peak2gene <- data.frame(PeakId = mcols(peak_gr)$PeakId[peak_map$row_idx], gene = peak_map$gene)
notch_meta <- as.data.frame(colData(vsd))[, c("sample", "cell_line", "time", "batch")]
notch_meta$col <- colnames(vsd)

notch_long <- as.data.frame(notch_acc) %>%
  tibble::rownames_to_column("PeakId") %>%
  tidyr::pivot_longer(-PeakId, names_to = "col", values_to = "accessibility") %>%
  dplyr::left_join(peak2gene, by = "PeakId") %>%
  dplyr::left_join(notch_meta, by = "col")

# Per-sample locus accessibility = mean batch-corrected VST across the gene's peaks.
notch_summary <- notch_long %>%
  dplyr::group_by(gene, sample, cell_line, time, batch) %>%
  dplyr::summarise(mean_access = mean(accessibility), n_peaks = dplyr::n(), .groups = "drop") %>%
  dplyr::mutate(gene = factor(gene, levels = c("MAML2", "NOTCH1", "NOTCH2")))

write.csv(notch_summary, file.path(notch_dir, "notch_gene_accessibility_per_sample.csv"), row.names = FALSE)
write.csv(notch_long,    file.path(notch_dir, "notch_gene_accessibility_per_peak.csv"),   row.names = FALSE)

# ---- Statistics: pairwise comparison of timepoints -------------------------
# The figure uses the batch-corrected `vsd` for DISPLAY. For INFERENCE we instead
# model the UNCORRECTED VST (`vsd_uncorrected`, created in the differential-analysis
# chunk) and put `batch` in the model. Testing on values from which batch was
# already removed genome-wide, and THEN adding a batch term, would double-count the
# correction and make p-values anticonservative; modelling batch explicitly on the
# uncorrected values is the honest route.
#
# Per gene x cell line we fit  lm(mean_access ~ batch + time)  on the per-sample
# locus means and read off the three pairwise timepoint contrasts as general linear
# hypotheses (base R: estimate and SE from the model coefficients and vcov).
#
# NOTE: this is a locus-level summary test to accompany the plot (n = 3 replicates
# per timepoint per line, so power is limited). The authoritative, peak-resolution
# differential test is the DESeq2 DAR analysis (`ATAC_dds_results_24h/48h`), which
# models raw counts with the negative-binomial GLM.
notch_acc_raw <- SummarizedExperiment::assay(vsd_uncorrected)[peak_map$row_idx, , drop = FALSE]
rownames(notch_acc_raw) <- mcols(peak_gr)$PeakId[peak_map$row_idx]
notch_summary_raw <- as.data.frame(notch_acc_raw) %>%
  tibble::rownames_to_column("PeakId") %>%
  tidyr::pivot_longer(-PeakId, names_to = "col", values_to = "accessibility") %>%
  dplyr::left_join(peak2gene, by = "PeakId") %>%
  dplyr::left_join(notch_meta, by = "col") %>%
  dplyr::group_by(gene, sample, cell_line, time, batch) %>%
  dplyr::summarise(mean_access = mean(accessibility), .groups = "drop")

pairwise_time_tests <- function(df) {
  df$time <- droplevels(factor(df$time, levels = c("control", "24h", "48h")))
  fit <- lm(mean_access ~ batch + time, data = df)
  b <- coef(fit); V <- vcov(fit); dfr <- df.residual(fit)
  cf <- function(nm) { v <- setNames(rep(0, length(b)), names(b)); if (nm %in% names(v)) v[nm] <- 1; v }
  contrasts <- list(control_vs_24h = cf("time24h"),
                    control_vs_48h = cf("time48h"),
                    `24h_vs_48h`   = cf("time48h") - cf("time24h"))
  do.call(rbind, lapply(names(contrasts), function(nm) {
    L <- contrasts[[nm]]; est <- sum(L * b); se <- sqrt(as.numeric(t(L) %*% V %*% L))
    tval <- est / se
    data.frame(contrast = nm, estimate = est, se = se, t = tval, df = dfr,
               p_value = 2 * pt(-abs(tval), dfr))
  }))
}
notch_stats <- do.call(rbind, lapply(
  split(notch_summary_raw, list(notch_summary_raw$gene, notch_summary_raw$cell_line), drop = TRUE),
  function(df) cbind(gene = as.character(df$gene[1]),
                     cell_line = as.character(df$cell_line[1]),
                     pairwise_time_tests(df))))
notch_stats$p_adj <- p.adjust(notch_stats$p_value, method = "BH")   # BH across all 18 tests
rownames(notch_stats) <- NULL
notch_stats <- notch_stats %>%
  dplyr::mutate(gene = factor(gene, levels = c("MAML2", "NOTCH1", "NOTCH2")))
write.csv(notch_stats, file.path(notch_dir, "notch_gene_accessibility_timepoint_stats.csv"), row.names = FALSE)
message("Pairwise timepoint statistics (uncorrected VST + batch covariate):")
print(notch_stats)

# Per-panel p-value labels for the figures (raw p shown; adjusted p_adj is in the CSV).
notch_labels <- notch_stats %>%
  dplyr::mutate(txt = sprintf("%s: p=%.3g", sub("_vs_", "–", contrast), p_value)) %>%
  dplyr::group_by(gene, cell_line) %>%
  dplyr::summarise(label = paste(txt, collapse = "\n"), .groups = "drop")

# Batch colours are derived from the actual `batch` factor levels so the code works
# whether batch is 2-level (legacy / new) or 3-level (legacy / new_rep1 / new_rep2).
# Known levels get fixed colours; anything else falls back to a generated palette.
.notch_batch_base   <- c(legacy = "#D9BE6D", new = "#0a9396",
                         new_rep1 = "#0a9396", new_rep2 = "#94d2bd")
notch_batch_levels  <- levels(droplevels(as.factor(colData(vsd)$batch)))
notch_batch_colors  <- if (all(notch_batch_levels %in% names(.notch_batch_base))) {
  .notch_batch_base[notch_batch_levels]
} else {
  setNames(grDevices::hcl.colors(length(notch_batch_levels), "Teal"), notch_batch_levels)
}

# Combined figure: genes (rows) x cell lines (cols); one point per replicate,
# coloured by batch, boxplot over the replicates of each timepoint.
notch_plot_combined <- ggplot(notch_summary, aes(time, mean_access)) +
  geom_boxplot(outlier.shape = NA, colour = "grey70", fill = NA, width = 0.5) +
  geom_point(aes(colour = batch), size = 2.6, position = position_jitter(width = 0.12, height = 0)) +
  facet_grid(gene ~ cell_line, scales = "free_y") +
  scale_colour_manual(values = notch_batch_colors) +
  geom_text(data = notch_labels, aes(x = -Inf, y = Inf, label = label),
            hjust = -0.05, vjust = 1.15, size = 2.4, lineheight = 0.9, inherit.aes = FALSE) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.22))) +
  labs(title = "Chromatin accessibility of Notch-pathway genes",
       subtitle = "Batch-corrected VST, mean over consensus peaks per gene locus (±5 kb)",
       x = "K-975 treatment", y = "Mean accessibility (VST)", colour = "batch") +
  theme_bw(base_size = 12)
print(notch_plot_combined)
ggsave(file.path(notch_dir, "notch_accessibility_all_genes.png"), notch_plot_combined,
       width = 18, height = 22, units = "cm", dpi = 300)
ggsave(file.path(notch_dir, "notch_accessibility_all_genes.pdf"), notch_plot_combined,
       width = 18, height = 22, units = "cm")

# One separate figure per gene (both cell lines side by side).
for (g in c("MAML2", "NOTCH1", "NOTCH2")) {
  df_g <- notch_summary[notch_summary$gene == g, ]
  np   <- unique(df_g$n_peaks)
  p_g  <- ggplot(df_g, aes(time, mean_access)) +
    geom_boxplot(outlier.shape = NA, colour = "grey70", fill = NA, width = 0.5) +
    geom_point(aes(colour = batch), size = 3, position = position_jitter(width = 0.12, height = 0)) +
    facet_wrap(~ cell_line) +
    scale_colour_manual(values = notch_batch_colors) +
    geom_text(data = notch_labels[notch_labels$gene == g, ], aes(x = -Inf, y = Inf, label = label),
              hjust = -0.05, vjust = 1.15, size = 2.6, lineheight = 0.9, inherit.aes = FALSE) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.22))) +
    labs(title = sprintf("%s accessibility (K-975 treatment)", g),
         subtitle = sprintf("Batch-corrected VST, mean over %d consensus peaks in the locus", np),
         x = "K-975 treatment", y = "Mean accessibility (VST)", colour = "batch") +
    theme_bw(base_size = 12)
  print(p_g)
  ggsave(file.path(notch_dir, sprintf("notch_accessibility_%s.png", g)), p_g,
         width = 16, height = 10, units = "cm", dpi = 300)
  ggsave(file.path(notch_dir, sprintf("notch_accessibility_%s.pdf", g)), p_g,
         width = 16, height = 10, units = "cm")
}

# ===========================================================================
# (a) Per-peak DESeq2 statistics for each Notch locus
# ---------------------------------------------------------------------------
# The locus-mean plots above can hide a single responsive element: e.g. an MAML2
# intron that opens while flanking peaks drift the other way, so the average
# barely moves. Here we pull the peak-level DESeq2 results already computed in
# this document (`ATAC_dds_results_48h/24h`, design ~ batch + cell_line + time)
# for every peak that falls in a Notch locus, and report the most significant one.
#
# IMPORTANT: these are the COMBINED (both cell lines, cell_line-adjusted) treatment
# effects. A cell-line-specific effect is diluted here - e.g. the MAML2 intron
# Interval_88730 is padj~0.0015 in the SH-only analysis but only ~0.07 in this
# pooled test - so a peak can be a DAR in the per-line analysis yet not here.
notch_res_all_48h <- ATAC_dds_results_48h$results_all
notch_res_all_24h <- ATAC_dds_results_24h$results_all

pull_locus_peak_stats <- function(res_all, contrast_label) {
  res_all %>%
    dplyr::semi_join(peak2gene, by = "PeakId") %>%          # keep only locus peaks
    dplyr::left_join(peak2gene, by = "PeakId") %>%
    dplyr::transmute(gene, PeakId, Annotation, Distance.to.TSS,
                     log2FoldChange = as.numeric(log2FoldChange),
                     padj           = as.numeric(padj),
                     contrast       = contrast_label)
}
notch_peak_stats <- dplyr::bind_rows(
  pull_locus_peak_stats(notch_res_all_48h, "48h_vs_control"),
  pull_locus_peak_stats(notch_res_all_24h, "24h_vs_control")) %>%
  dplyr::mutate(gene = factor(gene, levels = c("MAML2", "NOTCH1", "NOTCH2")))
write.csv(notch_peak_stats, file.path(notch_dir, "notch_locus_peaks_DESeq2_stats.csv"), row.names = FALSE)

# Most significant peak per gene x contrast + count of peaks passing the DAR filter.
notch_best_peak <- notch_peak_stats %>%
  dplyr::filter(!is.na(padj)) %>%
  dplyr::group_by(gene, contrast) %>%
  dplyr::summarise(
    n_tested_peaks  = dplyr::n(),
    n_signif_peaks  = sum(padj < 0.05 & abs(log2FoldChange) > 1),
    best_PeakId     = PeakId[which.min(padj)],
    best_log2FC     = round(log2FoldChange[which.min(padj)], 3),
    best_padj       = signif(min(padj), 3),
    best_annotation = substr(Annotation[which.min(padj)], 1, 30),
    .groups = "drop")
message("Most significant peak per Notch gene (combined DESeq2, cell_line-adjusted):")
print(as.data.frame(notch_best_peak))
write.csv(notch_best_peak, file.path(notch_dir, "notch_locus_best_peak_DESeq2.csv"), row.names = FALSE)

# ===========================================================================
# (b) Per-peak heatmap of each Notch locus
# ---------------------------------------------------------------------------
# Rows = individual consensus peaks in the locus (genomic order, 5'->3'); columns
# = samples grouped by cell line then time. Batch-corrected VST, row-scaled, so a
# single opening/closing element stands out instead of being averaged away. The
# left-hand row annotations give each peak's 48h-vs-control log2FC and whether it
# is a significant DAR (padj<0.05 & |log2FC|>1) in the combined analysis.
locus_peak_heatmap <- function(g) {
  ids <- as.character(peak2gene$PeakId[peak2gene$gene == g])
  gr  <- peak_gr[as.character(mcols(peak_gr)$PeakId) %in% ids]
  ids_ordered <- as.character(mcols(gr)$PeakId[order(GenomicRanges::start(gr))])  # 5'->3'
  mat <- SummarizedExperiment::assay(vsd)[match(ids_ordered, as.character(mcols(peak_gr)$PeakId)), , drop = FALSE]
  rownames(mat) <- ids_ordered
  keep <- apply(mat, 1, function(x) stats::sd(x) > 0 & !any(is.na(x)))  # row-scaling needs variance
  mat <- mat[keep, , drop = FALSE]; ids_ordered <- ids_ordered[keep]

  col_ann <- as.data.frame(colData(vsd))[, c("cell_line", "time", "batch")]
  col_ann <- col_ann[order(col_ann$cell_line, col_ann$time), , drop = FALSE]
  # Drop any unused factor levels: pheatmap errors if an annotation level has no
  # matching colour, so the annotation levels must match annotation_colors exactly.
  col_ann$cell_line <- droplevels(as.factor(col_ann$cell_line))
  col_ann$time      <- droplevels(as.factor(col_ann$time))
  col_ann$batch     <- droplevels(as.factor(col_ann$batch))
  mat     <- mat[, rownames(col_ann), drop = FALSE]
  grp      <- paste(col_ann$cell_line, col_ann$time)
  gaps_col <- utils::head(cumsum(rle(grp)$lengths), -1)   # visual gaps between the 6 groups

  lfc <- notch_res_all_48h[match(ids_ordered, notch_res_all_48h$PeakId), ]
  row_ann <- data.frame(
    log2FC_48h = as.numeric(lfc$log2FoldChange),
    DAR_48h    = factor(ifelse(!is.na(as.numeric(lfc$padj)) &
                               as.numeric(lfc$padj) < 0.05 &
                               abs(as.numeric(lfc$log2FoldChange)) > 1, "yes", "no"),
                        levels = c("no", "yes")),
    row.names = ids_ordered)

  time_pal <- c(control = "#005f73", `24h` = "#0a9396", `48h` = "#94d2bd")
  cl_pal   <- c(CLBMa = "#E9D8A6", SK_N_SH = "#D9BE6D")
  ann_cols <- list(
    time      = time_pal[levels(col_ann$time)],
    cell_line = cl_pal[levels(col_ann$cell_line)],
    batch     = notch_batch_colors[levels(col_ann$batch)],
    DAR_48h   = c(no = "grey85", yes = "#ae2012"))

  hm <- pheatmap::pheatmap(
    mat, scale = "row",
    cluster_rows = FALSE, cluster_cols = FALSE, gaps_col = gaps_col,
    annotation_col = col_ann, annotation_row = row_ann, annotation_colors = ann_cols,
    show_rownames = nrow(mat) <= 40, show_colnames = FALSE,
    color = colorRampPalette(c("navy", "white", "firebrick3"))(50),
    main = sprintf("%s locus - per-peak accessibility (batch-corrected VST, row-scaled)", g),
    fontsize = 8, fontsize_row = 6)
  h_cm <- max(12, min(45, nrow(mat) * 0.35 + 4))
  ggsave(file.path(notch_dir, sprintf("notch_locus_heatmap_%s.png", g)), hm, width = 20, height = h_cm, units = "cm")
  ggsave(file.path(notch_dir, sprintf("notch_locus_heatmap_%s.pdf", g)), hm, width = 20, height = h_cm, units = "cm")
  invisible(hm)
}
for (g in c("MAML2", "NOTCH1", "NOTCH2")) locus_peak_heatmap(g)
```

```{r notch-SH-significant-peak-boxplots}
# Boxplots of accessibility for the individual Notch-locus peaks that are
# SIGNIFICANT DARs in the SH-specific analysis (48h vs control), shown for both
# cell lines. This isolates the exact regions the SH-only DESeq2 test called
# significant (padj<0.05 & |log2FC|>1) and shows how they behave in SK_N_SH vs
# CLBMa - making clear that the effect is cell-line specific (strong in SH, flat
# in CLBMa) and therefore diluted in the pooled analysis. Accessibility is the
# batch-corrected VST `vsd`, as in the plots above. Boxplots only (no points).
#
# NOTE: only genes that actually have a significant SH peak in their locus are
# plotted. In the current SH 48h results that is MAML2 only - NOTCH1 and NOTCH2
# have no significant peak in the SH-specific analysis (a message lists them).
sh_dar_dir  <- "~/neuroblastoma/results/ATAC-seq/ATAC-seq_drug_treatment_SH_with_bathc_variable/"
sh_sig <- read_xlsx(file.path(sh_dar_dir, "DAR_results_time_48h_vs_control.xlsx"),
                    sheet = "results_signif")

# Notch loci (hg38 gene body + 5 kb flank), same definition as the block above.
notch_loci_sh <- GenomicRanges::GRanges(
  seqnames = c("chr11", "chr9", "chr1"),
  ranges   = IRanges::IRanges(start = c(95976614, 136494433, 119911553),
                              end   = c(96343547, 136545786, 120100779)),
  gene = c("MAML2", "NOTCH1", "NOTCH2"))
notch_loci_sh <- GenomicRanges::resize(
  notch_loci_sh, width = GenomicRanges::width(notch_loci_sh) + 2 * 5000, fix = "center")

# Map each significant SH peak to a Notch locus by genomic overlap.
sh_sig_gr <- GenomicRanges::GRanges(
  seqnames = paste0("chr", sub("^chr", "", as.character(sh_sig$Chr))),
  ranges   = IRanges::IRanges(as.integer(sh_sig$Start), as.integer(sh_sig$End)))
ov_sh <- GenomicRanges::findOverlaps(sh_sig_gr, notch_loci_sh)
sh_notch <- sh_sig[S4Vectors::queryHits(ov_sh), c("PeakId", "Chr", "Start", "End", "log2FoldChange", "padj")]
sh_notch$gene <- notch_loci_sh$gene[S4Vectors::subjectHits(ov_sh)]
sh_notch <- sh_notch[!duplicated(sh_notch$PeakId), ]

genes_without <- setdiff(c("MAML2", "NOTCH1", "NOTCH2"), unique(sh_notch$gene))
if (length(genes_without))
  message("No significant SH-specific peak in locus (not plotted): ",
          paste(genes_without, collapse = ", "))

# Keep only peaks present in the accessibility matrix.
sh_notch <- sh_notch[sh_notch$PeakId %in% as.character(mcols(peak_gr)$PeakId), ]
stopifnot(nrow(sh_notch) > 0)

# Facet label per peak, carrying the SH stats + direction of change.
sh_notch <- sh_notch %>%
  dplyr::mutate(
    direction  = ifelse(as.numeric(log2FoldChange) > 0, "up", "down"),
    peak_label = sprintf("%s | %s (%s)\nchr%s:%s-%s | SH 48h LFC %+.2f, padj %.2g",
                         gene, PeakId, direction, Chr,
                         format(as.integer(Start), big.mark = ","),
                         format(as.integer(End), big.mark = ","),
                         as.numeric(log2FoldChange), as.numeric(padj))) %>%
  dplyr::arrange(gene, dplyr::desc(as.numeric(log2FoldChange)))

# Long accessibility table for exactly these peaks (batch-corrected VST).
sh_acc <- SummarizedExperiment::assay(vsd)[match(sh_notch$PeakId, as.character(mcols(peak_gr)$PeakId)), , drop = FALSE]
rownames(sh_acc) <- sh_notch$PeakId
sh_meta <- as.data.frame(colData(vsd))[, c("cell_line", "time", "batch")]
sh_meta$col <- colnames(vsd)
sh_long <- as.data.frame(sh_acc) %>%
  tibble::rownames_to_column("PeakId") %>%
  tidyr::pivot_longer(-PeakId, names_to = "col", values_to = "accessibility") %>%
  dplyr::left_join(sh_meta, by = "col") %>%
  dplyr::left_join(sh_notch[, c("PeakId", "gene", "peak_label")], by = "PeakId") %>%
  dplyr::mutate(peak_label = factor(peak_label, levels = unique(sh_notch$peak_label)))

sh_time_fill <- c(control = "#005f73", `24h` = "#0a9396", `48h` = "#94d2bd")
sh_box <- ggplot(sh_long, aes(time, accessibility, fill = time)) +
  geom_boxplot(colour = "grey30", width = 0.6, outlier.size = 0.6) +
  facet_grid(peak_label ~ cell_line, scales = "free_y") +
  scale_fill_manual(values = sh_time_fill, guide = "none") +
  labs(title = "Accessibility of SH-significant Notch-locus peaks",
       subtitle = "Peaks called DAR in the SH-only 48h-vs-control analysis; batch-corrected VST",
       x = "K-975 treatment", y = "Accessibility (VST)") +
  theme_bw(base_size = 11) +
  theme(strip.text.y = element_text(size = 7, angle = 0))
print(sh_box)

n_pk <- nrow(sh_notch)
ggsave(file.path(notch_dir, "notch_SH_significant_peaks_boxplots.png"), sh_box,
       width = 18, height = max(10, 4 + n_pk * 5), units = "cm", dpi = 300)
ggsave(file.path(notch_dir, "notch_SH_significant_peaks_boxplots.pdf"), sh_box,
       width = 18, height = max(10, 4 + n_pk * 5), units = "cm")
```

```{r SE analysis}
# load-atac-data results and SEs from previous analysis}

ATAC_dds_results_48h <- read_xlsx(file.path(deg_dir,  paste0("DAR_results_time_48h_vs_control.xlsx")), sheet = "results_signif")
ATAC_dds_results_24h <- read_xlsx(file.path(deg_dir,  paste0("DAR_results_time_24h_vs_control.xlsx")), sheet = "results_signif")


# Load SEs that were determined in 05_Lily_CNR_SuperEnhancer_detection.
CLB_SKN_A_SEs_annot <- read.csv("~/neuroblastoma/temp_results/BEDs/CLB_SKN_A_homer_annot.csv", sep = "\t")
CLB_SKN_M_SEs_annot <- read.csv("~/neuroblastoma/temp_results/BEDs/CLB_SKN_M_homer_annot.csv", sep = "\t")
CLB_SKN_AM_SEs_annot <-read.csv("~/neuroblastoma/temp_results/BEDs/CLB_SKN_AM_homer_annot.csv", sep = "\t")


# Filter regions that present only in SEs
# Convert DESeq2 results into GRanges, that can be used for finding overlap with SEs
ATAC_dds_results_24h_grange <- GRanges(seqnames = paste0("chr", ATAC_dds_results_24h$Chr), 
                                       ranges = IRanges(
                                         start = ATAC_dds_results_24h$Start,
                                         end = ATAC_dds_results_24h$End),
                                       mcols = ATAC_dds_results_24h[, c(1,6:29)]
                                       )
ATAC_dds_results_48h_grange <- GRanges(seqnames = paste0("chr", ATAC_dds_results_48h$Chr), 
                                       ranges = IRanges(
                                         start = ATAC_dds_results_48h$Start,
                                         end = ATAC_dds_results_48h$End),
                                       mcols =  ATAC_dds_results_48h[, c(1,6:29)])

# Convert SEs BEDs into GRanges
CLB_SKN_M_SEs_annot_grange <- GRanges(seqnames = CLB_SKN_M_SEs_annot$Chr, 
                                      ranges = IRanges(
                                        start = CLB_SKN_M_SEs_annot$Start,
                                        end = CLB_SKN_M_SEs_annot$End))
CLB_SKN_A_SEs_annot_grange <- GRanges(seqnames = CLB_SKN_A_SEs_annot$Chr, 
                                      ranges = IRanges(
                                        start = CLB_SKN_A_SEs_annot$Start,
                                        end = CLB_SKN_A_SEs_annot$End))
CLB_SKN_AM_SEs_annot_grange <- GRanges(seqnames = CLB_SKN_AM_SEs_annot$Chr, 
                                       ranges = IRanges(
                                         start = CLB_SKN_AM_SEs_annot$Start,
                                         end = CLB_SKN_AM_SEs_annot$End))


ATAC_24h_AM_overlaps <- ChIPpeakAnno::findOverlapsOfPeaks(ATAC_dds_results_24h_grange,
                                                          CLB_SKN_AM_SEs_annot_grange,
                                                          maxgap = 50)
ATAC_24h_M_overlaps <- ChIPpeakAnno::findOverlapsOfPeaks(ATAC_dds_results_24h_grange,
                                                          CLB_SKN_M_SEs_annot_grange,
                                                          maxgap = 50)
write.csv(ATAC_24h_M_overlaps$overlappingPeaks, 
          file = "~/neuroblastoma/results/ATAC-seq/ATAC-seq_drug_treatment/ATAC_24h_MES_SEs_overlaps.tsv",
          sep = "\t", 
          row.names = TRUE,
          col.names = TRUE)
ATAC_24h_A_overlaps <- ChIPpeakAnno::findOverlapsOfPeaks(ATAC_dds_results_24h_grange,
                                                          CLB_SKN_A_SEs_annot_grange,
                                                          maxgap = 50)



ATAC_48h_AM_overlaps <- ChIPpeakAnno::findOverlapsOfPeaks(ATAC_dds_results_48h_grange,
                                                          CLB_SKN_AM_SEs_annot_grange,
                                                          maxgap = 50)
ATAC_48h_M_overlaps <- ChIPpeakAnno::findOverlapsOfPeaks(ATAC_dds_results_48h_grange,
                                                          CLB_SKN_M_SEs_annot_grange,
                                                          maxgap = 50)
write.csv(ATAC_48h_M_overlaps$overlappingPeaks, 
          file = "~/neuroblastoma/results/ATAC-seq/ATAC-seq_drug_treatment/ATAC_48h_MES_SEs_overlaps.tsv",
          sep = "\t", 
          row.names = TRUE,
          col.names = TRUE)
ATAC_48h_A_overlaps <- ChIPpeakAnno::findOverlapsOfPeaks(ATAC_dds_results_48h_grange,
                                                          CLB_SKN_A_SEs_annot_grange,
                                                          maxgap = 50)
```


Here we identify what are the motifs that that present in differentially accessible regions and how they are differ in control (DMSO) vs 24h and 48h samples.
```{r chromvar-motif-analysis}
# ChromVAR motif enrichment analysis for drug treatment samples
# Identify transcription factor motifs that change accessibility over time
opts <- list()
opts[["tax_group"]] <- "vertebrates"
opts[["species"]] <- "9606"
opts[["collection"]] <- "CORE"
opts[["all_versions"]] <- FALSE
motifsToScan <- TFBSTools::getMatrixSet(JASPAR2022, opts)

# Using different subsets of peaks to check 
subsets_to_run <- c("ALL")

sbst <-  c("ALL")
# Filter peaks with low read counts across samples (minimum 5 reads total)
dim(drg_trtm_ATAC)
counts_consensus_filt <- drg_trtm_ATAC[rowSums(assay(drg_trtm_ATAC)) > 5, ]
dim(counts_consensus_filt)

print(paste("procesing: ", sbst ))

# ChromVar package uses GC content to identify background peaks that are likely to be non-functional and exclude them from further analysis. 
counts_consensus_filt <- chromVAR::addGCBias(counts_consensus_filt, 
                                             genome = BSgenome.Hsapiens.UCSC.hg38) 

# Having corrected for bias, we can use the matchMotifs function to identify motifs under our ATACseq peaks.
motif_matches <- motifmatchr::matchMotifs(pwms = motifsToScan, 
                                          subject = counts_consensus_filt, 
                                          genome = BSgenome.Hsapiens.UCSC.hg38, 
                                          out = "matches")
motif_matches

BiocParallel::register(BiocParallel::MulticoreParam(8, progressbar = FALSE))
set.seed(42069)

# Run chromVar ----
#Following the identification of motifs in our peaks, we can perform the summarization of ATACseq signal to motifs using the computeDeviations and the computeVariability functions.
background_peaks <- chromVAR::getBackgroundPeaks(object = counts_consensus_filt) 
access_expectation <- chromVAR::computeExpectations(object = counts_consensus_filt)
chrom_access_deviations <- chromVAR::computeDeviations(object = counts_consensus_filt, 
                                                       annotations = motif_matches,
                                                       background_peaks = background_peaks,
                                                       expectation = access_expectation)

# Generate sample correlation matrix to assess data quality and batch effects
sample_cor <- chromVAR::getSampleCorrelation(chrom_access_deviations)
annotation_row_sampleCor <- as.data.frame(SummarizedExperiment::colData(chrom_access_deviations)) %>%
  dplyr::select(sample)
pheatmap::pheatmap(as.dist(sample_cor), 
                   annotation_row = annotation_row_sampleCor, 
                   clustering_distance_rows = as.dist(1-sample_cor), 
                   clustering_distance_cols = as.dist(1-sample_cor),
                   annotation_names_row = TRUE 
                  ) 

# Perform t-SNE dimensionality reduction to visualize sample clustering
tsne_results <- chromVAR::deviationsTsne(chrom_access_deviations, 
                                         threshold = 1.5,
                                         perplexity = 8, 
                                         what = "samples",
                                         shiny = FALSE)

tsne_plots <- chromVAR::plotDeviationsTsne(chrom_access_deviations, 
                                           tsne_results,
                                           sample_column = "sample",
                                           shiny = FALSE)

diff_acc <- chromVAR::differentialDeviations(chrom_access_deviations, 
                                             groups = "time",
                                             parametric = FALSE)

motif_annot <- as.data.frame(SummarizedExperiment::rowData(chrom_access_deviations)) %>%
  tibble::rownames_to_column(var = "motif") %>%
  dplyr::select(motif, name)
diff_acc_annot <- diff_acc %>%
  tibble::rownames_to_column(var = "motif") %>%
  dplyr::left_join(., motif_annot, by = "motif")
head(diff_acc_annot) 

diff_var <- chromVAR::differentialVariability(chrom_access_deviations, "time", 
                                              parametric = FALSE)
diff_var_annot <- diff_var %>%
  tibble::rownames_to_column(var = "motif") %>%
  dplyr::left_join(., motif_annot, by = "motif")
head(diff_var_annot)

devZscores <- chromVAR::deviationScores(chrom_access_deviations)
devZscores_df <- tibble::rownames_to_column(as.data.frame(devZscores),
                                            var = "jaspar_id")

#The function computeVariability returns a data.frame that contains the variability (standard deviation of the z scores computed above
# across all cell/samples for a set of peaks), bootstrap confidence intervals for that variability (by resampling cells/samples), 
# and a p-value for the variability being greater than the null hypothesis of 1.
chrom_access_variability <- chromVAR::computeVariability(chrom_access_deviations)
chrom_access_variability_plot <- chromVAR::plotVariability(chrom_access_variability,
                                                           use_plotly = FALSE)

ggsave(chrom_access_variability_plot, 
       filename = file.path(deg_dir, paste0("ChromVAR_chrom_access_variability_corr_plot_", sbst, ".pdf")))

BiocParallel::register(BiocParallel::SerialParam())

chrom_access_variability_ord <- chrom_access_variability[order(chrom_access_variability$p_value), ]  # ordering based on p-value

#! to store
chrom_access_variability_ord_results <- chrom_access_variability_ord %>%
  tibble::rownames_to_column(var="jaspar_id") %>%
  dplyr::left_join(., devZscores_df, by = "jaspar_id")

message("Saving motif enrichment results")
ntop <- 100
topVariable <- chrom_access_variability_ord[1:ntop, ]
topVariable <- tibble::rownames_to_column(topVariable, var="jaspar_id")

devTop <- topVariable %>%
  dplyr::left_join(., devZscores_df, by = "jaspar_id")

devTop <- devTop %>% mutate(long_name = paste0(jaspar_id, "_", name))

devToPlot <- devTop %>%
  dplyr::select(long_name, 
                CM_control_ATAC_S165169_REP1,
                CM_24h_ATAC_S165170_REP1,
                CM_48h_ATAC_S165167_REP1,
                SH_control_ATAC_S165166_REP1,
                SH_24h_ATAC_S165164_REP1,
                SH_48h_ATAC_S165165_REP1,
                SH_DMSO_ctrl_rep1_REP1,
                SH_DMSO_ctrl_rep2_REP1,
                SH_24h_rep1_REP1,     
                SH_24h_rep2_REP1,     
                SH_48h_rep1_REP1,
                SH_48h_rep2_REP1,
                CM_DMSO_ctrl_rep1_REP1,
                CM_DMSO_ctrl_rep2_REP1,
                CM_24h_rep1_REP1,
                CM_24h_rep2_REP1,
                CM_48h_rep1_REP1,
                CM_48h_rep2_REP1
  ) %>%
  tibble::column_to_rownames(var = "long_name")

annotCol_forHeatmap <- as.data.frame(colData(drg_trtm_ATAC)) %>%
  dplyr::select(time)
# annotCol_forHeatmap_colors <- list(phenotype = c(A = "#525252", M = "#fc4e2a"))
annotCol_forHeatmap_colors <- list(phenotype = c(`24h` = "#525252",
                                                 `48h` = "#fc4e2a",
                                                 control = "green"))


color.scheme <- colorRampPalette(c("navy", "white", "firebrick3"))(50)

heatmap <- pheatmap::pheatmap(as.matrix(devToPlot),
                              color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
                              border_color = NA,
                              #color = colorRampPalette(rev(brewer.pal(7, "RdBu")))(100),
                              scale = "row",
                              cluster_cols = FALSE,
                              cluster_rows = TRUE,
                              show_colnames = TRUE,
                              show_rownames = TRUE,
                              annotation_col = annotCol_forHeatmap,
                              annotation_colors = annotCol_forHeatmap_colors,  
                              fontsize_row = 9,
                              fontsize = 9,
                              main = paste0("Heatmap of signif. ATAC-seq Drug treatment control vs 24h vs 48h. ", sbst, "- used"))

ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_drug_treatment_control_vs_24h_48h_Motifs", sbst, "- used", ".png")), 
       plot=heatmap,
       width = 18, height = 40, units = "cm")
ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_drug_treatment_control_vs_24h_48h_Motifs", sbst, "- used", ".pdf")), 
       plot=heatmap,
       width = 18, height = 40, units = "cm")

# Generate separate heatmap for CM cell line only
devToPlot_CM <- devTop %>%
  dplyr::select(long_name, 
                CM_control_ATAC_S165169_REP1,
                CM_DMSO_ctrl_rep1_REP1,
                CM_DMSO_ctrl_rep2_REP1,
                CM_24h_ATAC_S165170_REP1,
                CM_24h_rep1_REP1,
                CM_24h_rep2_REP1,
                CM_48h_ATAC_S165167_REP1,
                CM_48h_rep1_REP1,
                CM_48h_rep2_REP1 
  ) %>%
  tibble::column_to_rownames(var = "long_name")

heatmap <- pheatmap::pheatmap(as.matrix(devToPlot_CM),
                              color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
                              border_color = NA,
                              #color = colorRampPalette(rev(brewer.pal(7, "RdBu")))(100),
                              scale = "row",
                              cluster_cols = FALSE,
                              cluster_rows = TRUE,
                              show_colnames = TRUE,
                              show_rownames = TRUE,
                              annotation_col = annotCol_forHeatmap,
                              annotation_colors = annotCol_forHeatmap_colors,  
                              fontsize_row = 9,
                              fontsize = 9,
                              main = paste0("Heatmap of signif. CM line ATAC-seq Drug treatment control vs 24h vs 48h. ", sbst, "- used"))

ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_CM_line_drug_treatment_control_vs_24h_48h_Motifs_", sbst, "- used", ".png")), 
       plot=heatmap,
       width = 18, height = 40, units = "cm")
ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_CM_line_drug_treatment_control_vs_24h_48h_Motifs_", sbst, "- used", ".pdf")), 
       plot=heatmap,
       width = 18, height = 40, units = "cm")


# Generate separate heatmap for SH cell line only
devToPlot_SH <- devTop %>%
  dplyr::select(long_name, 
                SH_control_ATAC_S165166_REP1,
                SH_DMSO_ctrl_rep1_REP1,
                SH_DMSO_ctrl_rep2_REP1,
                SH_24h_ATAC_S165164_REP1,
                SH_24h_rep1_REP1,     
                SH_24h_rep2_REP1,     
                SH_48h_ATAC_S165165_REP1,
                SH_48h_rep1_REP1,
                SH_48h_rep2_REP1,
  ) %>%
  tibble::column_to_rownames(var = "long_name")

heatmap <- pheatmap::pheatmap(as.matrix(devToPlot_SH),
                              color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
                              border_color = NA,
                              #color = colorRampPalette(rev(brewer.pal(7, "RdBu")))(100),
                              scale = "row",
                              cluster_cols = FALSE,
                              cluster_rows = TRUE,
                              show_colnames = TRUE,
                              show_rownames = TRUE,
                              annotation_col = annotCol_forHeatmap,
                              annotation_colors = annotCol_forHeatmap_colors,  
                              fontsize_row = 9,
                              fontsize = 9,
                              main = paste0("Heatmap of signif. SH line ATAC-seq Drug treatment control vs 24h vs 48h. ", sbst, "- used"))

ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_SH_line_drug_treatment_control_vs_24h_48h_Motifs_", sbst, "- used", ".png")), 
       plot=heatmap,
       width = 18, height = 40, units = "cm")
ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_SH_line_drug_treatment_control_vs_24h_48h_Motifs_", sbst, "- used", ".pdf")),
       plot=heatmap,
       width = 18, height = 40, units = "cm")


# Generate heatmap with averaged replicates: 6 columns (control/24h/48h x CM/SH)
# Each group's replicates (legacy single-rep + new rep1/rep2) are averaged into one column.
replicate_groups <- list(
  control_CM = c("CM_control_ATAC_S165169_REP1", "CM_DMSO_ctrl_rep1_REP1", "CM_DMSO_ctrl_rep2_REP1"),
  control_SH = c("SH_control_ATAC_S165166_REP1", "SH_DMSO_ctrl_rep1_REP1", "SH_DMSO_ctrl_rep2_REP1"),
  `24h_CM`   = c("CM_24h_ATAC_S165170_REP1", "CM_24h_rep1_REP1", "CM_24h_rep2_REP1"),
  `24h_SH`   = c("SH_24h_ATAC_S165164_REP1", "SH_24h_rep1_REP1", "SH_24h_rep2_REP1"),
  `48h_CM`   = c("CM_48h_ATAC_S165167_REP1", "CM_48h_rep1_REP1", "CM_48h_rep2_REP1"),
  `48h_SH`   = c("SH_48h_ATAC_S165165_REP1", "SH_48h_rep1_REP1", "SH_48h_rep2_REP1")
)

devToPlot_avg <- sapply(replicate_groups, function(cols) {
  rowMeans(as.matrix(devToPlot[, cols, drop = FALSE]))
})
devToPlot_avg <- as.data.frame(devToPlot_avg)

# Column annotation for the 6 averaged groups
annotCol_avg <- data.frame(
  time      = c("control", "control", "24h", "24h", "48h", "48h"),
  cell_line = c("CM", "SH", "CM", "SH", "CM", "SH"),
  row.names = colnames(devToPlot_avg)
)
annotCol_avg_colors <- list(
  time = c(control = "green", `24h` = "#525252", `48h` = "#fc4e2a"),
  cell_line = c(CM = "#1f78b4", SH = "#33a02c")
)

heatmap <- pheatmap::pheatmap(as.matrix(devToPlot_avg),
                              color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
                              border_color = NA,
                              scale = "row",
                              cluster_cols = FALSE,
                              cluster_rows = TRUE,
                              show_colnames = TRUE,
                              show_rownames = TRUE,
                              annotation_col = annotCol_avg,
                              annotation_colors = annotCol_avg_colors,
                              fontsize_row = 9,
                              fontsize = 9,
                              main = paste0("Heatmap of signif. ATAC-seq Drug treatment (averaged replicates). ", sbst, "- used"))

ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_drug_treatment_averaged_replicates_Motifs_", sbst, "- used", ".png")),
       plot=heatmap,
       width = 18, height = 40, units = "cm")
ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_drug_treatment_averaged_replicates_Motifs_", sbst, "- used", ".pdf")),
       plot=heatmap,
       width = 18, height = 40, units = "cm")


# Generate heatmap averaged over replicates AND cell lines: 3 columns (control/24h/48h)
# Each time point averages all replicates of both CM and SH cell lines into one column.
time_groups <- list(
  control = c("CM_control_ATAC_S165169_REP1", "CM_DMSO_ctrl_rep1_REP1", "CM_DMSO_ctrl_rep2_REP1",
              "SH_control_ATAC_S165166_REP1", "SH_DMSO_ctrl_rep1_REP1", "SH_DMSO_ctrl_rep2_REP1"),
  `24h`   = c("CM_24h_ATAC_S165170_REP1", "CM_24h_rep1_REP1", "CM_24h_rep2_REP1",
              "SH_24h_ATAC_S165164_REP1", "SH_24h_rep1_REP1", "SH_24h_rep2_REP1"),
  `48h`   = c("CM_48h_ATAC_S165167_REP1", "CM_48h_rep1_REP1", "CM_48h_rep2_REP1",
              "SH_48h_ATAC_S165165_REP1", "SH_48h_rep1_REP1", "SH_48h_rep2_REP1")
)

devToPlot_time_avg <- sapply(time_groups, function(cols) {
  rowMeans(as.matrix(devToPlot[, cols, drop = FALSE]))
})
devToPlot_time_avg <- as.data.frame(devToPlot_time_avg)

# Column annotation for the 3 time-averaged groups
annotCol_time_avg <- data.frame(
  time      = c("control", "24h", "48h"),
  row.names = colnames(devToPlot_time_avg)
)
annotCol_time_avg_colors <- list(
  time = c(control = "green", `24h` = "#525252", `48h` = "#fc4e2a")
)

heatmap <- pheatmap::pheatmap(as.matrix(devToPlot_time_avg),
                              color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
                              border_color = NA,
                              scale = "row",
                              cluster_cols = FALSE,
                              cluster_rows = TRUE,
                              show_colnames = TRUE,
                              show_rownames = TRUE,
                              annotation_col = annotCol_time_avg,
                              annotation_colors = annotCol_time_avg_colors,
                              fontsize_row = 9,
                              fontsize = 9,
                              main = paste0("Heatmap of signif. ATAC-seq Drug treatment (averaged replicates & cell lines). ", sbst, "- used"))

ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_drug_treatment_averaged_replicates_celllines_Motifs_", sbst, "- used", ".png")),
       plot=heatmap,
       width = 18, height = 40, units = "cm")
ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_drug_treatment_averaged_replicates_celllines_Motifs_", sbst, "- used", ".pdf")),
       plot=heatmap,
       width = 18, height = 40, units = "cm")




# --- Heatmaps restricted to a curated list of motifs ------------------------
# Motifs to visualise (long_name = jaspar_id + "_" + motif name)
motifs_to_visualise <- c(
  "MA0808.1_TEAD3",
  "MA0090.3_TEAD1",
  "MA0809.2_TEAD4",
  "MA1121.1_TEAD2",
  "MA0496.3_MAFK",
  "MA1928.1_BNC2",
  "MA0835.2_BATF3",
  "MA0462.2_BATF::JUN",
  "MA1634.1_BATF",
  "MA0501.1_MAF::NFE2",
  "MA0089.2_MAFG::NFE2L1",
  "MA1101.2_BACH2",
  "MA0655.1_JDP2",
  "MA0490.2_JUNB",
  "MA0491.2_JUND",
  "MA0477.2_FOSL1",
  "MA1137.1_FOSL1::JUNB",
  "MA1141.1_FOS::JUND",
  "MA1132.1_JUN::JUND",
  "MA1134.1_FOS::JUNB",
  "MA1128.1_FOSL1::JUN",
  "MA1138.1_FOSL2::JUNB",
  "MA1135.1_FOSB::JUNB",
  "MA1130.1_FOSL2::JUN",
  "MA1144.1_FOSL2::JUND",
  "MA0478.1_FOSL2",
  "MA0476.1_FOS",
  "MA0099.3_FOS::JUN",
  "MA1142.1_FOSL1::JUND",
  "MA0841.1_NFE2",
  "MA1633.2_BACH1"
)

# Full deviation-score matrix over ALL motifs, indexed by long_name
devAll <- devZscores_df %>%
  dplyr::left_join(motif_annot, by = c("jaspar_id" = "motif")) %>%
  dplyr::mutate(long_name = paste0(jaspar_id, "_", name)) %>%
  dplyr::select(-jaspar_id, -name) %>%
  tibble::column_to_rownames(var = "long_name")

# Keep requested motifs that are present, preserving the requested order
present_motifs <- motifs_to_visualise[motifs_to_visualise %in% rownames(devAll)]
missing_motifs <- setdiff(motifs_to_visualise, rownames(devAll))
if (length(missing_motifs) > 0) {
  message("Motifs not found and skipped: ", paste(missing_motifs, collapse = ", "))
}
devSel <- devAll[present_motifs, , drop = FALSE]

# 6-column heatmap: replicates averaged, cell lines kept separate
devSel_avg <- sapply(replicate_groups, function(cols) {
  rowMeans(as.matrix(devSel[, cols, drop = FALSE]))
})
devSel_avg <- as.data.frame(devSel_avg)

heatmap <- pheatmap::pheatmap(as.matrix(devSel_avg),
                              color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
                              border_color = NA,
                              scale = "row",
                              cluster_cols = FALSE,
                              cluster_rows = TRUE,
                              show_colnames = TRUE,
                              show_rownames = TRUE,
                              annotation_col = annotCol_avg,
                              annotation_colors = annotCol_avg_colors,
                              fontsize_row = 9,
                              fontsize = 9,
                              main = paste0("Heatmap of selected motifs ATAC-seq Drug treatment (averaged replicates). ", sbst, "- used"))

ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_drug_treatment_selected_motifs_averaged_replicates_", sbst, "- used", ".png")),
       plot=heatmap,
       width = 18, height = 20, units = "cm")
ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_drug_treatment_selected_motifs_averaged_replicates_", sbst, "- used", ".pdf")),
       plot=heatmap,
       width = 18, height = 20, units = "cm")

# 3-column heatmap: replicates AND cell lines averaged
devSel_time_avg <- sapply(time_groups, function(cols) {
  rowMeans(as.matrix(devSel[, cols, drop = FALSE]))
})
devSel_time_avg <- as.data.frame(devSel_time_avg)

heatmap <- pheatmap::pheatmap(as.matrix(devSel_time_avg),
                              color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
                              border_color = NA,
                              scale = "row",
                              cluster_cols = FALSE,
                              cluster_rows = TRUE,
                              show_colnames = TRUE,
                              show_rownames = TRUE,
                              annotation_col = annotCol_time_avg,
                              annotation_colors = annotCol_time_avg_colors,
                              fontsize_row = 9,
                              fontsize = 9,
                              main = paste0("Heatmap of selected motifs ATAC-seq Drug treatment (averaged replicates & cell lines). ", sbst, "- used"))

ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_drug_treatment_selected_motifs_averaged_replicates_celllines_", sbst, "- used", ".png")),
       plot=heatmap,
       width = 18, height = 20, units = "cm")
ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_drug_treatment_selected_motifs_averaged_replicates_celllines_", sbst, "- used", ".pdf")),
       plot=heatmap,
       width = 18, height = 20, units = "cm")

# All-replicates heatmap: curated motifs (see motifs_to_visualise) across every
# individual sample column (same samples/order as devToPlot), no averaging.
devSel_all <- devSel[, colnames(devToPlot), drop = FALSE]

heatmap <- pheatmap::pheatmap(as.matrix(devSel_all),
                              color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
                              border_color = NA,
                              scale = "row",
                              cluster_cols = FALSE,
                              cluster_rows = TRUE,
                              show_colnames = TRUE,
                              show_rownames = TRUE,
                              annotation_col = annotCol_forHeatmap,
                              annotation_colors = annotCol_forHeatmap_colors,
                              fontsize_row = 9,
                              fontsize = 9,
                              main = paste0("Heatmap of selected motifs ATAC-seq Drug treatment (all replicates). ", sbst, "- used"))

ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_drug_treatment_selected_motifs_all_replicates_", sbst, "- used", ".png")),
       plot=heatmap,
       width = 18, height = 20, units = "cm")
ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_drug_treatment_selected_motifs_all_replicates_", sbst, "- used", ".pdf")),
       plot=heatmap,
       width = 18, height = 20, units = "cm")
```

The heatmaps are row-scaled, so they hide the absolute magnitude of the batch
offset. To see how the raw chromVAR deviation z-score for a single motif actually
moves over the treatment time course - and how differently the legacy vs new
batch behave - we plot it directly. Each line is the mean per cell line x batch;
points are individual replicates.
```{r fosl1jun-lineplot}
# Track one motif's chromVAR deviation z-score across the time course, split by batch
motif_of_interest <- "MA1128.1"          # FOSL1::JUN
motif_label       <- "MA1128.1_FOSL1::JUN"

motif_of_interest <- "MA0808.1"          # FOSL1::JUN
motif_label       <- "MA0808.1_TEAD3"

# Per-sample metadata (rownames == column names of the deviation-score matrix)
motif_meta <- as.data.frame(colData(drg_trtm_ATAC)) %>%
  tibble::rownames_to_column(var = "sample_id") %>%
  dplyr::select(sample_id, cell_line, time, batch)

# Long table of the z-scores for this motif, one row per sample
motif_scores <- devZscores_df %>%
  dplyr::filter(str_detect(jaspar_id, motif_of_interest)) %>%
  tidyr::pivot_longer(-jaspar_id, names_to = "sample_id", values_to = "dev_zscore") %>%
  dplyr::left_join(motif_meta, by = "sample_id") %>%
  dplyr::mutate(
    time  = factor(time, levels = c("control", "24h", "48h")),
    group = paste(cell_line, batch, sep = " - ")
  )

if (nrow(motif_scores) == 0) {
  warning("Motif ", motif_of_interest, " not found in devZscores_df - skipping line plot.")
} else {
  # Mean per cell_line x batch x time drives the connecting lines
  motif_summary <- motif_scores %>%
    dplyr::group_by(cell_line, batch, group, time) %>%
    dplyr::summarise(mean_z = mean(dev_zscore), .groups = "drop")

  fosl1jun_lineplot <- ggplot(
      motif_scores,
      aes(x = time, y = dev_zscore, color = cell_line, linetype = batch, group = group)
    ) +
    geom_point(size = 2, alpha = 0.7,
               position = position_dodge(width = 0.15)) +
    geom_line(data = motif_summary, aes(y = mean_z), linewidth = 1) +
    labs(
      title    = paste0("chromVAR deviation z-score: ", motif_label),
      subtitle = "Line = mean per cell line x batch; points = individual replicates",
      x = "Treatment time", y = "Deviation z-score",
      color = "Cell line", linetype = "Batch"
    ) +
    theme_bw()

  print(fosl1jun_lineplot)

  ggsave(filename = file.path(deg_dir, paste0("lineplot_", motif_label, "_dev_zscore_by_batch.png")),
         plot = fosl1jun_lineplot, width = 18, height = 12, units = "cm")
  ggsave(filename = file.path(deg_dir, paste0("lineplot_", motif_label, "_dev_zscore_by_batch.pdf")),
         plot = fosl1jun_lineplot, width = 18, height = 12, units = "cm")
}
```


We want to check how TEAD/AP1 motifs are co-localized in the 24h sample. We will test the real data and make the comparison to simultaed data.
```{r Comparison to simulated data }
# Compare real vs simulated data to assess enrichment significance
h24_control_specific_peaks <- ATAC_dds_results_24h$results_signif %>% 
  dplyr::filter(log2FoldChange < 0) %>% 
  dplyr::pull(PeakId)
h24_specific_peaks <- ATAC_dds_results_24h$results_signif %>% 
  dplyr::filter(log2FoldChange > 0) %>% 
  dplyr::pull(PeakId)

opts <- list()
opts[["tax_group"]] <- "vertebrates"
opts[["species"]] <- "9606"
opts[["collection"]] <- "CORE"
opts[["all_versions"]] <- FALSE
motifsToScan <- TFBSTools::getMatrixSet(JASPAR2022, opts)
TEAD_AP1_motifs_to_scan_names <- c("MA0090.3", "MA0808.1", 
                                 "MA0809.2", "MA1121.1",
                                 "MA0099.3")
TEAD_AP1_motifsToScan <- motifsToScan[TEAD_AP1_motifs_to_scan_names,]

# Using different subsets of peaks to check 
drg_trtm_ATAC
counts_consensus_filt_control_24h <- drg_trtm_ATAC[drg_trtm_ATAC@rowRanges@elementMetadata@listData[["PeakId"]] %in% h24_control_specific_peaks, ]
counts_consensus_filt_24h <- drg_trtm_ATAC[drg_trtm_ATAC@rowRanges@elementMetadata@listData[["PeakId"]] %in% h24_specific_peaks, ]

# Generate randomized peak sets as background controls
# Create matched random regions for statistical comparison
# Load genome
genome_used <- getBSgenome("BSgenome.Hsapiens.UCSC.hg38")
seqnames(genome_used) <- sub("^.{1,2}_", "", seqnames(genome_used))
seqnames(genome_used) <- sub("v", ".", seqnames(genome_used))
seqnames(genome_used) <- sub("^M", "MT", seqnames(genome_used))

# load mask regions 
blacklist_regions <- read.delim("~/neuroblastoma/resources/hg38-blacklist.v2.bed", header = FALSE)
blacklist_regions <- makeGRangesFromDataFrame(blacklist_regions, 
                                              seqnames.field = "V1", 
                                              start.field = "V2", 
                                              end.field = "V3")

# Create set of random Peaks for control_24h 
initial_peak_set <- reduce(counts_consensus_filt_control_24h@rowRanges)
random_peaks_control_24h <- initial_peak_set
random_peaks_control_24h@seqnames <- droplevels(random_peaks_control_24h@seqnames)
random_peaks_control_24h <- regioneR::randomizeRegions(random_peaks_control_24h,
                                               allow.overlaps = FALSE,
                                               genome = genome_used,
                                               per.chromosome = TRUE,
                                               mask = blacklist_regions)
random_peaks_control_24h <- as.data.frame(random_peaks_control_24h, 
                                          row.names = NULL,
                                          optional = FALSE)
random_peaks_control_24h$seqnames <- droplevels(random_peaks_control_24h$seqnames)
random_peaks_control_24h <- makeGRangesFromDataFrame(random_peaks_control_24h)

# Create set of random Peaks for 24h 
initial_peak_set <- reduce(counts_consensus_filt_24h@rowRanges)
random_peaks_24h <- initial_peak_set
random_peaks_24h@seqnames <- droplevels(random_peaks_24h@seqnames)
random_peaks_24h <- regioneR::randomizeRegions(random_peaks_24h,
                                               allow.overlaps = FALSE,
                                               genome = genome_used,
                                               per.chromosome = TRUE,
                                               mask = blacklist_regions)
random_peaks_24h <- as.data.frame(random_peaks_24h, 
                                  row.names = NULL,
                                  optional = FALSE)
random_peaks_24h$seqnames <- droplevels(random_peaks_24h$seqnames)
random_peaks_24h <- makeGRangesFromDataFrame(random_peaks_24h)

counts_consensus_filt_control_24h <- chromVAR::addGCBias(counts_consensus_filt_control_24h, 
                                                         genome = BSgenome.Hsapiens.UCSC.hg38) 
counts_consensus_filt_24h <- chromVAR::addGCBias(counts_consensus_filt_24h, 
                                                 genome = BSgenome.Hsapiens.UCSC.hg38) 

# find TEAD and AP1 motifs in real control_24h data
motif_matches_control_24h <- motifmatchr::matchMotifs(pwms = TEAD_AP1_motifsToScan, 
                                              subject = counts_consensus_filt_control_24h, 
                                              genome = BSgenome.Hsapiens.UCSC.hg38, 
                                              out = "positions")
control_24h_TEAD <- c(motif_matches_control_24h$MA0090.3, 
                      motif_matches_control_24h$MA0808.1, 
                      motif_matches_control_24h$MA0809.2, 
                      motif_matches_control_24h$MA1121.1)
control_24h_TEAD <- reduce(control_24h_TEAD)
control_24h_AP1 <- reduce(motif_matches_control_24h$MA0099.3)
control_24h_AP1_TEAD_overlaps <- findOverlaps(control_24h_AP1, 
                                              control_24h_TEAD, 
                                              ignore.strand = TRUE, 
                                              maxgap = 50)

# find TEAD and AP1 motifs in simulated MES data
motif_matches_control_24h_random <- motifmatchr::matchMotifs(pwms = TEAD_AP1_motifsToScan, 
                                                     subject = random_peaks_control_24h, 
                                                     genome = BSgenome.Hsapiens.UCSC.hg38, 
                                                     out = "positions")
control_24h_random_TEAD <- c(motif_matches_control_24h_random$MA0090.3, 
                             motif_matches_control_24h_random$MA0808.1, 
                             motif_matches_control_24h_random$MA0809.2, 
                             motif_matches_control_24h_random$MA1121.1)
control_24h_random_TEAD <- reduce(control_24h_random_TEAD)
control_24h_random_AP1 <- reduce(motif_matches_control_24h_random$MA0099.3)
control_24h_random_AP1_TEAD_overlaps <- findOverlaps(control_24h_random_AP1, 
                                                     control_24h_random_TEAD, 
                                                     ignore.strand = TRUE, 
                                                     maxgap = 50)

# find TEAD and AP1 motifs in real 24h data
motif_matches_24h <- motifmatchr::matchMotifs(pwms = TEAD_AP1_motifsToScan, 
                                              subject = counts_consensus_filt_24h, 
                                              genome = BSgenome.Hsapiens.UCSC.hg38, 
                                              out = "positions")
h24_TEAD <- c(motif_matches_24h$MA0090.3, 
              motif_matches_24h$MA0808.1, 
              motif_matches_24h$MA0809.2, 
              motif_matches_24h$MA1121.1)
h24_TEAD <- reduce(h24_TEAD)
h24_AP1 <- reduce(motif_matches_24h$MA0099.3)
h24_AP1_TEAD_overlaps <- findOverlaps(h24_AP1, 
                                      h24_TEAD, 
                                      ignore.strand = TRUE, 
                                      maxgap = 50)

# find TEAD and AP1 motifs in simulated 24h data
motif_matches_24h_random <- motifmatchr::matchMotifs(pwms = TEAD_AP1_motifsToScan, 
                                                     subject = random_peaks_24h, 
                                                     genome = BSgenome.Hsapiens.UCSC.hg38, 
                                                     out = "positions")
h24_random_TEAD <- c(motif_matches_24h_random$MA0090.3, 
                     motif_matches_24h_random$MA0808.1, 
                     motif_matches_24h_random$MA0809.2, 
                     motif_matches_24h_random$MA1121.1)
h24_random_TEAD <- reduce(h24_random_TEAD)
h24_random_AP1 <- reduce(motif_matches_24h_random$MA0099.3)
h24_random_AP1_TEAD_overlaps <- findOverlaps(h24_random_AP1, 
                                             h24_random_TEAD, 
                                             ignore.strand = TRUE, 
                                             maxgap = 50)


summary_table_peaks <- data.frame(total_number_of_peaks = c(length(counts_consensus_filt_control_24h), 
                                                            length(random_peaks_control_24h),
                                                            length(counts_consensus_filt_24h),
                                                            length(random_peaks_24h)),
                                  TEAD_sites = c(length(control_24h_TEAD),
                                                 length(control_24h_random_TEAD),
                                                 length(h24_TEAD),
                                                 length(h24_random_TEAD)),
                                  AP1_sites = c(length(control_24h_AP1),
                                                length(control_24h_random_AP1),
                                                length(h24_AP1),
                                                length(h24_random_AP1)),
                                  TEAD_AP1_coloc_sites = c(length(control_24h_AP1_TEAD_overlaps),
                                                           length(control_24h_random_AP1_TEAD_overlaps),
                                                           length(h24_AP1_TEAD_overlaps),
                                                           length(h24_random_AP1_TEAD_overlaps)),
                                  row.names = c("control_24h", 
                                                "Simulated control_24h", 
                                                "h24", 
                                                "Simulated h24")
)
summary_table_peaks

# Perform Fisher's exact tests to assess statistical significance of TEAD/AP1 co-localization
contingency_table <- matrix(c(length(h24_AP1_TEAD_overlaps), 
                              length(h24_TEAD) + length(h24_AP1) - length(h24_AP1_TEAD_overlaps), 
                              length(control_24h_AP1_TEAD_overlaps), 
                              length(control_24h_TEAD) + length(control_24h_AP1) - length(control_24h_AP1_TEAD_overlaps)), 
                            nrow = 2, 
                            byrow = TRUE)
fisher_results <- fisher.test(contingency_table, alternative = "two.sided")
fisher_summary_table <- data.table::data.table(comparison = "24h_vs_control_24h",
                                               pval = fisher_results$p.value,
                                               odds_ratio = fisher_results$estimate)

# Fisher test comparing real h24 AP1_TEAD peaks vs simulated
contingency_table_h24_vs_simulated <- matrix(c(length(h24_AP1_TEAD_overlaps), 
                                               length(h24_TEAD) + length(h24_AP1) - length(h24_AP1_TEAD_overlaps),
                                               length(h24_random_AP1_TEAD_overlaps), 
                                               length(h24_random_TEAD) + length(h24_random_AP1) - length(h24_random_AP1_TEAD_overlaps)), 
                                             nrow = 2, 
                                             byrow = TRUE)
row.names(contingency_table_h24_vs_simulated) <- c("h24_peaks", "h24_peaks_random")
colnames(contingency_table_h24_vs_simulated) <- c("TEAD_AB1_colocalisation_present", "TEAD_AB1_colocalisation_absent")
contingency_table_h24_vs_simulated
fisher_results <- fisher.test(contingency_table_h24_vs_simulated, alternative="two.sided")
fisher_summary_table <- rbind(fisher_summary_table,
                              data.table::data.table(comparison = "h24_vs_h24sim",
                                                     pval = fisher_results$p.value,
                                                     odds_ratio = fisher_results$estimate))

# Fisher test for TEAD peaks
contingency_table_TEAD <- matrix(c(length(h24_TEAD), 
                                   length(counts_consensus_filt_24h) - length(h24_TEAD),
                                   length(control_24h_TEAD), 
                                   length(counts_consensus_filt_control_24h) - length(control_24h_TEAD)), 
                                 nrow = 2, 
                                 byrow = TRUE)
row.names(contingency_table_TEAD) <- c("h24_peaks", "control_24h_peaks")
colnames(contingency_table_TEAD) <- c("TEAD_site_present", "TEAD_site_absent")
contingency_table_TEAD
fisher.test(contingency_table_TEAD, 
            alternative = "two.sided")
chisq.test(contingency_table_TEAD)

# Fisher test for AP1 peaks
contingency_table_AP1 <- matrix(c(length(h24_AP1), 
                                  length(counts_consensus_filt_24h) - length(h24_AP1),
                                  length(control_24h_AP1), 
                                  length(counts_consensus_filt_control_24h) - length(control_24h_AP1)),
                                nrow = 2,
                                byrow = TRUE)
row.names(contingency_table_AP1) <- c("h24_peaks", "control_24h_peaks")
colnames(contingency_table_AP1) <- c("AP1_site_present", "AP1_site_absent")
contingency_table_AP1
fisher.test(contingency_table_AP1, alternative="two.sided")
chisq.test(contingency_table_AP1)

```

```{r tead-ap1-analysis-48h}
# TEAD/AP1 motif co-localization analysis for 48h treatment
# Repeat analysis for longer treatment timepoint
drg_trtm_ATAC
ATAC_dds_results_control_48h <- ATAC_dds_results_48h$results_signif %>% 
  dplyr::filter(log2FoldChange < 0) %>% 
  dplyr::pull(PeakId)
h48_specific_peaks <- ATAC_dds_results_48h$results_signif %>% 
  dplyr::filter(log2FoldChange > 0) %>% 
  dplyr::pull(PeakId)

opts <- list()
opts[["tax_group"]] <- "vertebrates"
opts[["species"]] <- "9606"
opts[["collection"]] <- "CORE"
opts[["all_versions"]] <- FALSE
motifsToScan <- TFBSTools::getMatrixSet(JASPAR2022, opts)
TEAD_AP1_motifsToScan_names <- c("MA0090.3", 
                                 "MA0808.1", 
                                 "MA0809.2", 
                                 "MA1121.1",
                                 "MA0099.3")
TEAD_AP1_motifsToScan <- motifsToScan[TEAD_AP1_motifsToScan_names,]

# Using different subsets of peaks to check 
drg_trtm_ATAC
counts_consensus_filt_control_48h <- drg_trtm_ATAC[drg_trtm_ATAC@rowRanges@elementMetadata@listData[["PeakId"]] %in% ATAC_dds_results_control_48h, ]
counts_consensus_filt_48h <- drg_trtm_ATAC[drg_trtm_ATAC@rowRanges@elementMetadata@listData[["PeakId"]] %in% h48_specific_peaks, ]

# Create a set of random peaks for 48h 48h and control_48h
# Load genome
genome_used <- getBSgenome("BSgenome.Hsapiens.UCSC.hg38")
seqnames(genome_used) <- sub("^.{1,2}_", "", seqnames(genome_used))
seqnames(genome_used) <- sub("v", ".", seqnames(genome_used))
seqnames(genome_used) <- sub("^M", "MT", seqnames(genome_used))
# load mask regions 
blacklist_regions <- read.delim("~/workspace/neuroblastoma/resources/hg38-blacklist.v2.bed", header = FALSE)
blacklist_regions <- makeGRangesFromDataFrame(blacklist_regions, 
                                              seqnames.field = "V1", 
                                              start.field = "V2", 
                                              end.field = "V3")

# Create set of random Peaks for control_48h 
initial_peak_set <- reduce(counts_consensus_filt_control_48h@rowRanges)
random_peaks_control_48h <- initial_peak_set
random_peaks_control_48h@seqnames <- droplevels(random_peaks_control_48h@seqnames)
random_peaks_control_48h <- regioneR::randomizeRegions(random_peaks_control_48h,
                                                   allow.overlaps = FALSE,
                                                   genome = genome_used,
                                                   per.chromosome = TRUE,
                                                   mask = blacklist_regions)
random_peaks_control_48h <- as.data.frame(random_peaks_control_48h, row.names = NULL, optional = FALSE)
random_peaks_control_48h$seqnames <- droplevels(random_peaks_control_48h$seqnames)
random_peaks_control_48h <- makeGRangesFromDataFrame(random_peaks_control_48h)

# Create set of random Peaks for 48h 
initial_peak_set <- reduce(counts_consensus_filt_48h@rowRanges)
random_peaks_48h <- initial_peak_set
random_peaks_48h@seqnames <- droplevels(random_peaks_48h@seqnames)
random_peaks_48h <- regioneR::randomizeRegions(random_peaks_48h,
                                               allow.overlaps = FALSE,
                                               genome = genome_used,
                                               per.chromosome = TRUE,
                                               mask = blacklist_regions)
random_peaks_48h <- as.data.frame(random_peaks_48h, row.names = NULL, optional = FALSE)
random_peaks_48h$seqnames <- droplevels(random_peaks_48h$seqnames)
random_peaks_48h <- makeGRangesFromDataFrame(random_peaks_48h)

counts_consensus_filt_control_48h <- chromVAR::addGCBias(counts_consensus_filt_control_48h,
                                                         genome = BSgenome.Hsapiens.UCSC.hg38) 
counts_consensus_filt_48h <- chromVAR::addGCBias(counts_consensus_filt_48h,
                                                 genome = BSgenome.Hsapiens.UCSC.hg38) 

# Having corrected for bias, we can use the matchMotifs function to identify motifs under our ATACseq peaks.
# Here we supply our RangedSummarizedExperiment of counts in peaks and the genome of interest to the matchMotifs function and use the default out of matches.

# find TEAD and AP1 motifs in real control_48h data
motif_matches_control_48h <- motifmatchr::matchMotifs(pwms = TEAD_AP1_motifsToScan, 
                                                  subject = counts_consensus_filt_control_48h, 
                                                  genome = BSgenome.Hsapiens.UCSC.hg38, 
                                                  out = "positions")
control_48h_TEAD <- c(motif_matches_control_48h$MA0090.3, 
                      motif_matches_control_48h$MA0808.1, 
                      motif_matches_control_48h$MA0809.2, 
                      motif_matches_control_48h$MA1121.1)
control_48h_TEAD <- reduce(control_48h_TEAD)
control_48h_AP1 <- reduce(motif_matches_control_48h$MA0099.3)
control_48h_AP1_TEAD_overlaps <- findOverlaps(control_48h_AP1, 
                                              control_48h_TEAD, 
                                              ignore.strand = TRUE, 
                                              maxgap = 50)

# find TEAD and AP1 motifs in simulated MES data
motif_matches_control_48h_random <- motifmatchr::matchMotifs(pwms = TEAD_AP1_motifsToScan, 
                                                         subject = random_peaks_control_48h, 
                                                         genome = BSgenome.Hsapiens.UCSC.hg38, 
                                                         out = "positions")
control_48h_random_TEAD <- c(motif_matches_control_48h_random$MA0090.3, 
                             motif_matches_control_48h_random$MA0808.1, 
                             motif_matches_control_48h_random$MA0809.2, 
                             motif_matches_control_48h_random$MA1121.1)
control_48h_random_TEAD <- reduce(control_48h_random_TEAD)
control_48h_random_AP1 <- reduce(motif_matches_control_48h_random$MA0099.3)
control_48h_random_AP1_TEAD_overlaps <- findOverlaps(control_48h_random_AP1, 
                                                     control_48h_random_TEAD, 
                                                     ignore.strand = TRUE,
                                                     maxgap = 50)

# find TEAD and AP1 motifs in real 48h data
motif_matches_48h <- motifmatchr::matchMotifs(pwms = TEAD_AP1_motifsToScan, 
                                              subject = counts_consensus_filt_48h, 
                                              genome = BSgenome.Hsapiens.UCSC.hg38, 
                                              out = "positions")
h48_TEAD <- c(motif_matches_48h$MA0090.3, 
              motif_matches_48h$MA0808.1,
              motif_matches_48h$MA0809.2, 
              motif_matches_48h$MA1121.1)
h48_TEAD <- reduce(h48_TEAD)
h48_AP1 <- reduce(motif_matches_48h$MA0099.3)
h48_AP1_TEAD_overlaps <- findOverlaps(h48_AP1,
                                      h48_TEAD, 
                                      ignore.strand = TRUE,
                                      maxgap = 50)

# find TEAD and AP1 motifs in simulated 48h data
motif_matches_48h_random <- motifmatchr::matchMotifs(pwms = TEAD_AP1_motifsToScan, 
                                                     subject = random_peaks_48h, 
                                                     genome = BSgenome.Hsapiens.UCSC.hg38, 
                                                     out = "positions")
h48_random_TEAD <- c(motif_matches_48h_random$MA0090.3, 
                     motif_matches_48h_random$MA0808.1,
                     motif_matches_48h_random$MA0809.2, 
                     motif_matches_48h_random$MA1121.1)
h48_random_TEAD <- reduce(h48_random_TEAD)
h48_random_AP1 <- reduce(motif_matches_48h_random$MA0099.3)
h48_random_AP1_TEAD_overlaps <- findOverlaps(h48_random_AP1, 
                                             h48_random_TEAD, 
                                             ignore.strand = TRUE,
                                             maxgap = 50)

summary_table_peaks <- data.frame(total_number_of_peaks = c(length(counts_consensus_filt_control_48h), 
                                                            length(random_peaks_control_48h),
                                                            length(counts_consensus_filt_48h),
                                                            length(random_peaks_48h)),
                                  TEAD_sites = c(length(control_48h_TEAD),
                                                 length(control_48h_random_TEAD),
                                                 length(h48_TEAD),
                                                 length(h48_random_TEAD)),
                                  AP1_sites = c(length(control_48h_AP1),
                                                length(control_48h_random_AP1),
                                                length(h48_AP1),
                                                length(h48_random_AP1)),
                                  TEAD_AP1_coloc_sites = c(length(control_48h_AP1_TEAD_overlaps),
                                                           length(control_48h_random_AP1_TEAD_overlaps),
                                                           length(h48_AP1_TEAD_overlaps),
                                                           length(h48_random_AP1_TEAD_overlaps)),
                                  row.names = c("control_48h", "Simulated control_48h", "h48", "Simulated h48")
)
summary_table_peaks


# Perform Fisher's exact tests for 48h treatment comparison
contingency_table <- matrix(c(length(h48_AP1_TEAD_overlaps), 
                              length(h48_TEAD) + length(h48_AP1) - length(h48_AP1_TEAD_overlaps), 
                              length(control_48h_AP1_TEAD_overlaps), 
                              length(control_48h_TEAD) + length(control_48h_AP1) - length(control_48h_AP1_TEAD_overlaps)),
                            nrow = 2, 
                            byrow = TRUE)
fisher_results <- fisher.test(contingency_table,
                              alternative = "two.sided")
fisher_summary_table <- data.table::data.table(comparison = "48h_vs_control_48h",
                                               pval = fisher_results$p.value,
                                               odds_ratio = fisher_results$estimate)

# Fisher test comparing real h48 AP1_TEAD peaks vs simulated
contingency_table_h48_vs_simulated <- matrix(c(length(h48_AP1_TEAD_overlaps), 
                                               length(h48_TEAD) + length(h48_AP1) - length(h48_AP1_TEAD_overlaps),
                                               length(h48_random_AP1_TEAD_overlaps), 
                                               length(h48_random_TEAD) + length(h48_random_AP1) - length(h48_random_AP1_TEAD_overlaps)), 
                                             nrow = 2,
                                             byrow = TRUE)
row.names(contingency_table_h48_vs_simulated) <- c("h48_peaks", "h48_peaks_random")
colnames(contingency_table_h48_vs_simulated) <- c("TEAD_AB1_colocalisation_present", "TEAD_AB1_colocalisation_absent")
contingency_table_h48_vs_simulated
fisher_results <- fisher.test(contingency_table_h48_vs_simulated,
                              alternative = "two.sided")
fisher_summary_table <- rbind(fisher_summary_table,
                              data.table::data.table(comparison = "h48_vs_h48sim",
                                                     pval = fisher_results$p.value,
                                                     odds_ratio = fisher_results$estimate))

# Fisher test for TEAD peaks
contingency_table_TEAD <- matrix(c(length(h48_TEAD), 
                                   length(counts_consensus_filt_48h) - length(h48_TEAD),
                                   length(control_48h_TEAD), 
                                   length(counts_consensus_filt_control_48h) - length(control_48h_TEAD)), 
                                 nrow = 2, byrow = TRUE)
row.names(contingency_table_TEAD) <- c("h48_peaks", "control_48h_peaks")
colnames(contingency_table_TEAD) <- c("TEAD_site_present", "TEAD_site_absent")
contingency_table_TEAD
fisher.test(contingency_table_TEAD, 
            alternative="two.sided")
chisq.test(contingency_table_TEAD)

# Fisher test for AP1 peaks
contingency_table_AP1 <- matrix(c(length(h48_AP1), 
                                  length(counts_consensus_filt_48h) - length(h48_AP1),
                                  length(control_48h_AP1), 
                                  length(counts_consensus_filt_control_48h) - length(control_48h_AP1)), 
                                nrow = 2,
                                byrow = TRUE)
row.names(contingency_table_AP1) <- c("h48_peaks", "control_48h_peaks")
colnames(contingency_table_AP1) <- c("AP1_site_present", "AP1_site_absent")
contingency_table_AP1
fisher.test(contingency_table_AP1, 
            alternative = "two.sided")
chisq.test(contingency_table_AP1)

```

```{r gene-set-enrichment-24h, error=TRUE}
# Gene set enrichment analysis for 24h treatment
# Test for MES/ADRN signature enrichment in TEAD/AP1 peaks

import::from(
  .from = "~/workspace/neuroblastoma/resources/utilityScripts.R",
  "chipEnrichAndExport"
)
locusdef <- "5kb"
our_rna_seq_terms <- path.expand("~/neuroblastoma/resources/mes_adrn_GS_frm_RNA_seq.tsv")
res_dir <- path.expand("~/neuroblastoma/results/ATAC-seq/ATAC-seq_drug_treatment/")

drg_trtm_ATAC
ATAC_dds_results_control <- ATAC_dds_results_24h$results_signif %>% 
  dplyr::filter(log2FoldChange < 0) %>% 
  dplyr::pull(PeakId)
h24_specific_peaks <- ATAC_dds_results_24h$results_signif %>% 
  dplyr::filter(log2FoldChange > 0) %>% 
  dplyr::pull(PeakId)
opts <- list()
opts[["tax_group"]] <- "vertebrates"
opts[["species"]] <- "9606"
opts[["collection"]] <- "CORE"
opts[["all_versions"]] <- FALSE
motifsToScan <- TFBSTools::getMatrixSet(JASPAR2022, opts)
TEAD_AP1_motifsToScan_names <- c("MA0090.3", 
                                 "MA0808.1",
                                 "MA0809.2",
                                 "MA1121.1",
                                 "MA0099.3")
TEAD_AP1_motifsToScan <- motifsToScan[TEAD_AP1_motifsToScan_names,]

# Using different subsets of peaks to check 
drg_trtm_ATAC
counts_consensus_filt_control <- drg_trtm_ATAC[drg_trtm_ATAC@rowRanges@elementMetadata@listData[["PeakId"]] %in% ATAC_dds_results_control, ]
counts_consensus_filt_24h <- drg_trtm_ATAC[drg_trtm_ATAC@rowRanges@elementMetadata@listData[["PeakId"]] %in% h24_specific_peaks, ]

motif_matches_control <- motifmatchr::matchMotifs(pwms = TEAD_AP1_motifsToScan, 
                                                  subject = counts_consensus_filt_control, 
                                                  genome = BSgenome.Hsapiens.UCSC.hg38, 
                                                  out = "positions")
control_TEAD <- c(motif_matches_control$MA0090.3,
                  motif_matches_control$MA0808.1,
                  motif_matches_control$MA0809.2,
                  motif_matches_control$MA1121.1)
control_TEAD <- reduce(control_TEAD)
control_AP1 <- reduce(motif_matches_control$MA0099.3)


control_AP1_TEAD_overlaps <- ChIPpeakAnno::findOverlapsOfPeaks(control_AP1,
                                                               control_TEAD,
                                                               maxgap = 50)
control_AP1_TEAD_overlaps <- control_AP1_TEAD_overlaps$peaksInMergedPeaks
control_AP1_TEAD_overlaps <- as.data.frame(control_AP1_TEAD_overlaps)

if(nrow(control_AP1_TEAD_overlaps) != 0){
  chipEnrichAndExport(peaks = control_AP1_TEAD_overlaps,
                      peaksName = "control", 
                      TF_name = "control_AP1_TEAD", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}

if(nrow(as.data.frame(control_AP1)) != 0){
  chipEnrichAndExport(peaks = as.data.frame(control_AP1),
                      peaksName = "control", 
                      TF_name = "control_AP1", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}

if(nrow(as.data.frame(control_TEAD)) != 0){
  chipEnrichAndExport(peaks = as.data.frame(control_TEAD),
                      peaksName = "control", 
                      TF_name = "control_TEAD", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}

control <- reduce(counts_consensus_filt_control@rowRanges)
if(nrow(as.data.frame(control)) != 0){
  chipEnrichAndExport(peaks = as.data.frame(control),
                      peaksName = "control", 
                      TF_name = "control_Total", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}

h24_AP1_TEAD_overlaps <- ChIPpeakAnno::findOverlapsOfPeaks(h24_AP1, h24_TEAD, maxgap = 50)
h24_AP1_TEAD_overlaps <- h24_AP1_TEAD_overlaps$peaksInMergedPeaks
h24_AP1_TEAD_overlaps <- as.data.frame(h24_AP1_TEAD_overlaps)

if(nrow(as.data.frame(h24_AP1_TEAD_overlaps)) != 0){
  chipEnrichAndExport(peaks = h24_AP1_TEAD_overlaps,
                      peaksName = "h24", 
                      TF_name = "h24_AP1_TEAD", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}

#the intersection here doesn't exist. The function returns error. This is why it's commented out.
# if(nrow(as.data.frame(h24_AP1)) != 0){
#   chipEnrichAndExport(peaks = as.data.frame(h24_AP1),
#                       peaksName = "h24", 
#                       TF_name = "h24_AP1", 
#                       res_dir = res_dir, 
#                       genesets = our_rna_seq_terms, 
#                       genesets_name = "Our_data_",
#                       locusdef = locusdef
#   )
# }

if(nrow(as.data.frame(h24_TEAD)) != 0){
  chipEnrichAndExport(peaks = as.data.frame(h24_TEAD),
                      peaksName = "h24", 
                      TF_name = "h24_TEAD", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}

h24 <- reduce(counts_consensus_filt_24h@rowRanges)
if(nrow(as.data.frame(h24)) != 0){
  chipEnrichAndExport(peaks = as.data.frame(h24),
                      peaksName = "h24", 
                      TF_name = "h24_Total", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}

enrich_results_files <- unlist(list.files(path = res_dir, 
                                          pattern = "^Enricher_.*(h24|control).*.xlsx", 
                                          full.names = TRUE))
summary_table <- data.frame(Protein = NULL,
                            Data_source = NULL,
                            Selected_peaks = NULL,
                            Description = NULL,
                            P.value = NULL,
                            FDR = NULL,
                            Effect = NULL,
                            Status = NULL,
                            Gene_set_size = NULL,
                            Peaks_in_set = NULL,
                            Odds_ratio = NULL)

pattern <-  ".*(control_Total|control_AP1|control_TEAD|control_AP1_TEAD|h24_Total|h24_AP1|h24_TEAD|h24_AP1_TEAD)_(Our_data|Groen).*(control|h24).*"

for (file_name in enrich_results_files){
  tmp <- openxlsx2::wb_to_df(file_name, sheet = 1)
  
  basename_tmp <- basename(file_name)
  basename_tmp <- str_extract(basename_tmp, 
                              pattern, 
                              group = c(1,2,3))
  
  summary_table_tmp <- data.frame(Protein = basename_tmp[1],
                                  Data_source = basename_tmp[2],
                                  Selected_peaks = basename_tmp[3],
                                  Description = tmp$Description,
                                  P.value = tmp$P.value,
                                  FDR = tmp$FDR,
                                  Effect = tmp$Effect,
                                  Status = tmp$Status,
                                  Gene_set_size = tmp$N.Geneset.Genes,
                                  Peaks_in_set = tmp$N.Geneset.Peak.Genes,
                                  Odds_ratio = tmp$Odds.Ratio)
  
  summary_table <- rbind(summary_table,
                         summary_table_tmp)
  
}
# Summary: No significant enrichment detected in 24h treatment analysis
summary_table
```

```{r gene-set-enrichment-48h, error=TRUE}
# Gene set enrichment analysis for 48h treatment
# Repeat enrichment analysis for longer treatment timepoint
import::from(
  .from = "~/workspace/neuroblastoma/resources/utilityScripts.R",
  "chipEnrichAndExport"
)
locusdef <- "5kb"
our_rna_seq_terms <- path.expand("~/neuroblastoma/resources/mes_adrn_GS_frm_RNA_seq.tsv")
res_dir <- path.expand("~/neuroblastoma/results/ATAC-seq/ATAC-seq_drug_treatment/")

drg_trtm_ATAC
ATAC_dds_results_control <- ATAC_dds_results_48h$results_signif %>% 
  dplyr::filter(log2FoldChange < 0) %>% 
  dplyr::pull(PeakId)
h48_specific_peaks <- ATAC_dds_results_48h$results_signif %>% 
  dplyr::filter(log2FoldChange > 0) %>% 
  dplyr::pull(PeakId)
opts <- list()
opts[["tax_group"]] <- "vertebrates"
opts[["species"]] <- "9606"
opts[["collection"]] <- "CORE"
opts[["all_versions"]] <- FALSE
motifsToScan <- TFBSTools::getMatrixSet(JASPAR2022, opts)
TEAD_AP1_motifsToScan_names <- c("MA0090.3", 
                                 "MA0808.1",
                                 "MA0809.2",
                                 "MA1121.1",
                                 "MA0099.3")
TEAD_AP1_motifsToScan <- motifsToScan[TEAD_AP1_motifsToScan_names,]

# Using different subsets of peaks to check 
drg_trtm_ATAC
counts_consensus_filt_control <- drg_trtm_ATAC[drg_trtm_ATAC@rowRanges@elementMetadata@listData[["PeakId"]] %in% ATAC_dds_results_control, ]
counts_consensus_filt_48h <- drg_trtm_ATAC[drg_trtm_ATAC@rowRanges@elementMetadata@listData[["PeakId"]] %in% h48_specific_peaks, ]

motif_matches_control <- motifmatchr::matchMotifs(pwms = TEAD_AP1_motifsToScan, 
                                                  subject = counts_consensus_filt_control, 
                                                  genome = BSgenome.Hsapiens.UCSC.hg38, 
                                                  out = "positions")
control_TEAD <- c(motif_matches_control$MA0090.3,
                  motif_matches_control$MA0808.1,
                  motif_matches_control$MA0809.2,
                  motif_matches_control$MA1121.1)
control_TEAD <- reduce(control_TEAD)
control_AP1 <- reduce(motif_matches_control$MA0099.3)

control_AP1_TEAD_overlaps <- ChIPpeakAnno::findOverlapsOfPeaks(control_AP1,
                                                               control_TEAD,
                                                               maxgap = 50)
control_AP1_TEAD_overlaps <- control_AP1_TEAD_overlaps$peaksInMergedPeaks
control_AP1_TEAD_overlaps <- as.data.frame(control_AP1_TEAD_overlaps)

if(nrow(control_AP1_TEAD_overlaps) != 0){
  chipEnrichAndExport(peaks = control_AP1_TEAD_overlaps,
                      peaksName = "control", 
                      TF_name = "control_AP1_TEAD", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}

if(nrow(as.data.frame(control_AP1)) != 0){
  chipEnrichAndExport(peaks = as.data.frame(control_AP1),
                      peaksName = "control", 
                      TF_name = "control_AP1", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}

if(nrow(as.data.frame(control_TEAD)) != 0){
  chipEnrichAndExport(peaks = as.data.frame(control_TEAD),
                      peaksName = "control", 
                      TF_name = "control_TEAD", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}

control <- reduce(counts_consensus_filt_control@rowRanges)
if(nrow(as.data.frame(control)) != 0){
  chipEnrichAndExport(peaks = as.data.frame(control),
                      peaksName = "control", 
                      TF_name = "control_Total", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}


# Analysing 48h
h48_AP1_TEAD_overlaps <- ChIPpeakAnno::findOverlapsOfPeaks(h48_AP1,
                                                           h48_TEAD,
                                                           maxgap = 50)
h48_AP1_TEAD_overlaps <- h48_AP1_TEAD_overlaps$peaksInMergedPeaks
h48_AP1_TEAD_overlaps <- as.data.frame(h48_AP1_TEAD_overlaps)

# There are no overlaps between peaks, function returns error. The code is commented out.
# if(nrow(as.data.frame(h48_AP1_TEAD_overlaps)) != 0){
#   chipEnrichAndExport(peaks = h48_AP1_TEAD_overlaps,
#                       peaksName = "h48", 
#                       TF_name = "h48_AP1_TEAD", 
#                       res_dir = res_dir, 
#                       genesets = our_rna_seq_terms, 
#                       genesets_name = "Our_data_",
#                       locusdef = locusdef
#   )
# }

if(nrow(as.data.frame(h48_AP1)) != 0){
  chipEnrichAndExport(peaks = as.data.frame(h48_AP1),
                      peaksName = "h48", 
                      TF_name = "h48_AP1", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}

if(nrow(as.data.frame(h48_TEAD)) != 0){
  chipEnrichAndExport(peaks = as.data.frame(h48_TEAD),
                      peaksName = "h48", 
                      TF_name = "h48_TEAD", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}

h48 <- reduce(counts_consensus_filt_48h@rowRanges)
if(nrow(as.data.frame(h48)) != 0){
  chipEnrichAndExport(peaks = as.data.frame(h48),
                      peaksName = "h48", 
                      TF_name = "h48_Total", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}

enrich_results_files <- unlist(list.files(path = res_dir,
                                          pattern = "^Enricher_.*(h48|control).*.xlsx",
                                          full.names = TRUE))
summary_table <- data.frame(Protein = NULL,
                            Data_source = NULL,
                            Selected_peaks = NULL,
                            Description = NULL,
                            P.value = NULL,
                            FDR = NULL,
                            Effect = NULL,
                            Status = NULL,
                            Gene_set_size = NULL,
                            Peaks_in_set = NULL,
                            Odds_ratio = NULL)

pattern <-  ".*(control_Total|control_AP1|control_TEAD|control_AP1_TEAD|h48_Total|h48_AP1|h48_TEAD|h48_AP1_TEAD)_(Our_data|Groen).*(control|h48).*"

for (file_name in enrich_results_files){
  tmp <- openxlsx2::wb_to_df(file_name, sheet = 1)
  
  basename_tmp <- basename(file_name)
  basename_tmp <- str_extract(basename_tmp, 
                              pattern, 
                              group = c(1,2,3))
  
  summary_table_tmp <- data.frame(Protein = basename_tmp[1],
                                  Data_source = basename_tmp[2],
                                  Selected_peaks = basename_tmp[3],
                                  Description = tmp$Description,
                                  P.value = tmp$P.value,
                                  FDR = tmp$FDR,
                                  Effect = tmp$Effect,
                                  Status = tmp$Status,
                                  Gene_set_size = tmp$N.Geneset.Genes,
                                  Peaks_in_set = tmp$N.Geneset.Peak.Genes,
                                  Odds_ratio = tmp$Odds.Ratio)
  
  summary_table <- rbind(summary_table,
                         summary_table_tmp)
}
summary_table
# Summary: Only control condition shows significant enrichment in 48h analysis
```
