---
title: "ATAC-seq Overexpression of WWTR1 analysis "
author: "Aleksandr Bykov (Enhanced by Claude Code)"
date: '`r Sys.Date()`'
output:
  html_document: 
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 8,
  dpi = 300
)
```

```{r setup-libraries}
# OE ATAC-seq analysis
## Loading libraries ####
library(dplyr)
library(tidyr)
library(DESeq2)
library(ggplot2)
library(BSgenome.Hsapiens.UCSC.hg38)
library(IRanges)
library(patchwork)
library(stringr)
import::from(stringr, str_extract, str_detect)
import::from(GenomicRanges,GRanges)
import::from(rtracklayer, liftOver, import.chain)
import::from(openxlsx, addWorksheet, writeData, saveWorkbook)
import::from(openxlsx2, read_xlsx)
import::from(.from = TFBSTools, getMatrixSet)
import::from(chromVAR, addGCBias)
import::from(.from = RColorBrewer, brewer.pal)
import::from(.from = "~/neuroblastoma/resources/utilityScripts.R",
             "generatePCA", 
             "extract_results_nextflow_output",
             "meanExprsPerGroup_nextflow",
             "extract_results_DDS_HIC",
             "plotVolcano",
             "get_the_poissoon_p_val")
import::from(
  .from = here::here("~/neuroblastoma/resources/UtilityScriptsRNA-seq.R"),
  "filterDatasets",
  .character_only = TRUE
)
```

```{r setup-environment}
# BugFix - JASPAR2022 cannot be pulled from the web automatically, so it need to be loaded manually
download.file(url = "https://jaspar2022.genereg.net/download/database/JASPAR2022.sqlite",
              destfile = "~/neuroblastoma/resources/JASPAR2022.sqlite")
JASPAR2022 <-  "~/neuroblastoma/resources/JASPAR2022.sqlite"

# Set up the environment ####
deg_dir <- "~/neuroblastoma/results/ATAC-seq/ATAC-seq_OE/"
if(!dir.exists(deg_dir)){dir.create(deg_dir)}
```

```{r load-atac-data}
# Analysis part ###########################################################################
# Load and prepare ATAC-seq data
# We use only filtered consensus peaks. And combined annotation from genecode, homer and reg.
tmp_dir <- "~/neuroblastoma/data/ATACseq/"
path_to_the_ATAC_counts_data <- file.path(tmp_dir, "consensus_peaks.mLb.clN.rds")
path_annnotation_ATAC_data <- file.path(tmp_dir, "consensus_peaks.mLb.clN.annotatePeaks.txt")
ATAC_counts_data <- readRDS(path_to_the_ATAC_counts_data)
ATAC_annotation_data <- read.csv(path_annnotation_ATAC_data, sep = "\t")

colnames(ATAC_annotation_data)[1] <- "PeakId"
ATAC_annotation_data <- ATAC_annotation_data %>%
  mutate(PeakId = factor(PeakId, levels = row.names(ATAC_counts_data))) %>%
  arrange(PeakId)

#remove Scaffolds chrs
ATAC_annotation_data <- ATAC_annotation_data[ATAC_annotation_data$Chr %in% c(1:22, "X", "Y"), ]
ATAC_counts_data <- ATAC_counts_data[rownames(ATAC_counts_data) %in% ATAC_annotation_data$PeakId, ]

#add ranges
rowRanges(ATAC_counts_data) <- GenomicRanges::GRanges(seqnames = paste0("chr", ATAC_annotation_data$Chr), 
                                              ranges = IRanges::IRanges(start = ATAC_annotation_data$Start, 
                                                                        end = ATAC_annotation_data$End
                                              )
)
mcols(ATAC_counts_data) <- DataFrame(mcols(ATAC_counts_data), ATAC_annotation_data)
```


```{r subset-overexpression-samples}
# Subset to overexpression samples and control knockdown samples
# Extract sample information and create metadata for differential analysis
samples_to_subset <- c(
  "SH_A_ATAC_S131296_REP1",
  "SK_N_SH_ADR_rep1_REP1",
  "SK_N_SH_ADR_rep2_REP1",
  "SH_M_ATAC_S131295_REP1",
  "SK_N_SH_MES_rep1_REP1",
  "SK_N_SH_MES_rep2_REP1",
  
  "OE_WWTR1_DMSO_ATAC_S165172_REP1",
  "OE_WWTR1_DMSO_rep1_REP1",
  "OE_WWTR1_DMSO_rep3_REP1",
  
  "OE_WWTR1_dox_TDI_ATAC_S165175_REP1",
  "OE_WWTR1_dox_TDI_rep1_REP1",
  "OE_WWTR1_dox_TDI_rep3_REP1",
  
  "OE_WWTR1_dox_ATAC_S165177_REP1",
  "OE_WWTR1_dox_rep1_REP1",
  "OE_WWTR1_dox_rep_3_REP1",
  
  "OE_WWTR1_TDI_ATAC_S165176_REP1",
  "OE_WWTR1_TDI_rep1_REP1",
  "OE_WWTR1_TDI_rep3_REP1"
) 
  
OE_ATAC <- ATAC_counts_data[,samples_to_subset]

OE_ATAC$experiment <- as.factor(case_when(
  OE_ATAC$sample %in%  c(
    "SH_A_ATAC_S131296_REP1",
    "SK_N_SH_ADR_rep1_REP1",
    "SK_N_SH_ADR_rep2_REP1",
    "SH_M_ATAC_S131295_REP1",
    "SK_N_SH_MES_rep1_REP1",
    "SK_N_SH_MES_rep2_REP1") ~ "old",
  OE_ATAC$sample %in%  c(
    "OE_WWTR1_DMSO_ATAC_S165172_REP1",
    "OE_WWTR1_DMSO_rep1_REP1",
    "OE_WWTR1_DMSO_rep3_REP1",
    "OE_WWTR1_dox_TDI_ATAC_S165175_REP1",
    "OE_WWTR1_dox_TDI_rep1_REP1",
    "OE_WWTR1_dox_TDI_rep3_REP1",
    "OE_WWTR1_dox_ATAC_S165177_REP1",
    "OE_WWTR1_dox_rep1_REP1",
    "OE_WWTR1_dox_rep_3_REP1",
    "OE_WWTR1_TDI_ATAC_S165176_REP1",
    "OE_WWTR1_TDI_rep1_REP1",
    "OE_WWTR1_TDI_rep3_REP1") ~ "new",
))
  
OE_ATAC$genotype <- as.factor(case_when(
  OE_ATAC$sample %in%  c(
    "SH_A_ATAC_S131296_REP1",
    "SK_N_SH_ADR_rep1_REP1",
    "SK_N_SH_ADR_rep2_REP1") ~ "SH_ADR",
  OE_ATAC$sample %in%  c(
    "OE_WWTR1_DMSO_ATAC_S165172_REP1",
    "OE_WWTR1_DMSO_rep1_REP1",
    "OE_WWTR1_DMSO_rep3_REP1",
    "OE_WWTR1_dox_TDI_ATAC_S165175_REP1",
    "OE_WWTR1_dox_TDI_rep1_REP1",
    "OE_WWTR1_dox_TDI_rep3_REP1",
    "OE_WWTR1_dox_ATAC_S165177_REP1",
    "OE_WWTR1_dox_rep1_REP1",
    "OE_WWTR1_dox_rep_3_REP1",
    "OE_WWTR1_TDI_ATAC_S165176_REP1",
    "OE_WWTR1_TDI_rep1_REP1",
    "OE_WWTR1_TDI_rep3_REP1") ~ "WWTR1",
  OE_ATAC$sample %in%  c(
    "SH_M_ATAC_S131295_REP1",
    "SK_N_SH_MES_rep1_REP1",
    "SK_N_SH_MES_rep2_REP1") ~ "SH_M",
))

OE_ATAC$genotype <- as.factor(case_when(
  OE_ATAC$sample %in%  c(
    "SH_A_ATAC_S131296_REP1",
    "SK_N_SH_ADR_rep1_REP1",
    "SK_N_SH_ADR_rep2_REP1") ~ "SH_ADR",
  OE_ATAC$sample %in%  c(
    "OE_WWTR1_DMSO_ATAC_S165172_REP1",
    "OE_WWTR1_DMSO_rep1_REP1",
    "OE_WWTR1_DMSO_rep3_REP1",
    "OE_WWTR1_dox_TDI_ATAC_S165175_REP1",
    "OE_WWTR1_dox_TDI_rep1_REP1",
    "OE_WWTR1_dox_TDI_rep3_REP1",
    "OE_WWTR1_dox_ATAC_S165177_REP1",
    "OE_WWTR1_dox_rep1_REP1",
    "OE_WWTR1_dox_rep_3_REP1",
    "OE_WWTR1_TDI_ATAC_S165176_REP1",
    "OE_WWTR1_TDI_rep1_REP1",
    "OE_WWTR1_TDI_rep3_REP1") ~ "WWTR1",
  OE_ATAC$sample %in%  c(
    "SH_M_ATAC_S131295_REP1",
    "SK_N_SH_MES_rep1_REP1",
    "SK_N_SH_MES_rep2_REP1") ~ "SH_M",
))

OE_ATAC$treatment <- as.factor(case_when(
  OE_ATAC$sample %in%  c(
    "SH_A_ATAC_S131296_REP1",
    "SK_N_SH_ADR_rep1_REP1",
    "SK_N_SH_ADR_rep2_REP1",
    "SH_M_ATAC_S131295_REP1",
    "SK_N_SH_MES_rep1_REP1",
    "SK_N_SH_MES_rep2_REP1"
    ) ~ "NO",
  OE_ATAC$sample %in%  c(
    "OE_WWTR1_DMSO_ATAC_S165172_REP1",
    "OE_WWTR1_DMSO_rep1_REP1",
    "OE_WWTR1_DMSO_rep3_REP1") ~ "DMSO",
  OE_ATAC$sample %in%  c(
    "OE_WWTR1_dox_TDI_ATAC_S165175_REP1",
    "OE_WWTR1_dox_TDI_rep1_REP1",
    "OE_WWTR1_dox_TDI_rep3_REP1") ~ "dox_TDI",
  OE_ATAC$sample %in%  c(
    "OE_WWTR1_dox_ATAC_S165177_REP1",
    "OE_WWTR1_dox_rep1_REP1",
    "OE_WWTR1_dox_rep_3_REP1") ~ "dox",
  OE_ATAC$sample %in%  c(
    "OE_WWTR1_TDI_ATAC_S165176_REP1",
    "OE_WWTR1_TDI_rep1_REP1",
    "OE_WWTR1_TDI_rep3_REP1"
    ) ~ "TDI"
))

design(OE_ATAC) <- as.formula("~ treatment")

#OE_ATAC@rowRanges@elementMetadata@listData[["PeakId"]]

RNA_SEQ_data <- openxlsx2::read_xlsx("~/neuroblastoma/results/RNA-seq/RNA-seq_MES_vs_ADR/tables/MES_vs_ADR_comprehensive_results.xlsx", sheet = 2)
RNA_SEQ_data <- RNA_SEQ_data %>%
  dplyr::mutate(Term = if_else(log2FoldChange < 0, "ADRN", "MES")) %>%
  dplyr::select(gene_symbol, Term)
mcols(OE_ATAC)$gene_category_our_RNAseq <- "ND"
mcols(OE_ATAC)$gene_category_our_RNAseq <- case_when(
  mcols(OE_ATAC)$Gene.Name %in% RNA_SEQ_data[RNA_SEQ_data$Term == "ADRN", 1] ~ "g_ADRN",
  mcols(OE_ATAC)$Gene.Name %in% RNA_SEQ_data[RNA_SEQ_data$Term == "MES", 1] ~ "g_MES",
  TRUE ~ mcols(OE_ATAC)$gene_category_our_RNAseq
)

# Add metadata - marks peaks that are around 3kb from TSS
mcols(OE_ATAC)$is_promoter_3kb <- FALSE
mcols(OE_ATAC)$is_promoter_3kb[abs(mcols(OE_ATAC)$Distance.to.TSS) <= 3000] <- TRUE

OE_ATAC <- filterDatasets(OE_ATAC,
                          abs_filt = TRUE,
                          abs_filt_samples = 2)

OE_ATAC <- estimateSizeFactors(OE_ATAC)
OE_ATAC <- DESeq(OE_ATAC)
vsd <- vst(OE_ATAC, blind = F)
```


```{r pca-analysis}
# Perform differential accessibility analysis and generate PCA plots
# First filter low-count peaks, then run DESeq2 normalization and transformation
# remove replicate 2
pca_dar <- generatePCA(transf_object = vsd, 
                       cond_interest_varPart = c("treatment", "genotype"), 
                       color_variable = "treatment", 
                       shape_variable = "genotype",
                       ntop_genes = 1000) +
  ggtitle("OE data + old ADNR and MES samples") 
pca_dar

# Batch correction
transf_batch_NObatch_experiment <- vsd
transf_batch_NObatch_experiment_count <- limma::removeBatchEffect(
  SummarizedExperiment::assay(transf_batch_NObatch_experiment),
  transf_batch_NObatch_experiment$experiment)
SummarizedExperiment::assay(transf_batch_NObatch_experiment) <- transf_batch_NObatch_experiment_count
pca_deg_NObatch_experiment <- generatePCA(transf_object = transf_batch_NObatch_experiment, 
                                          cond_interest_varPart = c("treatment", "genotype"), 
                                          color_variable = "treatment", 
                                          shape_variable = "genotype",
                                          ntop_genes = 1000) +
  ggtitle("OE data + old ADNR and MES samples batch corrected") 
pca_deg_NObatch_experiment

```


```{r chromvar-motif-analysis}
# ChromVAR motif enrichment analysis
# Identify transcription factor motifs that are differentially accessible
# between overexpression conditions
opts <- list()
opts[["tax_group"]] <- "vertebrates"
opts[["species"]] <- "9606"
opts[["collection"]] <- "CORE"
opts[["all_versions"]] <- FALSE
motifsToScan <- TFBSTools::getMatrixSet(JASPAR2022, opts)

# Filter peaks with low read counts across samples
# Using different subsets of peaks to check 
sbst <-  c("ALL")
# here we check if there are any peaks that have less than 5 reads across all samples.
dim(OE_ATAC)
counts_consensus_filt <- OE_ATAC[rowSums(assay(OE_ATAC)) > 5, ]
dim(counts_consensus_filt)

print(paste("procesing: ", sbst ))

# ChromVar package uses GC content to identify background peaks that are likely to be non-functional and exclude them from further analysis. 
counts_consensus_filt <- chromVAR::addGCBias(counts_consensus_filt,
                                             genome = BSgenome.Hsapiens.UCSC.hg38) 

# Having corrected for bias, we can use the matchMotifs function to identify motifs under our ATACseq peaks.
motif_matches <- motifmatchr::matchMotifs(pwms = motifsToScan, 
                                          subject = counts_consensus_filt, 
                                          genome = BSgenome.Hsapiens.UCSC.hg38, 
                                          out = "matches")
motif_matches

BiocParallel::register(BiocParallel::MulticoreParam(8, progressbar = FALSE))
set.seed(42069)

# Run chromVar ----
#Following the identification of motifs in our peaks, we can perform the summarization of ATACseq signal to motifs using the computeDeviations and the computeVariability functions.
background_peaks <- chromVAR::getBackgroundPeaks(object = counts_consensus_filt) 
access_expectation <- chromVAR::computeExpectations(object = counts_consensus_filt)
chrom_access_deviations <- chromVAR::computeDeviations(object = counts_consensus_filt, 
                                                       annotations = motif_matches,
                                                       background_peaks = background_peaks,
                                                       expectation = access_expectation)

# Generate sample correlation heatmap to assess data quality
sample_cor <- chromVAR::getSampleCorrelation(chrom_access_deviations)
annotation_row_sampleCor <- as.data.frame(SummarizedExperiment::colData(chrom_access_deviations)) %>%
  dplyr::select(sample)
pheatmap::pheatmap(as.dist(sample_cor), 
                   annotation_row = annotation_row_sampleCor, 
                   clustering_distance_rows = as.dist(1-sample_cor), 
                   clustering_distance_cols = as.dist(1-sample_cor),  
                   annotation_names_row = TRUE, annotation_names_col = TRUE) 

# Perform t-SNE dimensionality reduction for sample clustering visualization
tsne_results <- chromVAR::deviationsTsne(chrom_access_deviations, 
                                         threshold = 1.5, 
                                         perplexity = 8, 
                                         what = "samples", 
                                         shiny = FALSE)

tsne_plots <- chromVAR::plotDeviationsTsne(chrom_access_deviations, 
                                           tsne_results,
                                           sample_column = "sample", 
                                           shiny = FALSE)
tsne_plots

diff_acc <- chromVAR::differentialDeviations(chrom_access_deviations, 
                                             groups="treatment", 
                                             parametric = FALSE)
motif_annot <- as.data.frame(SummarizedExperiment::rowData(chrom_access_deviations)) %>%
  tibble::rownames_to_column(var = "motif") %>%
  dplyr::select(motif, name)
diff_acc_annot <- diff_acc %>%
  tibble::rownames_to_column(var = "motif") %>%
  dplyr::left_join(., motif_annot, by = "motif")
head(diff_acc_annot) 

diff_var <- chromVAR::differentialVariability(chrom_access_deviations,
                                              "treatment",
                                              parametric = FALSE)
diff_var_annot <- diff_var %>%
  tibble::rownames_to_column(var = "motif") %>%
  dplyr::left_join(., motif_annot, by = "motif")
head(diff_var_annot)

devZscores <- chromVAR::deviationScores(chrom_access_deviations)
devZscores_df <- tibble::rownames_to_column(as.data.frame(devZscores), var="jaspar_id")

#The function computeVariability returns a data.frame that contains the variability (standard deviation of the z scores computed above
# across all cell/samples for a set of peaks), bootstrap confidence intervals for that variability (by resampling cells/samples), 
# and a p-value for the variability being greater than the null hypothesis of 1.
chrom_access_variability <- chromVAR::computeVariability(chrom_access_deviations)
chrom_access_variability_plot <- chromVAR::plotVariability(chrom_access_variability,
                                                           use_plotly = FALSE)
chrom_access_variability_plot

ggsave(chrom_access_variability_plot, 
       filename = file.path(deg_dir, paste0("ChromVAR_chrom_access_variability_corr_plot_", sbst, ".pdf")))

BiocParallel::register(BiocParallel::SerialParam())

chrom_access_variability_ord <- chrom_access_variability[order(chrom_access_variability$p_value), ]  # ordering based on p-value

#! to store
chrom_access_variability_ord_results <- chrom_access_variability_ord %>%
  tibble::rownames_to_column(var="jaspar_id") %>%
  dplyr::left_join(., devZscores_df, by = "jaspar_id")

message("Saving motif enrichment results")
ntop <- 100
topVariable <- chrom_access_variability_ord[1:ntop, ]
topVariable <- tibble::rownames_to_column(topVariable, var="jaspar_id")

devTop <- topVariable %>%
  dplyr::left_join(., devZscores_df, by = "jaspar_id")

devTop <- devTop %>% mutate(long_name = paste0(jaspar_id, "_", name))

devToPlot <- devTop %>%
  dplyr::select(long_name, 
                SH_A_ATAC_S131296_REP1,
                SK_N_SH_ADR_rep1_REP1,
                SK_N_SH_ADR_rep2_REP1,
                SH_M_ATAC_S131295_REP1,
                SK_N_SH_MES_rep1_REP1,
                SK_N_SH_MES_rep2_REP1,
                OE_WWTR1_DMSO_ATAC_S165172_REP1,
                OE_WWTR1_DMSO_rep1_REP1,
                OE_WWTR1_DMSO_rep3_REP1,
                OE_WWTR1_dox_TDI_ATAC_S165175_REP1,
                OE_WWTR1_dox_TDI_rep1_REP1,
                OE_WWTR1_dox_TDI_rep3_REP1,
                OE_WWTR1_dox_ATAC_S165177_REP1,
                OE_WWTR1_dox_rep1_REP1,
                OE_WWTR1_dox_rep_3_REP1,
                OE_WWTR1_TDI_ATAC_S165176_REP1,
                OE_WWTR1_TDI_rep1_REP1,
                OE_WWTR1_TDI_rep3_REP1
                ) %>%
  tibble::column_to_rownames(var = "long_name")

annotCol_forHeatmap <- as.data.frame(colData(OE_ATAC)) %>%
  dplyr::select(treatment)
# annotCol_forHeatmap_colors <- list(phenotype = c(A = "#525252", M = "#fc4e2a"))
annotCol_forHeatmap_colors <- list(phenotype = c(dox = "#525252", 
                                                 TDI = "#fc4e2a", 
                                                 DMSO = "green", 
                                                 dox_TDI = "yellow", 
                                                 NO = "pink"))

color.scheme <- colorRampPalette(c("navy", "white", "firebrick3"))(50)

heatmap <- pheatmap::pheatmap(as.matrix(devToPlot),
                              color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
                              border_color = NA,
                              #color = colorRampPalette(rev(brewer.pal(7, "RdBu")))(100),
                              scale = "row",
                              cluster_cols = FALSE,
                              cluster_rows = TRUE,
                              show_colnames = TRUE,
                              show_rownames = TRUE,
                              annotation_col = annotCol_forHeatmap,
                              annotation_colors = annotCol_forHeatmap_colors,  
                              fontsize_row = 9,
                              fontsize = 9,
                              main = paste0("Heatmap of signif. ATAC-seq OE DMSO TDI dox TDI_DOX ", sbst, "- used"))

ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_drug_OE_DMSO_TDI_dox_TDI_dox_Motifs", sbst, "- used", ".png")), 
       plot=heatmap,
       width = 18, height = 38, units = "cm")
ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_drug_OE_DMSO_TDI_dox_TDI_dox_Motifs", sbst, "- used", ".pdf")), 
       plot=heatmap,
       width = 18, height = 38, units = "cm")

# Generate heatmap without control samples (SH_M and SH_A) to focus on overexpression effects
devToPlot <- devTop %>%
  dplyr::select(long_name, 
                OE_WWTR1_DMSO_ATAC_S165172_REP1,
                OE_WWTR1_DMSO_rep1_REP1,
                OE_WWTR1_DMSO_rep3_REP1,
                OE_WWTR1_dox_TDI_ATAC_S165175_REP1,
                OE_WWTR1_dox_TDI_rep1_REP1,
                OE_WWTR1_dox_TDI_rep3_REP1,
                OE_WWTR1_dox_ATAC_S165177_REP1,
                OE_WWTR1_dox_rep1_REP1,
                OE_WWTR1_dox_rep_3_REP1,
                OE_WWTR1_TDI_ATAC_S165176_REP1,
                OE_WWTR1_TDI_rep1_REP1,
                OE_WWTR1_TDI_rep3_REP1
  ) %>%
  tibble::column_to_rownames(var = "long_name")

heatmap <- pheatmap::pheatmap(as.matrix(devToPlot),
                              color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
                              border_color = NA,
                              #color = colorRampPalette(rev(brewer.pal(7, "RdBu")))(100),
                              scale = "row",
                              cluster_cols = FALSE,
                              cluster_rows = TRUE,
                              show_colnames = TRUE,
                              show_rownames = TRUE,
                              annotation_col = annotCol_forHeatmap,
                              annotation_colors = annotCol_forHeatmap_colors,  
                              fontsize_row = 9,
                              fontsize = 9,
                              main = paste0(" Heatmap of signif. ATAC-seq OE DMSO TDI dox TDI_DOX \n no SH_A SH_M ", sbst, "- used"))

ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_drug_OE_DMSO_TDI_dox_TDI_dox_Motifs_no_SHA_SHM", sbst, "- used", ".png")), 
       plot=heatmap,
       width = 18, height = 38, units = "cm")
ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_drug_OE_DMSO_TDI_dox_TDI_dox_Motifs_no_SHA_SHM", sbst, "- used", ".pdf")),
       plot=heatmap,
       width = 18, height = 38, units = "cm")

```

The heatmaps above show every replicate as its own column. To see the overall
signal per condition without replicate-to-replicate noise, we collapse all
replicates of a condition into a single averaged column. Grouping is
genotype x treatment, which keeps the two untreated controls (SH_ADR and SH_M,
both treatment == "NO") as separate columns.
```{r motif-heatmap-averaged}
# Averaged motif heatmap: one mean deviation z-score column per condition group
# instead of one column per replicate.

# Per-sample metadata. rownames(colData) == column names of the z-score matrix,
# so `sample_id` lines up with the per-sample columns carried in devTop.
sample_meta <- as.data.frame(colData(OE_ATAC)) %>%
  tibble::rownames_to_column("sample_id") %>%
  dplyr::mutate(group = paste(genotype, treatment, sep = "_")) %>%
  dplyr::select(sample_id, group, genotype, treatment)

# Long form -> mean per (motif, group) -> wide matrix (motifs x condition groups).
# Averaging on the raw deviation z-scores; pheatmap still row-scales for display.
devToPlot_mean <- devTop %>%
  dplyr::select(long_name, dplyr::all_of(sample_meta$sample_id)) %>%
  tidyr::pivot_longer(-long_name, names_to = "sample_id", values_to = "dev_zscore") %>%
  dplyr::left_join(sample_meta, by = "sample_id") %>%
  dplyr::group_by(long_name, group) %>%
  dplyr::summarise(mean_z = mean(dev_zscore), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = group, values_from = mean_z) %>%
  tibble::column_to_rownames("long_name")

# Fixed, readable column order (drop any group that happens to be absent)
group_order <- c("SH_ADR_NO", "WWTR1_DMSO", "WWTR1_dox", "WWTR1_TDI",
                 "WWTR1_dox_TDI", "SH_M_NO")
group_order <- group_order[group_order %in% colnames(devToPlot_mean)]
devToPlot_mean <- devToPlot_mean[, group_order, drop = FALSE]

# Column annotation: one row per group. Note the annotation_colors list names MUST
# match the annotation columns ("treatment"/"genotype") or pheatmap ignores them.
group_annot <- sample_meta %>%
  dplyr::distinct(group, genotype, treatment) %>%
  tibble::column_to_rownames("group")
group_annot <- group_annot[colnames(devToPlot_mean), , drop = FALSE]

group_annot_colors <- list(
  treatment = c(dox = "#525252", TDI = "#fc4e2a", DMSO = "green",
                dox_TDI = "yellow", NO = "pink"),
  genotype  = c(SH_ADR = "#0a9396", SH_M = "#ee9b00", WWTR1 = "#9b2226")
)

heatmap_avg <- pheatmap::pheatmap(as.matrix(devToPlot_mean),
                                  color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
                                  border_color = NA,
                                  scale = "row",
                                  cluster_cols = FALSE,
                                  cluster_rows = TRUE,
                                  show_colnames = TRUE,
                                  show_rownames = TRUE,
                                  annotation_col = group_annot,
                                  annotation_colors = group_annot_colors,
                                  fontsize_row = 9,
                                  fontsize = 9,
                                  main = paste0("Mean motif deviation z-score per condition (replicates averaged) ", sbst))

ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_OE_Motifs_group_mean_", sbst, ".png")),
       plot = heatmap_avg, width = 16, height = 38, units = "cm")
ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_OE_Motifs_group_mean_", sbst, ".pdf")),
       plot = heatmap_avg, width = 16, height = 38, units = "cm")
```


```{r}
motifs_to_visualise <- c(
  "MA0808.1_TEAD3",
  "MA0090.3_TEAD1",
  "MA0809.2_TEAD4",
  "MA1121.1_TEAD2",
  "MA0496.3_MAFK",
  "MA1928.1_BNC2",
  "MA0835.2_BATF3",
  "MA0462.2_BATF::JUN",
  "MA1634.1_BATF",
  "MA0501.1_MAF::NFE2",
  "MA0089.2_MAFG::NFE2L1",
  "MA1101.2_BACH2",
  "MA0655.1_JDP2",
  "MA0490.2_JUNB",
  "MA0491.2_JUND",
  "MA0477.2_FOSL1",
  "MA1137.1_FOSL1::JUNB",
  "MA1141.1_FOS::JUND",
  "MA1132.1_JUN::JUND",
  "MA1134.1_FOS::JUNB",
  "MA1128.1_FOSL1::JUN",
  "MA1138.1_FOSL2::JUNB",
  "MA1135.1_FOSB::JUNB",
  "MA1130.1_FOSL2::JUN",
  "MA1144.1_FOSL2::JUND",
  "MA0478.1_FOSL2",
  "MA0476.1_FOS",
  "MA0099.3_FOS::JUN",
  "MA1142.1_FOSL1::JUND",
  "MA0841.1_NFE2",
  "MA1633.2_BACH1"
)
```

Visualise the hand-picked motif set above (TEAD / MAF / BATF / AP-1 families) as
two heatmaps of chromVAR deviation z-scores: one ungrouped (one column per
replicate) and one grouped (replicates averaged into one column per condition).
Motifs are drawn in the order listed in `motifs_to_visualise` (rows not
clustered) so the TF families stay together.
```{r selected-motif-heatmaps}
# Build a long_name-keyed z-score table from ALL motifs (not just devTop's top-N),
# so every hand-picked motif is available regardless of its variability ranking.
dev_named <- devZscores_df %>%
  dplyr::left_join(motif_annot, by = c("jaspar_id" = "motif")) %>%
  dplyr::mutate(long_name = paste0(jaspar_id, "_", name))

# Keep the requested motifs, preserving the order given in motifs_to_visualise.
motif_present <- motifs_to_visualise[motifs_to_visualise %in% dev_named$long_name]
motif_missing <- setdiff(motifs_to_visualise, dev_named$long_name)
if (length(motif_missing) > 0) {
  warning("Selected motifs not found in the data and skipped: ",
          paste(motif_missing, collapse = ", "))
}

# Shared per-sample metadata (rownames of colData == z-score column names).
sample_meta <- as.data.frame(colData(OE_ATAC)) %>%
  tibble::rownames_to_column("sample_id") %>%
  dplyr::mutate(group = paste(genotype, treatment, sep = "_")) %>%
  dplyr::select(sample_id, group, genotype, treatment)

annot_colors <- list(
  treatment = c(dox = "#525252", TDI = "#fc4e2a", DMSO = "green",
                dox_TDI = "yellow", NO = "pink"),
  genotype  = c(SH_ADR = "#0a9396", SH_M = "#ee9b00", WWTR1 = "#9b2226")
)
color.scheme_100 <- colorRampPalette(c("navy", "white", "firebrick3"))(100)

## --- Ungrouped: one column per replicate ---------------------------------
# samples_to_subset already lists samples in a sensible grouped order.
sample_order <- samples_to_subset[samples_to_subset %in% colnames(dev_named)]
mat_samples <- dev_named %>%
  dplyr::filter(long_name %in% motif_present) %>%
  dplyr::select(long_name, dplyr::all_of(sample_order)) %>%
  tibble::column_to_rownames("long_name")
mat_samples <- as.matrix(mat_samples[motif_present, , drop = FALSE])

sample_annot <- as.data.frame(colData(OE_ATAC)) %>%
  dplyr::select(treatment, genotype)
sample_annot <- sample_annot[colnames(mat_samples), , drop = FALSE]

heatmap_sel_samples <- pheatmap::pheatmap(mat_samples,
                                          color = color.scheme_100,
                                          border_color = NA,
                                          scale = "row",
                                          cluster_cols = FALSE,
                                          cluster_rows = FALSE,
                                          show_colnames = TRUE,
                                          show_rownames = TRUE,
                                          annotation_col = sample_annot,
                                          annotation_colors = annot_colors,
                                          fontsize_row = 9,
                                          fontsize = 8,
                                          main = "Selected motifs - deviation z-score per replicate")

ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_OE_selected_motifs_per_sample_", sbst, ".png")),
       plot = heatmap_sel_samples, width = 22, height = 26, units = "cm")
ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_OE_selected_motifs_per_sample_", sbst, ".pdf")),
       plot = heatmap_sel_samples, width = 22, height = 26, units = "cm")

## --- Grouped: replicates averaged into one column per condition -----------
mat_group <- dev_named %>%
  dplyr::filter(long_name %in% motif_present) %>%
  dplyr::select(long_name, dplyr::all_of(sample_meta$sample_id)) %>%
  tidyr::pivot_longer(-long_name, names_to = "sample_id", values_to = "dev_zscore") %>%
  dplyr::left_join(sample_meta, by = "sample_id") %>%
  dplyr::group_by(long_name, group) %>%
  dplyr::summarise(mean_z = mean(dev_zscore), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = group, values_from = mean_z) %>%
  tibble::column_to_rownames("long_name")

group_order <- c("SH_ADR_NO", "WWTR1_DMSO", "WWTR1_dox", "WWTR1_TDI",
                 "WWTR1_dox_TDI", "SH_M_NO")
group_order <- group_order[group_order %in% colnames(mat_group)]
mat_group <- as.matrix(mat_group[motif_present, group_order, drop = FALSE])

group_annot <- sample_meta %>%
  dplyr::distinct(group, genotype, treatment) %>%
  tibble::column_to_rownames("group")
group_annot <- group_annot[colnames(mat_group), , drop = FALSE]

heatmap_sel_group <- pheatmap::pheatmap(mat_group,
                                        color = color.scheme_100,
                                        border_color = NA,
                                        scale = "row",
                                        cluster_cols = FALSE,
                                        cluster_rows = FALSE,
                                        show_colnames = TRUE,
                                        show_rownames = TRUE,
                                        annotation_col = group_annot,
                                        annotation_colors = annot_colors,
                                        fontsize_row = 9,
                                        fontsize = 8,
                                        main = "Selected motifs - mean deviation z-score per condition")

ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_OE_selected_motifs_group_mean_", sbst, ".png")),
       plot = heatmap_sel_group, width = 16, height = 26, units = "cm")
ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_OE_selected_motifs_group_mean_", sbst, ".pdf")),
       plot = heatmap_sel_group, width = 16, height = 26, units = "cm")

## --- Same two heatmaps but WITHOUT the untreated "NO" controls ------------
# Drop the treatment == "NO" columns (SH_ADR / SH_M). Columns are subset from the
# unscaled matrices, so scale = "row" re-standardises over the remaining columns.

# Ungrouped, no NO
samples_noNO <- sample_meta$sample_id[sample_meta$treatment != "NO"]
sample_order_noNO <- sample_order[sample_order %in% samples_noNO]
mat_samples_noNO <- mat_samples[, sample_order_noNO, drop = FALSE]
sample_annot_noNO <- sample_annot[sample_order_noNO, , drop = FALSE]

heatmap_sel_samples_noNO <- pheatmap::pheatmap(mat_samples_noNO,
                                               color = color.scheme_100,
                                               border_color = NA,
                                               scale = "row",
                                               cluster_cols = FALSE,
                                               cluster_rows = FALSE,
                                               show_colnames = TRUE,
                                               show_rownames = TRUE,
                                               annotation_col = sample_annot_noNO,
                                               annotation_colors = annot_colors,
                                               fontsize_row = 9,
                                               fontsize = 8,
                                               main = "Selected motifs - deviation z-score per replicate (no NO)")

ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_OE_selected_motifs_per_sample_no_NO_", sbst, ".png")),
       plot = heatmap_sel_samples_noNO, width = 18, height = 26, units = "cm")
ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_OE_selected_motifs_per_sample_no_NO_", sbst, ".pdf")),
       plot = heatmap_sel_samples_noNO, width = 18, height = 26, units = "cm")

# Grouped, no NO
group_order_noNO <- group_order[!group_order %in% c("SH_ADR_NO", "SH_M_NO")]
mat_group_noNO <- mat_group[, group_order_noNO, drop = FALSE]
group_annot_noNO <- group_annot[group_order_noNO, , drop = FALSE]

heatmap_sel_group_noNO <- pheatmap::pheatmap(mat_group_noNO,
                                             color = color.scheme_100,
                                             border_color = NA,
                                             scale = "row",
                                             cluster_cols = FALSE,
                                             cluster_rows = FALSE,
                                             show_colnames = TRUE,
                                             show_rownames = TRUE,
                                             annotation_col = group_annot_noNO,
                                             annotation_colors = annot_colors,
                                             fontsize_row = 9,
                                             fontsize = 8,
                                             main = "Selected motifs - mean deviation z-score per condition (no NO)")

ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_OE_selected_motifs_group_mean_no_NO_", sbst, ".png")),
       plot = heatmap_sel_group_noNO, width = 14, height = 26, units = "cm")
ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_OE_selected_motifs_group_mean_no_NO_", sbst, ".pdf")),
       plot = heatmap_sel_group_noNO, width = 14, height = 26, units = "cm")
```