---
title: "ATAC-seq addtitional analysis - YAP/TAZ inhibition using K-975"
author: "Aleksandr Bykov (Enhanced by Claude Code)"
date: '`r Sys.Date()`'
output:
  html_document: 
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 8,
  dpi = 300
)

```


```{r setup-libraries}
# Set up the environment ####
deg_dir <- "~/neuroblastoma/results/ATAC-seq/ATAC-seq_drug_treatment_no_batch_variable/"
if(!dir.exists(deg_dir)){dir.create(deg_dir)}
## Loading libraries ####
library(dplyr)
library(tidyr)
library(DESeq2)
library(limma)
library(ggplot2)
library(BSgenome.Hsapiens.UCSC.hg38)
library(IRanges)
library(patchwork)
import::from(stringr, str_extract, str_detect)
import::from(GenomicRanges,GRanges)
import::from(rtracklayer, liftOver, import.chain)
import::from(openxlsx, addWorksheet, writeData, saveWorkbook)
import::from(openxlsx2, read_xlsx)
import::from(.from = TFBSTools, getMatrixSet)
import::from(chromVAR, addGCBias)
import::from(.from = RColorBrewer, brewer.pal)
import::from(.from = "~/neuroblastoma/resources/utilityScripts.R",
             "generatePCA", 
             "extract_results_nextflow_output",
             "meanExprsPerGroup_nextflow",
             "extract_results_DDS_HIC",
             "plotVolcano",
             "get_the_poissoon_p_val")
import::from(
  .from = here::here("~/neuroblastoma/resources/UtilityScriptsRNA-seq.R"),
  "filterDatasets",
  .character_only = TRUE
)
```

```{r setup-jaspar}
# BugFix - JASPAR2022 cannot be pulled from the web automatically, so it need to be loaded manually
download.file(url = "https://jaspar2022.genereg.net/download/database/JASPAR2022.sqlite",
              destfile = "~/neuroblastoma/resources/JASPAR2022.sqlite")
JASPAR2022 <-  "~/neuroblastoma/resources/JASPAR2022.sqlite"

```

```{r load-atac-data}
# Analysis part ###########################################################################
# Load and prepare ATAC-seq data for drug treatment analysis
# We use only filtered consensus peaks. And combined annotation from genecode, homer and reg.
tmp_dir <- "~/neuroblastoma/data/ATACseq/"
path_to_the_ATAC_counts_data <- file.path(tmp_dir, "consensus_peaks.mLb.clN.rds")
path_annnotation_ATAC_data <- file.path(tmp_dir, "consensus_peaks.mLb.clN.annotatePeaks.txt")
ATAC_counts_data <- readRDS(path_to_the_ATAC_counts_data)
ATAC_annotation_data <- read.csv(path_annnotation_ATAC_data, sep = "\t")

colnames(ATAC_annotation_data)[1] <- "PeakId"
ATAC_annotation_data <- ATAC_annotation_data %>%
  mutate(PeakId = factor(PeakId, levels = row.names(ATAC_counts_data))) %>%
  arrange(PeakId)

#remove Scaffolds chrs
ATAC_annotation_data <- ATAC_annotation_data[ATAC_annotation_data$Chr %in% c(1:22, "X", "Y"), ]
ATAC_counts_data <- ATAC_counts_data[rownames(ATAC_counts_data) %in% ATAC_annotation_data$PeakId, ]

#add ranges
rowRanges(ATAC_counts_data) <- GenomicRanges::GRanges(seqnames = paste0("chr", ATAC_annotation_data$Chr), 
                                              ranges = IRanges::IRanges(start = ATAC_annotation_data$Start, 
                                                                        end = ATAC_annotation_data$End
                                              )
)
mcols(ATAC_counts_data) <- DataFrame(mcols(ATAC_counts_data), ATAC_annotation_data)
```


```{r subset-drug-treatment-samples}
# Subset to drug treatment samples (CM and SH cell lines)
# Extract metadata for control, 24h, and 48h time points
samples_to_subset <- c(
  "CM_control_ATAC_S165169_REP1",
  "CM_24h_ATAC_S165170_REP1",
  "CM_48h_ATAC_S165167_REP1",
  "SH_control_ATAC_S165166_REP1",
  "SH_24h_ATAC_S165164_REP1",
  "SH_48h_ATAC_S165165_REP1",
  "SH_DMSO_ctrl_rep1_REP1",
  "SH_DMSO_ctrl_rep2_REP1",
  "SH_24h_rep1_REP1",
  "SH_24h_rep2_REP1",
  "SH_48h_rep1_REP1",
  "SH_48h_rep2_REP1",
  "CM_DMSO_ctrl_rep1_REP1",
  "CM_DMSO_ctrl_rep2_REP1",
  "CM_24h_rep1_REP1",
  "CM_24h_rep2_REP1",
  "CM_48h_rep1_REP1",
  "CM_48h_rep2_REP1"
)


drg_trtm_ATAC <- ATAC_counts_data[, colData(ATAC_counts_data)$sample %in% samples_to_subset]

colData(drg_trtm_ATAC)$cell_line <- as.factor(case_when(
  str_detect(drg_trtm_ATAC$sample, "^CM") ~ "CLBMa",
  str_detect(drg_trtm_ATAC$sample, "^SH") ~ "SK_N_SH",
  .default = colData(drg_trtm_ATAC)$sample
))
colData(drg_trtm_ATAC)$time <- as.factor(case_when(
  str_detect(drg_trtm_ATAC$sample, "24h") ~ "24h",
  str_detect(drg_trtm_ATAC$sample, "48h") ~ "48h",
  str_detect(drg_trtm_ATAC$sample, "DMSO") ~ "control",
  str_detect(drg_trtm_ATAC$sample, "control") ~ "control",
  .default = colData(drg_trtm_ATAC)$cell_line
))
colData(drg_trtm_ATAC)$time <- relevel(colData(drg_trtm_ATAC)$time, ref = "control")

# Batch: legacy samples (single replicate, named *_ATAC_S165xxx_REP1) vs the new
# batch (2 replicates, named *_rep1/rep2_REP1). Every cell_line x time combination
# is present in BOTH batches, so batch is balanced and can be corrected for.
colData(drg_trtm_ATAC)$batch <- as.factor(dplyr::if_else(
  str_detect(drg_trtm_ATAC$sample, "_ATAC_S\\d+"), "legacy", "new"
))

# Variable of interest (time) stays last; batch and cell_line are adjusted for.
design(drg_trtm_ATAC) <- as.formula("~ cell_line + time")

#drg_trtm_ATAC@rowRanges@elementMetadata@listData[["PeakId"]]

#
RNA_SEQ_data <- openxlsx2::read_xlsx("~/neuroblastoma/results/RNA-seq/RNA-seq_MES_vs_ADR/tables/MES_vs_ADR_comprehensive_results.xlsx", sheet = 2)
RNA_SEQ_data <- RNA_SEQ_data %>%
  dplyr::mutate(Term = if_else(log2FoldChange < 0, "ADRN", "MES")) %>%
  dplyr::select(gene_symbol, Term)
mcols(drg_trtm_ATAC)$gene_category_our_RNAseq <- "ND"
mcols(drg_trtm_ATAC)$gene_category_our_RNAseq <- case_when(
  mcols(drg_trtm_ATAC)$Gene.Name %in% RNA_SEQ_data[RNA_SEQ_data$Term == "ADRN", 1] ~ "g_ADRN",
  mcols(drg_trtm_ATAC)$Gene.Name %in% RNA_SEQ_data[RNA_SEQ_data$Term == "MES", 1] ~ "g_MES",
  TRUE ~ mcols(drg_trtm_ATAC)$gene_category_our_RNAseq
)

# Add metadata - marks peaks that are around 3kb from TSS
mcols(drg_trtm_ATAC)$is_promoter_3kb <- FALSE
mcols(drg_trtm_ATAC)$is_promoter_3kb[abs(mcols(drg_trtm_ATAC)$Distance.to.TSS) <= 3000] <- TRUE
```

```{r differential-analysis}
# Perform differential accessibility analysis
# Filter low-count peaks and run DESeq2 normalization
drg_trtm_ATAC <- filterDatasets(drg_trtm_ATAC,
                                abs_filt = TRUE,
                                abs_filt_samples = 2
)

design(drg_trtm_ATAC) <- as.formula("~ cell_line + time")
#design(drg_trtm_ATAC) <- as.formula("~ batch + time")
drg_trtm_ATAC <- DESeq(drg_trtm_ATAC)
vsd <- vst(drg_trtm_ATAC, blind = F)

# For visualization only: the design corrects for batch statistically, but the VST
# matrix still carries the batch structure, so PCA/heatmaps would still split by
# batch. Strip it here (NEVER feed this back into DESeq2 - the model handles it).
vsd_uncorrected <- vsd
assay(vsd) <- limma::removeBatchEffect(
  assay(vsd),
  batch  = vsd$batch,
  design = model.matrix(~ cell_line + time, data = as.data.frame(colData(vsd)))
)

pca_dar_uncorrected_time <- generatePCA(transf_object = vsd_uncorrected,
                       cond_interest_varPart = c("time", "cell_line"),
                       color_variable = "time",
                       shape_variable = "cell_line",
                       ntop_genes = 1000) +
  ggtitle("Original dataset (batch effect present)")


pca_dar_uncorrected <- generatePCA(transf_object = vsd_uncorrected,
                       cond_interest_varPart = c("batch", "cell_line"),
                       color_variable = "batch",
                       shape_variable = "cell_line",
                       ntop_genes = 1000) +
  ggtitle("Original dataset (batch effect present)")

pca_dar_time <- generatePCA(transf_object = vsd,
                       cond_interest_varPart = c("time", "cell_line"), 
                       color_variable = "time", 
                       shape_variable = "cell_line",
                       ntop_genes = 1000) +
  ggtitle("Original dataset (batch corrected)")

pca_dar <- generatePCA(transf_object = vsd,
                       cond_interest_varPart = c("batch", "cell_line"),
                       color_variable = "batch",
                       shape_variable = "cell_line",
                       ntop_genes = 1000) +
  ggtitle("Original dataset (batch corrected)")



# ############# test
# # For visualization only: the design corrects for batch statistically, but the VST
# # matrix still carries the batch structure, so PCA/heatmaps would still split by
# # batch. Strip it here (NEVER feed this back into DESeq2 - the model handles it).
# vsd_uncorrected <- vsd
# assay(vsd) <- limma::removeBatchEffect(
#   assay(vsd),
#   batch  = vsd$batch,
#   design = model.matrix(~ time, data = as.data.frame(colData(vsd)))
# )
# 
# pca_dar_uncorrected_time <- generatePCA(transf_object = vsd_uncorrected,
#                        cond_interest_varPart = c("time"),
#                        color_variable = "time",
#                        shape_variable = "cell_line",
#                        ntop_genes = 1000) +
#   ggtitle("Original dataset (batch effect present)")
# 
# 
# pca_dar_uncorrected <- generatePCA(transf_object = vsd_uncorrected,
#                        cond_interest_varPart = c("batch"),
#                        color_variable = "batch",
#                        shape_variable = "cell_line",
#                        ntop_genes = 1000) +
#   ggtitle("Original dataset (batch effect present)")
# 
# pca_dar_time <- generatePCA(transf_object = vsd,
#                        cond_interest_varPart = c("time"), 
#                        color_variable = "time", 
#                        shape_variable = "cell_line",
#                        ntop_genes = 1000) +
#   ggtitle("Original dataset (batch corrected)")
# 
# pca_dar <- generatePCA(transf_object = vsd,
#                        cond_interest_varPart = c("batch"),
#                        color_variable = "batch",
#                        shape_variable = "cell_line",
#                        ntop_genes = 1000) +
#   ggtitle("Original dataset (batch corrected)")
```


```{r plot-pca}
# Generate PCA plot to visualize sample relationships
# Left: batch effect visible; Right: after limma::removeBatchEffect
pca_dar_uncorrected_time + pca_dar_time
pca_dar_uncorrected + pca_dar
```

```{r extract-dar-results}
# Extract differential accessibility results for 48h vs control comparison
resultsNames(drg_trtm_ATAC)
coeff_name <- "time_48h_vs_control"
cond_numerator <- "48h"
cond_denominator <- "control"
cond_variable <- "time"
padj_cutoff = 0.05
log2FC_cutoff = 1
sample_result <- results(drg_trtm_ATAC, contrast = c(cond_variable, cond_numerator, cond_denominator))
ATAC_dds_results_48h <- extract_results_nextflow_output(dds_object = drg_trtm_ATAC,
                                            dds_result = sample_result,
                                            coeff_name = coeff_name,
                                            cond_numerator = cond_numerator,
                                            cond_denominator = cond_denominator,
                                            cond_variable = cond_variable,
                                            padj_cutoff = padj_cutoff,
                                            log2FC_cutoff = log2FC_cutoff)
# Write the result to the xlsx file
XLSX_OUT <- openxlsx::createWorkbook()
openxlsx::addWorksheet(XLSX_OUT, "results_signif")
openxlsx::addWorksheet(XLSX_OUT, "de_details")
openxlsx::addWorksheet(XLSX_OUT, "results_all")
openxlsx::writeData(XLSX_OUT, x = ATAC_dds_results_48h$results_signif, sheet = "results_signif")
openxlsx::writeData(XLSX_OUT, x = ATAC_dds_results_48h$de_details, sheet = "de_details")
openxlsx::writeData(XLSX_OUT, x = ATAC_dds_results_48h$results_all, sheet = "results_all")
openxlsx::saveWorkbook(XLSX_OUT, file.path(deg_dir,  paste0("DAR_results_time_48h_vs_control.xlsx")), overwrite = T)


resultsNames(drg_trtm_ATAC)
coeff_name <- "time_24h_vs_control"
cond_numerator <- "24h"
cond_denominator <- "control"
cond_variable <- "time"
padj_cutoff = 0.05
log2FC_cutoff = 1
sample_result <- results(drg_trtm_ATAC, contrast = c(cond_variable, cond_numerator, cond_denominator))
ATAC_dds_results_24h <- extract_results_nextflow_output(dds_object = drg_trtm_ATAC,
                                            dds_result = sample_result,
                                            coeff_name = coeff_name,
                                            cond_numerator = cond_numerator,
                                            cond_denominator = cond_denominator,
                                            cond_variable = cond_variable,
                                            padj_cutoff = padj_cutoff,
                                            log2FC_cutoff = log2FC_cutoff)

XLSX_OUT <- openxlsx::createWorkbook()
openxlsx::addWorksheet(XLSX_OUT, "results_signif")
openxlsx::addWorksheet(XLSX_OUT, "de_details")
openxlsx::addWorksheet(XLSX_OUT, "results_all")
openxlsx::writeData(XLSX_OUT, x = ATAC_dds_results_24h$results_signif, sheet = "results_signif")
openxlsx::writeData(XLSX_OUT, x = ATAC_dds_results_24h$de_details, sheet = "de_details")
openxlsx::writeData(XLSX_OUT, x = ATAC_dds_results_24h$results_all, sheet = "results_all")
openxlsx::saveWorkbook(XLSX_OUT, file.path(deg_dir,  paste0("DAR_results_time_24h_vs_control.xlsx")), overwrite = T)
```



```{r generate-heatmaps}
# Generate heatmaps of differentially accessible regions
# Focus on peaks associated with MES and ADRN gene signatures
## Now the same analysis, but for the promoter regions - ± 3000bp around TSSs ####
ATAC_dds <- drg_trtm_ATAC
#ATAC_dds <- ATAC_dds[mcols(ATAC_dds)$is_promoter_3kb,]

# Uncoment only for promoter regions
# ATAC_dds <- DESeq(ATAC_dds)
# 
# vsd <- vst(ATAC_dds, blind = F)
# 
# # Visualization only: remove batch so heatmap columns cluster by treatment, not batch.
# assay(vsd) <- limma::removeBatchEffect(
#   assay(vsd),
#   batch  = vsd$batch,
#   design = model.matrix(~ cell_line + time, data = as.data.frame(colData(vsd)))
# )
# 
# pca_dar <- generatePCA(transf_object = vsd,
#                        cond_interest_varPart = c("time", "cell_line"), 
#                        color_variable = "time", 
#                        shape_variable = "cell_line",
#                        ntop_genes = 1000) +
#   ggtitle("3kb Promoters")
# plot(pca_dar)

# # extracting results
# resultsNames(ATAC_dds)
# coeff_name <- "time_48h_vs_control"
# cond_numerator <-  "48h"
# cond_denominator <-  "control"
# cond_variable <- "time"
# padj_cutoff = 0.05
# log2FC_cutoff = 1
# 
# ATAC_dds_results <- extract_results_nextflow_output(dds_object = ATAC_dds,
#                                         coeff_name = coeff_name,
#                                         cond_numerator = cond_numerator,
#                                         cond_denominator = cond_denominator,
#                                         cond_variable = cond_variable,
#                                         padj_cutoff = padj_cutoff,
#                                         log2FC_cutoff = log2FC_cutoff)

ATAC_dds_results <- ATAC_dds_results_48h
# Making Heatmap
metadata_heatmap <- as.data.frame(colData(ATAC_dds))
heatmap_counts <- SummarizedExperiment::assay(vsd)
rownames(heatmap_counts) <- vsd@rowRanges$PeakId

all_ADR_peaks <- c(mcols(drg_trtm_ATAC)$PeakId[mcols(drg_trtm_ATAC)$gene_category_our_RNAseq == "g_ADRN"])
all_MES_peaks <- c(mcols(drg_trtm_ATAC)$PeakId[mcols(drg_trtm_ATAC)$gene_category_our_RNAseq == "g_MES"])

# removed experiment batch effects! ? use experimen+cell_line removed???
ATAC_signif <- ATAC_dds_results$results_signif
heatmap_counts <- heatmap_counts[rownames(heatmap_counts) %in% ATAC_dds_results$results_signif$PeakId, ]
heatmap_counts_MES_spec_genes <- heatmap_counts[rownames(heatmap_counts) %in% all_MES_peaks, ]
heatmap_counts_ADRN_spec_genes <- heatmap_counts[rownames(heatmap_counts) %in% all_ADR_peaks, ]

annotation_col <- metadata_heatmap %>%
  dplyr::select(cell_line, time) %>% 
  dplyr::arrange(time, cell_line)
heatmap_counts <- heatmap_counts[, match(rownames(annotation_col), colnames(heatmap_counts))]

heatmap_counts_MES_spec_genes <- heatmap_counts_MES_spec_genes[, match(rownames(annotation_col), colnames(heatmap_counts_MES_spec_genes))]
rownames(heatmap_counts_MES_spec_genes) <- ATAC_signif[ATAC_signif$PeakId %in% row.names(heatmap_counts_MES_spec_genes), "Gene.Name"]

heatmap_counts_ADRN_spec_genes <- heatmap_counts_ADRN_spec_genes[, match(rownames(annotation_col), colnames(heatmap_counts_ADRN_spec_genes))]
rownames(heatmap_counts_ADRN_spec_genes) <- ATAC_signif[ATAC_signif$PeakId %in% row.names(heatmap_counts_ADRN_spec_genes), "Gene.Name"]

#heatmap_row_annot_MES_spec_genes <- data.frame(gene_name = ATAC_signif[ATAC_signif$PeakId %in% row.names(heatmap_counts_MES_spec_genes), "Gene.Name"],
#                                               row.names =  row.names(heatmap_counts_MES_spec_genes))

ensembl2symbol_annot <- ATAC_dds_results$results_all %>%
  dplyr::select(PeakId, Gene.Name)

color.scheme <- colorRampPalette(c("navy", "white", "firebrick3"))(50)
ann_colors = list(
  time = c( control = "#005f73", `24h` = "#0a9396", `48h` = "#94d2bd"),
  cell_line = c(CLBMa = "#E9D8A6", SK_N_SH = "#D9BE6D")
)

heatmap <- pheatmap::pheatmap(heatmap_counts,
                              main = "Heatmap of signif. All DARs 48h vs control.",
                              scale = "row",
                              annotation_col = annotation_col,
                              annotation_colors = ann_colors,
                              show_colnames = FALSE,
                              show_rownames = FALSE,
                              cluster_cols = FALSE,
                              color = color.scheme,
                              fontsize = 10, fontsize_row = 10) 
ggsave(filename = file.path(deg_dir, "heatmap_48h_vs_control_all_DARs_ATAC.png"), 
       plot = heatmap,
       width = 18, height = 20, units = "cm")
ggsave(filename = file.path(deg_dir, "heatmap_48h_vs_control_all_DARs_ATAC.pdf"), 
       plot = heatmap,
       width = 18, height = 20, units = "cm")

heatmap <- pheatmap::pheatmap(heatmap_counts_MES_spec_genes,
                              main = "Heatmap of signif. DAR ADRN vs MES all DARs \n MES-specific genes only (defined by RNA-seq)",
                              scale = "row",
                              annotation_col = annotation_col,
                              annotation_colors = ann_colors,
                              #annotation_row = heatmap_row_annot_MES_spec_genes,
                              show_colnames = FALSE,
                              show_rownames = TRUE,
                              cluster_cols = FALSE,
                              color = color.scheme,
                              fontsize = 10, fontsize_row = 10) #height=10, cellwidth = 11, cellheight = 11
ggsave(filename = file.path(deg_dir, "heatmap_48h_vs_control_all_DARs_MES-RNAseq-specific.png"), 
       plot=heatmap,
       width = 18, height = 20, units = "cm")
ggsave(filename = file.path(deg_dir, "heatmap_48h_vs_control_all_DARs_MES-RNAseq-specific.pdf"), 
       plot=heatmap,
       width = 18, height = 20, units = "cm")

heatmap <- pheatmap::pheatmap(heatmap_counts_ADRN_spec_genes,
                              main = "Heatmap of signif. DAR ADRN vs MES all DARs \n ADRN-specific genes only (defined by RNA-seq)",
                              scale = "row",
                              annotation_col = annotation_col,
                              annotation_colors = ann_colors,
                              #annotation_row = row_annot_symbols,
                              show_colnames = FALSE,
                              show_rownames = FALSE,
                              cluster_cols = FALSE,
                              color = color.scheme,
                              fontsize = 10, fontsize_row = 10) #height=10, cellwidth = 11, cellheight = 11
ggsave(filename = file.path(deg_dir, "heatmap_48h_vs_control_all_DARs_ADRN-RNAseq-specific.png"), 
       plot=heatmap,
       width = 18, height = 20, units = "cm")
ggsave(filename = file.path(deg_dir, "heatmap_48h_vs_control_all_DARs_ADRN-RNAseq-specific.pdf"), 
       plot=heatmap,
       width = 18, height = 20, units = "cm")
```

Here we identify what are the motifs that that present in differentially accessible regions and how they are differ in control (DMSO) vs 24h and 48h samples.
```{r chromvar-motif-analysis}
# ChromVAR motif enrichment analysis for drug treatment samples
# Identify transcription factor motifs that change accessibility over time
opts <- list()
opts[["tax_group"]] <- "vertebrates"
opts[["species"]] <- "9606"
opts[["collection"]] <- "CORE"
opts[["all_versions"]] <- FALSE
motifsToScan <- TFBSTools::getMatrixSet(JASPAR2022, opts)

# Using different subsets of peaks to check 
subsets_to_run <- c("ALL")

sbst <-  c("ALL")
# Filter peaks with low read counts across samples (minimum 5 reads total)
dim(drg_trtm_ATAC)
counts_consensus_filt <- drg_trtm_ATAC[rowSums(assay(drg_trtm_ATAC)) > 5, ]
dim(counts_consensus_filt)

print(paste("procesing: ", sbst ))

# ChromVar package uses GC content to identify background peaks that are likely to be non-functional and exclude them from further analysis. 
counts_consensus_filt <- chromVAR::addGCBias(counts_consensus_filt, 
                                             genome = BSgenome.Hsapiens.UCSC.hg38) 

# Having corrected for bias, we can use the matchMotifs function to identify motifs under our ATACseq peaks.
motif_matches <- motifmatchr::matchMotifs(pwms = motifsToScan, 
                                          subject = counts_consensus_filt, 
                                          genome = BSgenome.Hsapiens.UCSC.hg38, 
                                          out = "matches")
motif_matches

BiocParallel::register(BiocParallel::MulticoreParam(8, progressbar = FALSE))
set.seed(42069)

# Run chromVar ----
#Following the identification of motifs in our peaks, we can perform the summarization of ATACseq signal to motifs using the computeDeviations and the computeVariability functions.
background_peaks <- chromVAR::getBackgroundPeaks(object = counts_consensus_filt) 
access_expectation <- chromVAR::computeExpectations(object = counts_consensus_filt)
chrom_access_deviations <- chromVAR::computeDeviations(object = counts_consensus_filt, 
                                                       annotations = motif_matches,
                                                       background_peaks = background_peaks,
                                                       expectation = access_expectation)

# Generate sample correlation matrix to assess data quality and batch effects
sample_cor <- chromVAR::getSampleCorrelation(chrom_access_deviations)
annotation_row_sampleCor <- as.data.frame(SummarizedExperiment::colData(chrom_access_deviations)) %>%
  dplyr::select(sample)
pheatmap::pheatmap(as.dist(sample_cor), 
                   annotation_row = annotation_row_sampleCor, 
                   clustering_distance_rows = as.dist(1-sample_cor), 
                   clustering_distance_cols = as.dist(1-sample_cor),
                   annotation_names_row = TRUE 
                  ) 

# Perform t-SNE dimensionality reduction to visualize sample clustering
tsne_results <- chromVAR::deviationsTsne(chrom_access_deviations, 
                                         threshold = 1.5,
                                         perplexity = 8, 
                                         what = "samples",
                                         shiny = FALSE)

tsne_plots <- chromVAR::plotDeviationsTsne(chrom_access_deviations, 
                                           tsne_results,
                                           sample_column = "sample",
                                           shiny = FALSE)

diff_acc <- chromVAR::differentialDeviations(chrom_access_deviations, 
                                             groups = "time",
                                             parametric = FALSE)

motif_annot <- as.data.frame(SummarizedExperiment::rowData(chrom_access_deviations)) %>%
  tibble::rownames_to_column(var = "motif") %>%
  dplyr::select(motif, name)
diff_acc_annot <- diff_acc %>%
  tibble::rownames_to_column(var = "motif") %>%
  dplyr::left_join(., motif_annot, by = "motif")
head(diff_acc_annot) 

diff_var <- chromVAR::differentialVariability(chrom_access_deviations, "time", 
                                              parametric = FALSE)
diff_var_annot <- diff_var %>%
  tibble::rownames_to_column(var = "motif") %>%
  dplyr::left_join(., motif_annot, by = "motif")
head(diff_var_annot)

devZscores <- chromVAR::deviationScores(chrom_access_deviations)
devZscores_df <- tibble::rownames_to_column(as.data.frame(devZscores),
                                            var = "jaspar_id")

#The function computeVariability returns a data.frame that contains the variability (standard deviation of the z scores computed above
# across all cell/samples for a set of peaks), bootstrap confidence intervals for that variability (by resampling cells/samples), 
# and a p-value for the variability being greater than the null hypothesis of 1.
chrom_access_variability <- chromVAR::computeVariability(chrom_access_deviations)
chrom_access_variability_plot <- chromVAR::plotVariability(chrom_access_variability,
                                                           use_plotly = FALSE)

ggsave(chrom_access_variability_plot, 
       filename = file.path(deg_dir, paste0("ChromVAR_chrom_access_variability_corr_plot_", sbst, ".pdf")))

BiocParallel::register(BiocParallel::SerialParam())

chrom_access_variability_ord <- chrom_access_variability[order(chrom_access_variability$p_value), ]  # ordering based on p-value

#! to store
chrom_access_variability_ord_results <- chrom_access_variability_ord %>%
  tibble::rownames_to_column(var="jaspar_id") %>%
  dplyr::left_join(., devZscores_df, by = "jaspar_id")

message("Saving motif enrichment results")
ntop <- 50
topVariable <- chrom_access_variability_ord[1:ntop, ]
topVariable <- tibble::rownames_to_column(topVariable, var="jaspar_id")

devTop <- topVariable %>%
  dplyr::left_join(., devZscores_df, by = "jaspar_id")

devTop <- devTop %>% mutate(long_name = paste0(jaspar_id, "_", name))

devToPlot <- devTop %>%
  dplyr::select(long_name, 
                CM_control_ATAC_S165169_REP1,
                CM_24h_ATAC_S165170_REP1,
                CM_48h_ATAC_S165167_REP1,
                SH_control_ATAC_S165166_REP1,
                SH_24h_ATAC_S165164_REP1,
                SH_48h_ATAC_S165165_REP1,
                SH_DMSO_ctrl_rep1_REP1,
                SH_DMSO_ctrl_rep2_REP1,
                SH_24h_rep1_REP1,     
                SH_24h_rep2_REP1,     
                SH_48h_rep1_REP1,
                SH_48h_rep2_REP1,
                CM_DMSO_ctrl_rep1_REP1,
                CM_DMSO_ctrl_rep2_REP1,
                CM_24h_rep1_REP1,
                CM_24h_rep2_REP1,
                CM_48h_rep1_REP1,
                CM_48h_rep2_REP1
  ) %>%
  tibble::column_to_rownames(var = "long_name")

annotCol_forHeatmap <- as.data.frame(colData(drg_trtm_ATAC)) %>%
  dplyr::select(time)
# annotCol_forHeatmap_colors <- list(phenotype = c(A = "#525252", M = "#fc4e2a"))
annotCol_forHeatmap_colors <- list(phenotype = c(`24h` = "#525252",
                                                 `48h` = "#fc4e2a",
                                                 control = "green"))


color.scheme <- colorRampPalette(c("navy", "white", "firebrick3"))(50)

heatmap <- pheatmap::pheatmap(as.matrix(devToPlot),
                              color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
                              border_color = NA,
                              #color = colorRampPalette(rev(brewer.pal(7, "RdBu")))(100),
                              scale = "row",
                              cluster_cols = FALSE,
                              cluster_rows = TRUE,
                              show_colnames = TRUE,
                              show_rownames = TRUE,
                              annotation_col = annotCol_forHeatmap,
                              annotation_colors = annotCol_forHeatmap_colors,  
                              fontsize_row = 9,
                              fontsize = 9,
                              main = paste0("Heatmap of signif. ATAC-seq Drug treatment control vs 24h vs 48h. ", sbst, "- used"))

ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_drug_treatment_control_vs_24h_48h_Motifs", sbst, "- used", ".png")), 
       plot=heatmap,
       width = 18, height = 20, units = "cm")
ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_drug_treatment_control_vs_24h_48h_Motifs", sbst, "- used", ".pdf")), 
       plot=heatmap,
       width = 18, height = 20, units = "cm")

# Generate separate heatmap for CM cell line only
devToPlot_CM <- devTop %>%
  dplyr::select(long_name, 
                CM_control_ATAC_S165169_REP1,
                CM_DMSO_ctrl_rep1_REP1,
                CM_DMSO_ctrl_rep2_REP1,
                CM_24h_ATAC_S165170_REP1,
                CM_24h_rep1_REP1,
                CM_24h_rep2_REP1,
                CM_48h_ATAC_S165167_REP1,
                CM_48h_rep1_REP1,
                CM_48h_rep2_REP1 
  ) %>%
  tibble::column_to_rownames(var = "long_name")

heatmap <- pheatmap::pheatmap(as.matrix(devToPlot_CM),
                              color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
                              border_color = NA,
                              #color = colorRampPalette(rev(brewer.pal(7, "RdBu")))(100),
                              scale = "row",
                              cluster_cols = FALSE,
                              cluster_rows = TRUE,
                              show_colnames = TRUE,
                              show_rownames = TRUE,
                              annotation_col = annotCol_forHeatmap,
                              annotation_colors = annotCol_forHeatmap_colors,  
                              fontsize_row = 9,
                              fontsize = 9,
                              main = paste0("Heatmap of signif. CM line ATAC-seq Drug treatment control vs 24h vs 48h. ", sbst, "- used"))

ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_CM_line_drug_treatment_control_vs_24h_48h_Motifs_", sbst, "- used", ".png")), 
       plot=heatmap,
       width = 18, height = 20, units = "cm")
ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_CM_line_drug_treatment_control_vs_24h_48h_Motifs_", sbst, "- used", ".pdf")), 
       plot=heatmap,
       width = 18, height = 20, units = "cm")


# Generate separate heatmap for SH cell line only
devToPlot_SH <- devTop %>%
  dplyr::select(long_name, 
                SH_control_ATAC_S165166_REP1,
                SH_DMSO_ctrl_rep1_REP1,
                SH_DMSO_ctrl_rep2_REP1,
                SH_24h_ATAC_S165164_REP1,
                SH_24h_rep1_REP1,     
                SH_24h_rep2_REP1,     
                SH_48h_ATAC_S165165_REP1,
                SH_48h_rep1_REP1,
                SH_48h_rep2_REP1,
  ) %>%
  tibble::column_to_rownames(var = "long_name")

heatmap <- pheatmap::pheatmap(as.matrix(devToPlot_SH),
                              color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
                              border_color = NA,
                              #color = colorRampPalette(rev(brewer.pal(7, "RdBu")))(100),
                              scale = "row",
                              cluster_cols = FALSE,
                              cluster_rows = TRUE,
                              show_colnames = TRUE,
                              show_rownames = TRUE,
                              annotation_col = annotCol_forHeatmap,
                              annotation_colors = annotCol_forHeatmap_colors,  
                              fontsize_row = 9,
                              fontsize = 9,
                              main = paste0("Heatmap of signif. SH line ATAC-seq Drug treatment control vs 24h vs 48h. ", sbst, "- used"))

ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_SH_line_drug_treatment_control_vs_24h_48h_Motifs_", sbst, "- used", ".png")), 
       plot=heatmap,
       width = 18, height = 20, units = "cm")
ggsave(filename = file.path(deg_dir, paste0("heatmap_ATAC_SH_line_drug_treatment_control_vs_24h_48h_Motifs_", sbst, "- used", ".pdf")),
       plot=heatmap,
       width = 18, height = 20, units = "cm")
```

The heatmaps are row-scaled, so they hide the absolute magnitude of the batch
offset. To see how the raw chromVAR deviation z-score for a single motif actually
moves over the treatment time course - and how differently the legacy vs new
batch behave - we plot it directly. Each line is the mean per cell line x batch;
points are individual replicates.
```{r fosl1jun-lineplot}
# Track one motif's chromVAR deviation z-score across the time course, split by batch
motif_of_interest <- "MA1128.1"          # FOSL1::JUN
motif_label       <- "MA1128.1_FOSL1::JUN"

motif_of_interest <- "MA0808.1"          # FOSL1::JUN
motif_label       <- "MA0808.1_TEAD3"

# Per-sample metadata (rownames == column names of the deviation-score matrix)
motif_meta <- as.data.frame(colData(drg_trtm_ATAC)) %>%
  tibble::rownames_to_column(var = "sample_id") %>%
  dplyr::select(sample_id, cell_line, time, batch)

# Long table of the z-scores for this motif, one row per sample
motif_scores <- devZscores_df %>%
  dplyr::filter(str_detect(jaspar_id, motif_of_interest)) %>%
  tidyr::pivot_longer(-jaspar_id, names_to = "sample_id", values_to = "dev_zscore") %>%
  dplyr::left_join(motif_meta, by = "sample_id") %>%
  dplyr::mutate(
    time  = factor(time, levels = c("control", "24h", "48h")),
    group = paste(cell_line, batch, sep = " - ")
  )

if (nrow(motif_scores) == 0) {
  warning("Motif ", motif_of_interest, " not found in devZscores_df - skipping line plot.")
} else {
  # Mean per cell_line x batch x time drives the connecting lines
  motif_summary <- motif_scores %>%
    dplyr::group_by(cell_line, batch, group, time) %>%
    dplyr::summarise(mean_z = mean(dev_zscore), .groups = "drop")

  fosl1jun_lineplot <- ggplot(
      motif_scores,
      aes(x = time, y = dev_zscore, color = cell_line, linetype = batch, group = group)
    ) +
    geom_point(size = 2, alpha = 0.7,
               position = position_dodge(width = 0.15)) +
    geom_line(data = motif_summary, aes(y = mean_z), linewidth = 1) +
    labs(
      title    = paste0("chromVAR deviation z-score: ", motif_label),
      subtitle = "Line = mean per cell line x batch; points = individual replicates",
      x = "Treatment time", y = "Deviation z-score",
      color = "Cell line", linetype = "Batch"
    ) +
    theme_bw()

  print(fosl1jun_lineplot)

  ggsave(filename = file.path(deg_dir, paste0("lineplot_", motif_label, "_dev_zscore_by_batch.png")),
         plot = fosl1jun_lineplot, width = 18, height = 12, units = "cm")
  ggsave(filename = file.path(deg_dir, paste0("lineplot_", motif_label, "_dev_zscore_by_batch.pdf")),
         plot = fosl1jun_lineplot, width = 18, height = 12, units = "cm")
}
```


We want to check how TEAD/AP1 motifs are co-localized in the 24h sample. We will test the real data and make the comparison to simultaed data.
```{r Comparison to simulated data }
# Compare real vs simulated data to assess enrichment significance
h24_control_specific_peaks <- ATAC_dds_results_24h$results_signif %>% 
  dplyr::filter(log2FoldChange < 0) %>% 
  dplyr::pull(PeakId)
h24_specific_peaks <- ATAC_dds_results_24h$results_signif %>% 
  dplyr::filter(log2FoldChange > 0) %>% 
  dplyr::pull(PeakId)

opts <- list()
opts[["tax_group"]] <- "vertebrates"
opts[["species"]] <- "9606"
opts[["collection"]] <- "CORE"
opts[["all_versions"]] <- FALSE
motifsToScan <- TFBSTools::getMatrixSet(JASPAR2022, opts)
TEAD_AP1_motifs_to_scan_names <- c("MA0090.3", "MA0808.1", 
                                 "MA0809.2", "MA1121.1",
                                 "MA0099.3")
TEAD_AP1_motifsToScan <- motifsToScan[TEAD_AP1_motifs_to_scan_names,]

# Using different subsets of peaks to check 
drg_trtm_ATAC
counts_consensus_filt_control_24h <- drg_trtm_ATAC[drg_trtm_ATAC@rowRanges@elementMetadata@listData[["PeakId"]] %in% h24_control_specific_peaks, ]
counts_consensus_filt_24h <- drg_trtm_ATAC[drg_trtm_ATAC@rowRanges@elementMetadata@listData[["PeakId"]] %in% h24_specific_peaks, ]

# Generate randomized peak sets as background controls
# Create matched random regions for statistical comparison
# Load genome
genome_used <- getBSgenome("BSgenome.Hsapiens.UCSC.hg38")
seqnames(genome_used) <- sub("^.{1,2}_", "", seqnames(genome_used))
seqnames(genome_used) <- sub("v", ".", seqnames(genome_used))
seqnames(genome_used) <- sub("^M", "MT", seqnames(genome_used))

# load mask regions 
blacklist_regions <- read.delim("~/neuroblastoma/resources/hg38-blacklist.v2.bed", header = FALSE)
blacklist_regions <- makeGRangesFromDataFrame(blacklist_regions, 
                                              seqnames.field = "V1", 
                                              start.field = "V2", 
                                              end.field = "V3")

# Create set of random Peaks for control_24h 
initial_peak_set <- reduce(counts_consensus_filt_control_24h@rowRanges)
random_peaks_control_24h <- initial_peak_set
random_peaks_control_24h@seqnames <- droplevels(random_peaks_control_24h@seqnames)
random_peaks_control_24h <- regioneR::randomizeRegions(random_peaks_control_24h,
                                               allow.overlaps = FALSE,
                                               genome = genome_used,
                                               per.chromosome = TRUE,
                                               mask = blacklist_regions)
random_peaks_control_24h <- as.data.frame(random_peaks_control_24h, 
                                          row.names = NULL,
                                          optional = FALSE)
random_peaks_control_24h$seqnames <- droplevels(random_peaks_control_24h$seqnames)
random_peaks_control_24h <- makeGRangesFromDataFrame(random_peaks_control_24h)

# Create set of random Peaks for 24h 
initial_peak_set <- reduce(counts_consensus_filt_24h@rowRanges)
random_peaks_24h <- initial_peak_set
random_peaks_24h@seqnames <- droplevels(random_peaks_24h@seqnames)
random_peaks_24h <- regioneR::randomizeRegions(random_peaks_24h,
                                               allow.overlaps = FALSE,
                                               genome = genome_used,
                                               per.chromosome = TRUE,
                                               mask = blacklist_regions)
random_peaks_24h <- as.data.frame(random_peaks_24h, 
                                  row.names = NULL,
                                  optional = FALSE)
random_peaks_24h$seqnames <- droplevels(random_peaks_24h$seqnames)
random_peaks_24h <- makeGRangesFromDataFrame(random_peaks_24h)

counts_consensus_filt_control_24h <- chromVAR::addGCBias(counts_consensus_filt_control_24h, 
                                                         genome = BSgenome.Hsapiens.UCSC.hg38) 
counts_consensus_filt_24h <- chromVAR::addGCBias(counts_consensus_filt_24h, 
                                                 genome = BSgenome.Hsapiens.UCSC.hg38) 

# find TEAD and AP1 motifs in real control_24h data
motif_matches_control_24h <- motifmatchr::matchMotifs(pwms = TEAD_AP1_motifsToScan, 
                                              subject = counts_consensus_filt_control_24h, 
                                              genome = BSgenome.Hsapiens.UCSC.hg38, 
                                              out = "positions")
control_24h_TEAD <- c(motif_matches_control_24h$MA0090.3, 
                      motif_matches_control_24h$MA0808.1, 
                      motif_matches_control_24h$MA0809.2, 
                      motif_matches_control_24h$MA1121.1)
control_24h_TEAD <- reduce(control_24h_TEAD)
control_24h_AP1 <- reduce(motif_matches_control_24h$MA0099.3)
control_24h_AP1_TEAD_overlaps <- findOverlaps(control_24h_AP1, 
                                              control_24h_TEAD, 
                                              ignore.strand = TRUE, 
                                              maxgap = 50)

# find TEAD and AP1 motifs in simulated MES data
motif_matches_control_24h_random <- motifmatchr::matchMotifs(pwms = TEAD_AP1_motifsToScan, 
                                                     subject = random_peaks_control_24h, 
                                                     genome = BSgenome.Hsapiens.UCSC.hg38, 
                                                     out = "positions")
control_24h_random_TEAD <- c(motif_matches_control_24h_random$MA0090.3, 
                             motif_matches_control_24h_random$MA0808.1, 
                             motif_matches_control_24h_random$MA0809.2, 
                             motif_matches_control_24h_random$MA1121.1)
control_24h_random_TEAD <- reduce(control_24h_random_TEAD)
control_24h_random_AP1 <- reduce(motif_matches_control_24h_random$MA0099.3)
control_24h_random_AP1_TEAD_overlaps <- findOverlaps(control_24h_random_AP1, 
                                                     control_24h_random_TEAD, 
                                                     ignore.strand = TRUE, 
                                                     maxgap = 50)

# find TEAD and AP1 motifs in real 24h data
motif_matches_24h <- motifmatchr::matchMotifs(pwms = TEAD_AP1_motifsToScan, 
                                              subject = counts_consensus_filt_24h, 
                                              genome = BSgenome.Hsapiens.UCSC.hg38, 
                                              out = "positions")
h24_TEAD <- c(motif_matches_24h$MA0090.3, 
              motif_matches_24h$MA0808.1, 
              motif_matches_24h$MA0809.2, 
              motif_matches_24h$MA1121.1)
h24_TEAD <- reduce(h24_TEAD)
h24_AP1 <- reduce(motif_matches_24h$MA0099.3)
h24_AP1_TEAD_overlaps <- findOverlaps(h24_AP1, 
                                      h24_TEAD, 
                                      ignore.strand = TRUE, 
                                      maxgap = 50)

# find TEAD and AP1 motifs in simulated 24h data
motif_matches_24h_random <- motifmatchr::matchMotifs(pwms = TEAD_AP1_motifsToScan, 
                                                     subject = random_peaks_24h, 
                                                     genome = BSgenome.Hsapiens.UCSC.hg38, 
                                                     out = "positions")
h24_random_TEAD <- c(motif_matches_24h_random$MA0090.3, 
                     motif_matches_24h_random$MA0808.1, 
                     motif_matches_24h_random$MA0809.2, 
                     motif_matches_24h_random$MA1121.1)
h24_random_TEAD <- reduce(h24_random_TEAD)
h24_random_AP1 <- reduce(motif_matches_24h_random$MA0099.3)
h24_random_AP1_TEAD_overlaps <- findOverlaps(h24_random_AP1, 
                                             h24_random_TEAD, 
                                             ignore.strand = TRUE, 
                                             maxgap = 50)


summary_table_peaks <- data.frame(total_number_of_peaks = c(length(counts_consensus_filt_control_24h), 
                                                            length(random_peaks_control_24h),
                                                            length(counts_consensus_filt_24h),
                                                            length(random_peaks_24h)),
                                  TEAD_sites = c(length(control_24h_TEAD),
                                                 length(control_24h_random_TEAD),
                                                 length(h24_TEAD),
                                                 length(h24_random_TEAD)),
                                  AP1_sites = c(length(control_24h_AP1),
                                                length(control_24h_random_AP1),
                                                length(h24_AP1),
                                                length(h24_random_AP1)),
                                  TEAD_AP1_coloc_sites = c(length(control_24h_AP1_TEAD_overlaps),
                                                           length(control_24h_random_AP1_TEAD_overlaps),
                                                           length(h24_AP1_TEAD_overlaps),
                                                           length(h24_random_AP1_TEAD_overlaps)),
                                  row.names = c("control_24h", 
                                                "Simulated control_24h", 
                                                "h24", 
                                                "Simulated h24")
)
summary_table_peaks

# Perform Fisher's exact tests to assess statistical significance of TEAD/AP1 co-localization
contingency_table <- matrix(c(length(h24_AP1_TEAD_overlaps), 
                              length(h24_TEAD) + length(h24_AP1) - length(h24_AP1_TEAD_overlaps), 
                              length(control_24h_AP1_TEAD_overlaps), 
                              length(control_24h_TEAD) + length(control_24h_AP1) - length(control_24h_AP1_TEAD_overlaps)), 
                            nrow = 2, 
                            byrow = TRUE)
fisher_results <- fisher.test(contingency_table, alternative = "two.sided")
fisher_summary_table <- data.table::data.table(comparison = "24h_vs_control_24h",
                                               pval = fisher_results$p.value,
                                               odds_ratio = fisher_results$estimate)

# Fisher test comparing real h24 AP1_TEAD peaks vs simulated
contingency_table_h24_vs_simulated <- matrix(c(length(h24_AP1_TEAD_overlaps), 
                                               length(h24_TEAD) + length(h24_AP1) - length(h24_AP1_TEAD_overlaps),
                                               length(h24_random_AP1_TEAD_overlaps), 
                                               length(h24_random_TEAD) + length(h24_random_AP1) - length(h24_random_AP1_TEAD_overlaps)), 
                                             nrow = 2, 
                                             byrow = TRUE)
row.names(contingency_table_h24_vs_simulated) <- c("h24_peaks", "h24_peaks_random")
colnames(contingency_table_h24_vs_simulated) <- c("TEAD_AB1_colocalisation_present", "TEAD_AB1_colocalisation_absent")
contingency_table_h24_vs_simulated
fisher_results <- fisher.test(contingency_table_h24_vs_simulated, alternative="two.sided")
fisher_summary_table <- rbind(fisher_summary_table,
                              data.table::data.table(comparison = "h24_vs_h24sim",
                                                     pval = fisher_results$p.value,
                                                     odds_ratio = fisher_results$estimate))

# Fisher test for TEAD peaks
contingency_table_TEAD <- matrix(c(length(h24_TEAD), 
                                   length(counts_consensus_filt_24h) - length(h24_TEAD),
                                   length(control_24h_TEAD), 
                                   length(counts_consensus_filt_control_24h) - length(control_24h_TEAD)), 
                                 nrow = 2, 
                                 byrow = TRUE)
row.names(contingency_table_TEAD) <- c("h24_peaks", "control_24h_peaks")
colnames(contingency_table_TEAD) <- c("TEAD_site_present", "TEAD_site_absent")
contingency_table_TEAD
fisher.test(contingency_table_TEAD, 
            alternative = "two.sided")
chisq.test(contingency_table_TEAD)

# Fisher test for AP1 peaks
contingency_table_AP1 <- matrix(c(length(h24_AP1), 
                                  length(counts_consensus_filt_24h) - length(h24_AP1),
                                  length(control_24h_AP1), 
                                  length(counts_consensus_filt_control_24h) - length(control_24h_AP1)),
                                nrow = 2,
                                byrow = TRUE)
row.names(contingency_table_AP1) <- c("h24_peaks", "control_24h_peaks")
colnames(contingency_table_AP1) <- c("AP1_site_present", "AP1_site_absent")
contingency_table_AP1
fisher.test(contingency_table_AP1, alternative="two.sided")
chisq.test(contingency_table_AP1)

```

```{r tead-ap1-analysis-48h}
# TEAD/AP1 motif co-localization analysis for 48h treatment
# Repeat analysis for longer treatment timepoint
drg_trtm_ATAC
ATAC_dds_results_control_48h <- ATAC_dds_results_48h$results_signif %>% 
  dplyr::filter(log2FoldChange < 0) %>% 
  dplyr::pull(PeakId)
h48_specific_peaks <- ATAC_dds_results_48h$results_signif %>% 
  dplyr::filter(log2FoldChange > 0) %>% 
  dplyr::pull(PeakId)

opts <- list()
opts[["tax_group"]] <- "vertebrates"
opts[["species"]] <- "9606"
opts[["collection"]] <- "CORE"
opts[["all_versions"]] <- FALSE
motifsToScan <- TFBSTools::getMatrixSet(JASPAR2022, opts)
TEAD_AP1_motifsToScan_names <- c("MA0090.3", 
                                 "MA0808.1", 
                                 "MA0809.2", 
                                 "MA1121.1",
                                 "MA0099.3")
TEAD_AP1_motifsToScan <- motifsToScan[TEAD_AP1_motifsToScan_names,]

# Using different subsets of peaks to check 
drg_trtm_ATAC
counts_consensus_filt_control_48h <- drg_trtm_ATAC[drg_trtm_ATAC@rowRanges@elementMetadata@listData[["PeakId"]] %in% ATAC_dds_results_control_48h, ]
counts_consensus_filt_48h <- drg_trtm_ATAC[drg_trtm_ATAC@rowRanges@elementMetadata@listData[["PeakId"]] %in% h48_specific_peaks, ]

# Create a set of random peaks for 48h 48h and control_48h
# Load genome
genome_used <- getBSgenome("BSgenome.Hsapiens.UCSC.hg38")
seqnames(genome_used) <- sub("^.{1,2}_", "", seqnames(genome_used))
seqnames(genome_used) <- sub("v", ".", seqnames(genome_used))
seqnames(genome_used) <- sub("^M", "MT", seqnames(genome_used))
# load mask regions 
blacklist_regions <- read.delim("~/workspace/neuroblastoma/resources/hg38-blacklist.v2.bed", header = FALSE)
blacklist_regions <- makeGRangesFromDataFrame(blacklist_regions, 
                                              seqnames.field = "V1", 
                                              start.field = "V2", 
                                              end.field = "V3")

# Create set of random Peaks for control_48h 
initial_peak_set <- reduce(counts_consensus_filt_control_48h@rowRanges)
random_peaks_control_48h <- initial_peak_set
random_peaks_control_48h@seqnames <- droplevels(random_peaks_control_48h@seqnames)
random_peaks_control_48h <- regioneR::randomizeRegions(random_peaks_control_48h,
                                                   allow.overlaps = FALSE,
                                                   genome = genome_used,
                                                   per.chromosome = TRUE,
                                                   mask = blacklist_regions)
random_peaks_control_48h <- as.data.frame(random_peaks_control_48h, row.names = NULL, optional = FALSE)
random_peaks_control_48h$seqnames <- droplevels(random_peaks_control_48h$seqnames)
random_peaks_control_48h <- makeGRangesFromDataFrame(random_peaks_control_48h)

# Create set of random Peaks for 48h 
initial_peak_set <- reduce(counts_consensus_filt_48h@rowRanges)
random_peaks_48h <- initial_peak_set
random_peaks_48h@seqnames <- droplevels(random_peaks_48h@seqnames)
random_peaks_48h <- regioneR::randomizeRegions(random_peaks_48h,
                                               allow.overlaps = FALSE,
                                               genome = genome_used,
                                               per.chromosome = TRUE,
                                               mask = blacklist_regions)
random_peaks_48h <- as.data.frame(random_peaks_48h, row.names = NULL, optional = FALSE)
random_peaks_48h$seqnames <- droplevels(random_peaks_48h$seqnames)
random_peaks_48h <- makeGRangesFromDataFrame(random_peaks_48h)

counts_consensus_filt_control_48h <- chromVAR::addGCBias(counts_consensus_filt_control_48h,
                                                         genome = BSgenome.Hsapiens.UCSC.hg38) 
counts_consensus_filt_48h <- chromVAR::addGCBias(counts_consensus_filt_48h,
                                                 genome = BSgenome.Hsapiens.UCSC.hg38) 

# Having corrected for bias, we can use the matchMotifs function to identify motifs under our ATACseq peaks.
# Here we supply our RangedSummarizedExperiment of counts in peaks and the genome of interest to the matchMotifs function and use the default out of matches.

# find TEAD and AP1 motifs in real control_48h data
motif_matches_control_48h <- motifmatchr::matchMotifs(pwms = TEAD_AP1_motifsToScan, 
                                                  subject = counts_consensus_filt_control_48h, 
                                                  genome = BSgenome.Hsapiens.UCSC.hg38, 
                                                  out = "positions")
control_48h_TEAD <- c(motif_matches_control_48h$MA0090.3, 
                      motif_matches_control_48h$MA0808.1, 
                      motif_matches_control_48h$MA0809.2, 
                      motif_matches_control_48h$MA1121.1)
control_48h_TEAD <- reduce(control_48h_TEAD)
control_48h_AP1 <- reduce(motif_matches_control_48h$MA0099.3)
control_48h_AP1_TEAD_overlaps <- findOverlaps(control_48h_AP1, 
                                              control_48h_TEAD, 
                                              ignore.strand = TRUE, 
                                              maxgap = 50)

# find TEAD and AP1 motifs in simulated MES data
motif_matches_control_48h_random <- motifmatchr::matchMotifs(pwms = TEAD_AP1_motifsToScan, 
                                                         subject = random_peaks_control_48h, 
                                                         genome = BSgenome.Hsapiens.UCSC.hg38, 
                                                         out = "positions")
control_48h_random_TEAD <- c(motif_matches_control_48h_random$MA0090.3, 
                             motif_matches_control_48h_random$MA0808.1, 
                             motif_matches_control_48h_random$MA0809.2, 
                             motif_matches_control_48h_random$MA1121.1)
control_48h_random_TEAD <- reduce(control_48h_random_TEAD)
control_48h_random_AP1 <- reduce(motif_matches_control_48h_random$MA0099.3)
control_48h_random_AP1_TEAD_overlaps <- findOverlaps(control_48h_random_AP1, 
                                                     control_48h_random_TEAD, 
                                                     ignore.strand = TRUE,
                                                     maxgap = 50)

# find TEAD and AP1 motifs in real 48h data
motif_matches_48h <- motifmatchr::matchMotifs(pwms = TEAD_AP1_motifsToScan, 
                                              subject = counts_consensus_filt_48h, 
                                              genome = BSgenome.Hsapiens.UCSC.hg38, 
                                              out = "positions")
h48_TEAD <- c(motif_matches_48h$MA0090.3, 
              motif_matches_48h$MA0808.1,
              motif_matches_48h$MA0809.2, 
              motif_matches_48h$MA1121.1)
h48_TEAD <- reduce(h48_TEAD)
h48_AP1 <- reduce(motif_matches_48h$MA0099.3)
h48_AP1_TEAD_overlaps <- findOverlaps(h48_AP1,
                                      h48_TEAD, 
                                      ignore.strand = TRUE,
                                      maxgap = 50)

# find TEAD and AP1 motifs in simulated 48h data
motif_matches_48h_random <- motifmatchr::matchMotifs(pwms = TEAD_AP1_motifsToScan, 
                                                     subject = random_peaks_48h, 
                                                     genome = BSgenome.Hsapiens.UCSC.hg38, 
                                                     out = "positions")
h48_random_TEAD <- c(motif_matches_48h_random$MA0090.3, 
                     motif_matches_48h_random$MA0808.1,
                     motif_matches_48h_random$MA0809.2, 
                     motif_matches_48h_random$MA1121.1)
h48_random_TEAD <- reduce(h48_random_TEAD)
h48_random_AP1 <- reduce(motif_matches_48h_random$MA0099.3)
h48_random_AP1_TEAD_overlaps <- findOverlaps(h48_random_AP1, 
                                             h48_random_TEAD, 
                                             ignore.strand = TRUE,
                                             maxgap = 50)

summary_table_peaks <- data.frame(total_number_of_peaks = c(length(counts_consensus_filt_control_48h), 
                                                            length(random_peaks_control_48h),
                                                            length(counts_consensus_filt_48h),
                                                            length(random_peaks_48h)),
                                  TEAD_sites = c(length(control_48h_TEAD),
                                                 length(control_48h_random_TEAD),
                                                 length(h48_TEAD),
                                                 length(h48_random_TEAD)),
                                  AP1_sites = c(length(control_48h_AP1),
                                                length(control_48h_random_AP1),
                                                length(h48_AP1),
                                                length(h48_random_AP1)),
                                  TEAD_AP1_coloc_sites = c(length(control_48h_AP1_TEAD_overlaps),
                                                           length(control_48h_random_AP1_TEAD_overlaps),
                                                           length(h48_AP1_TEAD_overlaps),
                                                           length(h48_random_AP1_TEAD_overlaps)),
                                  row.names = c("control_48h", "Simulated control_48h", "h48", "Simulated h48")
)
summary_table_peaks


# Perform Fisher's exact tests for 48h treatment comparison
contingency_table <- matrix(c(length(h48_AP1_TEAD_overlaps), 
                              length(h48_TEAD) + length(h48_AP1) - length(h48_AP1_TEAD_overlaps), 
                              length(control_48h_AP1_TEAD_overlaps), 
                              length(control_48h_TEAD) + length(control_48h_AP1) - length(control_48h_AP1_TEAD_overlaps)),
                            nrow = 2, 
                            byrow = TRUE)
fisher_results <- fisher.test(contingency_table,
                              alternative = "two.sided")
fisher_summary_table <- data.table::data.table(comparison = "48h_vs_control_48h",
                                               pval = fisher_results$p.value,
                                               odds_ratio = fisher_results$estimate)

# Fisher test comparing real h48 AP1_TEAD peaks vs simulated
contingency_table_h48_vs_simulated <- matrix(c(length(h48_AP1_TEAD_overlaps), 
                                               length(h48_TEAD) + length(h48_AP1) - length(h48_AP1_TEAD_overlaps),
                                               length(h48_random_AP1_TEAD_overlaps), 
                                               length(h48_random_TEAD) + length(h48_random_AP1) - length(h48_random_AP1_TEAD_overlaps)), 
                                             nrow = 2,
                                             byrow = TRUE)
row.names(contingency_table_h48_vs_simulated) <- c("h48_peaks", "h48_peaks_random")
colnames(contingency_table_h48_vs_simulated) <- c("TEAD_AB1_colocalisation_present", "TEAD_AB1_colocalisation_absent")
contingency_table_h48_vs_simulated
fisher_results <- fisher.test(contingency_table_h48_vs_simulated,
                              alternative = "two.sided")
fisher_summary_table <- rbind(fisher_summary_table,
                              data.table::data.table(comparison = "h48_vs_h48sim",
                                                     pval = fisher_results$p.value,
                                                     odds_ratio = fisher_results$estimate))

# Fisher test for TEAD peaks
contingency_table_TEAD <- matrix(c(length(h48_TEAD), 
                                   length(counts_consensus_filt_48h) - length(h48_TEAD),
                                   length(control_48h_TEAD), 
                                   length(counts_consensus_filt_control_48h) - length(control_48h_TEAD)), 
                                 nrow = 2, byrow = TRUE)
row.names(contingency_table_TEAD) <- c("h48_peaks", "control_48h_peaks")
colnames(contingency_table_TEAD) <- c("TEAD_site_present", "TEAD_site_absent")
contingency_table_TEAD
fisher.test(contingency_table_TEAD, 
            alternative="two.sided")
chisq.test(contingency_table_TEAD)

# Fisher test for AP1 peaks
contingency_table_AP1 <- matrix(c(length(h48_AP1), 
                                  length(counts_consensus_filt_48h) - length(h48_AP1),
                                  length(control_48h_AP1), 
                                  length(counts_consensus_filt_control_48h) - length(control_48h_AP1)), 
                                nrow = 2,
                                byrow = TRUE)
row.names(contingency_table_AP1) <- c("h48_peaks", "control_48h_peaks")
colnames(contingency_table_AP1) <- c("AP1_site_present", "AP1_site_absent")
contingency_table_AP1
fisher.test(contingency_table_AP1, 
            alternative = "two.sided")
chisq.test(contingency_table_AP1)

```

```{r gene-set-enrichment-24h, error=TRUE}
# Gene set enrichment analysis for 24h treatment
# Test for MES/ADRN signature enrichment in TEAD/AP1 peaks

import::from(
  .from = "~/workspace/neuroblastoma/resources/utilityScripts.R",
  "chipEnrichAndExport"
)
locusdef <- "5kb"
our_rna_seq_terms <- "/home/rstudio/workspace/neuroblastoma/resources/mes_adrn_GS_frm_RNA_seq.tsv"
res_dir <- "/home/rstudio/workspace/neuroblastoma/results/ATAC-seq_drug_treatment/"

drg_trtm_ATAC
ATAC_dds_results_control <- ATAC_dds_results_24h$results_signif %>% 
  dplyr::filter(log2FoldChange < 0) %>% 
  dplyr::pull(PeakId)
h24_specific_peaks <- ATAC_dds_results_24h$results_signif %>% 
  dplyr::filter(log2FoldChange > 0) %>% 
  dplyr::pull(PeakId)
opts <- list()
opts[["tax_group"]] <- "vertebrates"
opts[["species"]] <- "9606"
opts[["collection"]] <- "CORE"
opts[["all_versions"]] <- FALSE
motifsToScan <- TFBSTools::getMatrixSet(JASPAR2022, opts)
TEAD_AP1_motifsToScan_names <- c("MA0090.3", 
                                 "MA0808.1",
                                 "MA0809.2",
                                 "MA1121.1",
                                 "MA0099.3")
TEAD_AP1_motifsToScan <- motifsToScan[TEAD_AP1_motifsToScan_names,]

# Using different subsets of peaks to check 
drg_trtm_ATAC
counts_consensus_filt_control <- drg_trtm_ATAC[drg_trtm_ATAC@rowRanges@elementMetadata@listData[["PeakId"]] %in% ATAC_dds_results_control, ]
counts_consensus_filt_24h <- drg_trtm_ATAC[drg_trtm_ATAC@rowRanges@elementMetadata@listData[["PeakId"]] %in% h24_specific_peaks, ]

motif_matches_control <- motifmatchr::matchMotifs(pwms = TEAD_AP1_motifsToScan, 
                                                  subject = counts_consensus_filt_control, 
                                                  genome = BSgenome.Hsapiens.UCSC.hg38, 
                                                  out = "positions")
control_TEAD <- c(motif_matches_control$MA0090.3,
                  motif_matches_control$MA0808.1,
                  motif_matches_control$MA0809.2,
                  motif_matches_control$MA1121.1)
control_TEAD <- reduce(control_TEAD)
control_AP1 <- reduce(motif_matches_control$MA0099.3)


control_AP1_TEAD_overlaps <- ChIPpeakAnno::findOverlapsOfPeaks(control_AP1,
                                                               control_TEAD,
                                                               maxgap = 50)
control_AP1_TEAD_overlaps <- control_AP1_TEAD_overlaps$peaksInMergedPeaks
control_AP1_TEAD_overlaps <- as.data.frame(control_AP1_TEAD_overlaps)

if(nrow(control_AP1_TEAD_overlaps) != 0){
  chipEnrichAndExport(peaks = control_AP1_TEAD_overlaps,
                      peaksName = "control", 
                      TF_name = "control_AP1_TEAD", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}

if(nrow(as.data.frame(control_AP1)) != 0){
  chipEnrichAndExport(peaks = as.data.frame(control_AP1),
                      peaksName = "control", 
                      TF_name = "control_AP1", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}

if(nrow(as.data.frame(control_TEAD)) != 0){
  chipEnrichAndExport(peaks = as.data.frame(control_TEAD),
                      peaksName = "control", 
                      TF_name = "control_TEAD", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}

control <- reduce(counts_consensus_filt_control@rowRanges)
if(nrow(as.data.frame(control)) != 0){
  chipEnrichAndExport(peaks = as.data.frame(control),
                      peaksName = "control", 
                      TF_name = "control_Total", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}

h24_AP1_TEAD_overlaps <- ChIPpeakAnno::findOverlapsOfPeaks(h24_AP1, h24_TEAD, maxgap = 50)
h24_AP1_TEAD_overlaps <- h24_AP1_TEAD_overlaps$peaksInMergedPeaks
h24_AP1_TEAD_overlaps <- as.data.frame(h24_AP1_TEAD_overlaps)

if(nrow(as.data.frame(h24_AP1_TEAD_overlaps)) != 0){
  chipEnrichAndExport(peaks = h24_AP1_TEAD_overlaps,
                      peaksName = "h24", 
                      TF_name = "h24_AP1_TEAD", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}

#the intersection here doesn't exist. The function returns error. This is why it's commented out.
# if(nrow(as.data.frame(h24_AP1)) != 0){
#   chipEnrichAndExport(peaks = as.data.frame(h24_AP1),
#                       peaksName = "h24", 
#                       TF_name = "h24_AP1", 
#                       res_dir = res_dir, 
#                       genesets = our_rna_seq_terms, 
#                       genesets_name = "Our_data_",
#                       locusdef = locusdef
#   )
# }

if(nrow(as.data.frame(h24_TEAD)) != 0){
  chipEnrichAndExport(peaks = as.data.frame(h24_TEAD),
                      peaksName = "h24", 
                      TF_name = "h24_TEAD", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}

h24 <- reduce(counts_consensus_filt_24h@rowRanges)
if(nrow(as.data.frame(h24)) != 0){
  chipEnrichAndExport(peaks = as.data.frame(h24),
                      peaksName = "h24", 
                      TF_name = "h24_Total", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}

enrich_results_files <- unlist(list.files(path = res_dir, 
                                          pattern = "^Enricher_.*(h24|control).*.xlsx", 
                                          full.names = TRUE))
summary_table <- data.frame(Protein = NULL,
                            Data_source = NULL,
                            Selected_peaks = NULL,
                            Description = NULL,
                            P.value = NULL,
                            FDR = NULL,
                            Effect = NULL,
                            Status = NULL,
                            Gene_set_size = NULL,
                            Peaks_in_set = NULL,
                            Odds_ratio = NULL)

pattern <-  ".*(control_Total|control_AP1|control_TEAD|control_AP1_TEAD|h24_Total|h24_AP1|h24_TEAD|h24_AP1_TEAD)_(Our_data|Groen).*(control|h24).*"

for (file_name in enrich_results_files){
  tmp <- openxlsx2::wb_to_df(file_name, sheet = 1)
  
  basename_tmp <- basename(file_name)
  basename_tmp <- str_extract(basename_tmp, 
                              pattern, 
                              group = c(1,2,3))
  
  summary_table_tmp <- data.frame(Protein = basename_tmp[1],
                                  Data_source = basename_tmp[2],
                                  Selected_peaks = basename_tmp[3],
                                  Description = tmp$Description,
                                  P.value = tmp$P.value,
                                  FDR = tmp$FDR,
                                  Effect = tmp$Effect,
                                  Status = tmp$Status,
                                  Gene_set_size = tmp$N.Geneset.Genes,
                                  Peaks_in_set = tmp$N.Geneset.Peak.Genes,
                                  Odds_ratio = tmp$Odds.Ratio)
  
  summary_table <- rbind(summary_table,
                         summary_table_tmp)
  
}
# Summary: No significant enrichment detected in 24h treatment analysis
summary_table
```

```{r gene-set-enrichment-48h, error=TRUE}
# Gene set enrichment analysis for 48h treatment
# Repeat enrichment analysis for longer treatment timepoint
import::from(
  .from = "~/workspace/neuroblastoma/resources/utilityScripts.R",
  "chipEnrichAndExport"
)
locusdef <- "5kb"
our_rna_seq_terms <- "/home/rstudio/workspace/neuroblastoma/resources/mes_adrn_GS_frm_RNA_seq.tsv"
res_dir <- "/home/rstudio/workspace/neuroblastoma/results/ATAC-seq_drug_treatment/"

drg_trtm_ATAC
ATAC_dds_results_control <- ATAC_dds_results_48h$results_signif %>% 
  dplyr::filter(log2FoldChange < 0) %>% 
  dplyr::pull(PeakId)
h48_specific_peaks <- ATAC_dds_results_48h$results_signif %>% 
  dplyr::filter(log2FoldChange > 0) %>% 
  dplyr::pull(PeakId)
opts <- list()
opts[["tax_group"]] <- "vertebrates"
opts[["species"]] <- "9606"
opts[["collection"]] <- "CORE"
opts[["all_versions"]] <- FALSE
motifsToScan <- TFBSTools::getMatrixSet(JASPAR2022, opts)
TEAD_AP1_motifsToScan_names <- c("MA0090.3", 
                                 "MA0808.1",
                                 "MA0809.2",
                                 "MA1121.1",
                                 "MA0099.3")
TEAD_AP1_motifsToScan <- motifsToScan[TEAD_AP1_motifsToScan_names,]

# Using different subsets of peaks to check 
drg_trtm_ATAC
counts_consensus_filt_control <- drg_trtm_ATAC[drg_trtm_ATAC@rowRanges@elementMetadata@listData[["PeakId"]] %in% ATAC_dds_results_control, ]
counts_consensus_filt_48h <- drg_trtm_ATAC[drg_trtm_ATAC@rowRanges@elementMetadata@listData[["PeakId"]] %in% h48_specific_peaks, ]

motif_matches_control <- motifmatchr::matchMotifs(pwms = TEAD_AP1_motifsToScan, 
                                                  subject = counts_consensus_filt_control, 
                                                  genome = BSgenome.Hsapiens.UCSC.hg38, 
                                                  out = "positions")
control_TEAD <- c(motif_matches_control$MA0090.3,
                  motif_matches_control$MA0808.1,
                  motif_matches_control$MA0809.2,
                  motif_matches_control$MA1121.1)
control_TEAD <- reduce(control_TEAD)
control_AP1 <- reduce(motif_matches_control$MA0099.3)

control_AP1_TEAD_overlaps <- ChIPpeakAnno::findOverlapsOfPeaks(control_AP1,
                                                               control_TEAD,
                                                               maxgap = 50)
control_AP1_TEAD_overlaps <- control_AP1_TEAD_overlaps$peaksInMergedPeaks
control_AP1_TEAD_overlaps <- as.data.frame(control_AP1_TEAD_overlaps)

if(nrow(control_AP1_TEAD_overlaps) != 0){
  chipEnrichAndExport(peaks = control_AP1_TEAD_overlaps,
                      peaksName = "control", 
                      TF_name = "control_AP1_TEAD", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}

if(nrow(as.data.frame(control_AP1)) != 0){
  chipEnrichAndExport(peaks = as.data.frame(control_AP1),
                      peaksName = "control", 
                      TF_name = "control_AP1", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}

if(nrow(as.data.frame(control_TEAD)) != 0){
  chipEnrichAndExport(peaks = as.data.frame(control_TEAD),
                      peaksName = "control", 
                      TF_name = "control_TEAD", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}

control <- reduce(counts_consensus_filt_control@rowRanges)
if(nrow(as.data.frame(control)) != 0){
  chipEnrichAndExport(peaks = as.data.frame(control),
                      peaksName = "control", 
                      TF_name = "control_Total", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}


# Analysing 48h
h48_AP1_TEAD_overlaps <- ChIPpeakAnno::findOverlapsOfPeaks(h48_AP1,
                                                           h48_TEAD,
                                                           maxgap = 50)
h48_AP1_TEAD_overlaps <- h48_AP1_TEAD_overlaps$peaksInMergedPeaks
h48_AP1_TEAD_overlaps <- as.data.frame(h48_AP1_TEAD_overlaps)

# There are no overlaps between peaks, function returns error. The code is commented out.
# if(nrow(as.data.frame(h48_AP1_TEAD_overlaps)) != 0){
#   chipEnrichAndExport(peaks = h48_AP1_TEAD_overlaps,
#                       peaksName = "h48", 
#                       TF_name = "h48_AP1_TEAD", 
#                       res_dir = res_dir, 
#                       genesets = our_rna_seq_terms, 
#                       genesets_name = "Our_data_",
#                       locusdef = locusdef
#   )
# }

if(nrow(as.data.frame(h48_AP1)) != 0){
  chipEnrichAndExport(peaks = as.data.frame(h48_AP1),
                      peaksName = "h48", 
                      TF_name = "h48_AP1", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}

if(nrow(as.data.frame(h48_TEAD)) != 0){
  chipEnrichAndExport(peaks = as.data.frame(h48_TEAD),
                      peaksName = "h48", 
                      TF_name = "h48_TEAD", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}

h48 <- reduce(counts_consensus_filt_48h@rowRanges)
if(nrow(as.data.frame(h48)) != 0){
  chipEnrichAndExport(peaks = as.data.frame(h48),
                      peaksName = "h48", 
                      TF_name = "h48_Total", 
                      res_dir = res_dir, 
                      genesets = our_rna_seq_terms, 
                      genesets_name = "Our_data_",
                      locusdef = locusdef
  )
}

enrich_results_files <- unlist(list.files(path = res_dir,
                                          pattern = "^Enricher_.*(h48|control).*.xlsx",
                                          full.names = TRUE))
summary_table <- data.frame(Protein = NULL,
                            Data_source = NULL,
                            Selected_peaks = NULL,
                            Description = NULL,
                            P.value = NULL,
                            FDR = NULL,
                            Effect = NULL,
                            Status = NULL,
                            Gene_set_size = NULL,
                            Peaks_in_set = NULL,
                            Odds_ratio = NULL)

pattern <-  ".*(control_Total|control_AP1|control_TEAD|control_AP1_TEAD|h48_Total|h48_AP1|h48_TEAD|h48_AP1_TEAD)_(Our_data|Groen).*(control|h48).*"

for (file_name in enrich_results_files){
  tmp <- openxlsx2::wb_to_df(file_name, sheet = 1)
  
  basename_tmp <- basename(file_name)
  basename_tmp <- str_extract(basename_tmp, 
                              pattern, 
                              group = c(1,2,3))
  
  summary_table_tmp <- data.frame(Protein = basename_tmp[1],
                                  Data_source = basename_tmp[2],
                                  Selected_peaks = basename_tmp[3],
                                  Description = tmp$Description,
                                  P.value = tmp$P.value,
                                  FDR = tmp$FDR,
                                  Effect = tmp$Effect,
                                  Status = tmp$Status,
                                  Gene_set_size = tmp$N.Geneset.Genes,
                                  Peaks_in_set = tmp$N.Geneset.Peak.Genes,
                                  Odds_ratio = tmp$Odds.Ratio)
  
  summary_table <- rbind(summary_table,
                         summary_table_tmp)
}
summary_table
# Summary: Only control condition shows significant enrichment in 48h analysis
```
